### 2024-09-18 docs/writing/outline.tex
Porting from markdown to tex, have not added edits from Lizzie

\documentclass{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\usepackage{outlines}
\usepackage{enumitem}
\setenumerate[1]{label=\Roman*.}
\setenumerate[2]{label=\Alph*.}
\setenumerate[3]{label=\roman*.}
\setenumerate[4]{label=\alph*.}
\begin{document}
\section {Paper Outline}
\begin{outline}[itemize]

\1 Abstract
\begin{itemize}
	\item  Digital twins of samples is important for building data repositories and promoting replicable science
	\item  Digitizing wood samples has limits to sample dimensions or lacks affordable implementations
	\item 	Current alternatives are very expensive (\$70,000USD), lack a high level of detail (printer scanners), or cannot scan cookies (CaptuRING)
\end{itemize}
\item Introduction
	\begin{itemize}
	\item Development of measurement and recording systems
	\item Digitizing wood samples for better data pooling / archiving
	\item Software can be used to help record annual growth 
	\item Main reasons 
		\begin{itemize}
		\item affordable
		\item open source
		\item  alternative to scanners, without size limitations to digitize tree cookies and cores 
		\end{itemize}
	\item Provide a means to study tree growth from more detailed data than tree cores alone
	\item Development of measurement and recording systems
	\end{itemize}
\item Materials and Methods
	\begin{itemize}
	\item System Design 
		\begin{itemize}
		\item Hardware
			\begin{itemize}
			\item Cartesian gantry robot
			\item Camera
			\item computer
			\item camera and computer mount
			\end{itemize}
		\item Software
			\begin{itemize}
			\item image focusing
			\item grid traverse
			\item feature based image stitching
			\subitem memory mapped 
			\item user interface
			\end{itemize}
		\end{itemize}
	\end{itemize}
\item Results
	\begin{itemize}
	\item ultra high resolution scans (DPI 15,000 +)
	\item scans of cookies / cores
	\item  Digitization time as a function of surface area
	\item File-size results in functional limits 
		\begin{itemize}
		\item max filesize for TIFF files is 2.5 GB
		\item other lossless filetypes for larger images are not compatible with standard image viewers
		\item viewing a pseudo-core of a larger image and cropping 
		\item downscaling
		\end{itemize}
	\end{itemize}
\item Discussion (for MEE Only)
	\begin{itemize}
	\item Potential for vessel counts for an entire ring 
	\item Potential for automatic tree ring identification with image processing / ML techniques 
	\item strengths
		\begin{itemize}
		\item  Ability to capture multiple cores / cookies simultaneously
		\item cost effective
		\end{itemize}
	\item limitations
		\item poor lense quality
		\item The final quality is only as good as the sample preparation, microtomes are very expensive
		\item lack of focus stacking
		\item Difficulty stitching with lower detail images, can't increase digitizing speed, must downscale high resolution images
	\end{itemize}
\item Conclusions (optional for MEE)

\end{outline}
\end{document}

### 2024-09-18 docs/writing/outline.tex
Adding suggested changes, still work in progress

@@
 	\begin{itemize}
+	\item Tree ring science overview
+		\begin{itemize}
+		\item **Lizzie add here**
+		\item tree rings can help reconstruct past climates 
+		\item 
+		\end{itemize}
 	\item Development of measurement and recording systems
@@
 	\begin{itemize}
-	\item Potential for vessel counts for an entire ring 
-	\item Potential for automatic tree ring identification with image processing / ML techniques 
-	\item strengths
+	\item strengths and opportunities
 		\begin{itemize}
@@
 		\item cost effective
+		\item potential for automatic tree ring identification with iamge processing / ML 
+		\item potential for vessel counts for an entire ring 
+		\subitem although final quality is only as good as teh sample preparation
 		\end{itemize}
-	\item limitations
-		\item poor lense quality
-		\item The final quality is only as good as the sample preparation, microtomes are very expensive
-		\item lack of focus stacking
+	\item opportunities for improvements
+		\item getting a better lens
+			\begin{itemize}
+			\item increases sharpness
+			\item improves stitching accuracy
+			\item obtain similar quality between the corner and center of the image
+			\subitem poor lens quality seemed to be the culprit of producing visible seams when stitching with larger field of view images 
+			\end{itemize}
+
+		\item focus stacking
+		\subitem could allow for non planar objects to be digitized and stitched
 		\item Difficulty stitching with lower detail images, can't increase digitizing speed, must downscale high resolution images 
+		\item Autofocusing with hardware would significantly decrease the time to capture a cookie (using a professional camera)
 	\end{itemize}

### 2024-09-19 docs/writing/outline.tex
Updating outline, including pdf render

@@
 \begin{itemize}
+	\item Flatbed scanner alternative with improved resolution to digitizing tree cookies and cores
 	\item Digital twins of samples is important for building data repositories and promoting replicable science
@@
 	\item 	Current alternatives are very expensive (\$70,000USD), lack a high level of detail (printer scanners), or cannot scan cookies (CaptuRING)
+	\item Tina is open source and able to be assembled with 3D printed parts, OpenBuilds parts kits, and hand tools
 \end{itemize}
@@
 		\item tree rings can help reconstruct past climates 
-		\item 
+		\item understanding tree growth mechanisms
 		\end{itemize}
-	\item Development of measurement and recording systems
+	\item Development of scanning alternatives
+		\begin{itemize}
+		\item beginning with ATRICS
+		\item expensive Gigapixel
+		\item DIY CaptuRING
+		\end{itemize}
+	\item Main reasons to build this 
+		\begin{itemize}
+		\item further develop the affordable / open source alternatives for scientific equipment
+		\item reduce the need for proprietary software such as PTGUI for image stitching
+		\item allow a method to digitize cookies to have higher quality data than cores alone
+		\item avoid size limitations to digitize tree cookies and cores 
 		\item Digitizing wood samples for better data pooling / archiving
+		\end{itemize}
 	\item Software can be used to help record annual growth 
-	\item Main reasons 
 		\begin{itemize}
-		\item affordable
-		\item open source
-		\item  alternative to scanners, without size limitations to digitize tree cookies and cores 
+		\item Coorecorder
+		\item Various ML based methods that Sandy has researched
 		\end{itemize}
-	\item Provide a means to study tree growth from more detailed data than tree cores alone
-	\item Development of measurement and recording systems
 	\end{itemize}
@@
 		\begin{itemize}
-	\item System Design 
+		\item Hardware Design
 			\begin{itemize}
-		\item Hardware
+			\item Cartesian gantry robot kit from OpenBuilds (ACRO system)
+			\item Camera 
 				\begin{itemize}
-			\item Cartesian gantry robot
-			\item Camera
+				\item chose Raspberry Pi HQ Camera for C-mount lens options, ability to stream, cost effectiveness, and clear obscelescence statement
+				\end{itemize}
 			\item computer
-			\item camera and computer mount
+				\begin{itemize}
+				\item NVIDIA Jetson Orin Nano for processing power in a light package, can connect to Raspberry Pi camera. Powerful enough to power a GUI and monitor display as well
+				\item NVMM Hardware accelerated video streaming reduces CPU load 
 				\end{itemize}
-		\item Software
+			\item 3D printed components
+				\begin{itemize}
+				\item 3D printed parts from a Bambu X1C were designed to mount the lens and computer to the ACRO system
+				\item other parts such as a drag chain for cable management were also 3D printed
+				\item levelling tables were designed to allow for the cookie samples to be arranged orthogonally to the lens. Imperative to capturing in focus images across the entire cookie
+				\end{itemize}
+			\end{itemize}
+		\item Software Design
 			\begin{itemize}
 			\item image focusing
+				\begin{itemize}
+				\item sample levelling is not perfect.
+At different points on the cookie, the Z-height of the gantry must vary to capture the in focus image.
+A routine was made to take multiple images at varied Z-heights.
+The image with the highest normalized variance score was chosen as the in focus image and is kept.
+				\item automatic PID control was implemented to adjust the initial Z-coordinate for each set of images to increase the probability of having an in focus image at each image set
+				\end{itemize}
 			\item grid traverse
+				\begin{itemize}
+				\item The software traverses the entire surface area of the sample. Each row and column is overlapping with its adjacent row and column. Allowing for feature based image sttiching.
+				\end{itemize}
 			\item feature based image stitching
-			\subitem memory mapped 
+				\begin{itemize}
+				\item An implementation, Stitch2D, of feature based OpenCV image stitching was used.
+Edits to the code were made to allow for more memory efficient methods of stitching with the use of NumPy memory maps
+				\end{itemize}
 			\item user interface
+				\begin{itemize}
+				\item A GUI was made to navigate the machine to samples, set the size of the sample and its center. After samples are set up the operation is passive.
 				\end{itemize}
@@
 		\end{itemize}
+
 \item Results
 	\begin{itemize}
+	\item scans of cookies / cores
+		\begin{itemize}
 		\item ultra high resolution scans (DPI 15,000 +)
-	\item scans of cookies / cores
-	\item  Digitization time as a function of surface area
+		\item downscaled versions as high resolution is not necessary for all applications
+		\item Digitization time as a function of surface area. Large cookies can scan only a portion of the cookie including the center and half of the rings.
+		\item large surface areas can produce files that are inconveniently large
+		\end{itemize}
 	\item File-size results in functional limits 
@@
 		\item other lossless filetypes for larger images are not compatible with standard image viewers
-		\item viewing a pseudo-core of a larger image and cropping 
-		\item downscaling
+		\item workaround with NumPy memory maps and cropped viewing 
 		\end{itemize}
@@
 		\item potential for vessel counts for an entire ring 
-		\subitem although final quality is only as good as teh sample preparation
+		\subitem although final quality is only as good as the sample preparation
 		\end{itemize}
 	\item opportunities for improvements
-		\item getting a better lens
+		\item Lenses
 			\begin{itemize}
-			\item increases sharpness
-			\item improves stitching accuracy
+			\item troubles with fixed aperature / non autofocusing / poorly manufactured lens
+				\begin{itemize}
+				\item lack of control to increase sharpness
+				\item autofocusing with Z-axis movement is time consuming to do without image blur
+				\item poor lens manufacturing can lead to drastically different quality / lighting between the edges and center of an image
+				\end{itemize}
+			\item good lens can increase sharpness and therefore stitching accuracy and ML inference
 			\item obtain similar quality between the corner and center of the image
@@
 		\item focus stacking
-		\subitem could allow for non planar objects to be digitized and stitched
+		\subitem when a sample is microtomed, the vessels are hollow. The autofocusing algorithm can be confused from this 
 		\item Difficulty stitching with lower detail images, can't increase digitizing speed, must downscale high resolution images 
@@
 \item Conclusions (optional for MEE)
+	\begin{itemize}
+	\item
+	\end{itemize}
 

### 2024-09-30 docs/writing/operation_manual.tex
Adding skeleton of operation manual. Needs visuals

\documentclass{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\usepackage{outlines}
\usepackage{enumitem}
\setenumerate[1]{label=\Roman*.}
\setenumerate[2]{label=\Alph*.}
\setenumerate[3]{label=\roman*.}
\setenumerate[4]{label=\alph*.}
\begin{document}
\section {Operation Manual}
\begin{outline}[enumerate]

\1 Device Setup
\begin{enumerate}
	\item Turn on the Jetson by plugging the power port in
	\item Turn on the motor controller with the switch
	\item Turn on the power supply with the switch 
	\item Ensure everything is plugged in correctly
        \begin{enumerate}
        \item USB ring light is powered
        \item Jetson is connected to motor controller with a USB-B to USB-A cable
        \item Jetson is plugged into the monitor with DisplayPort to HDMI converter
        \item Power supply is plugged into the wall and also into the motor controller
        \item Motor controller is connected to X, Y, and Z motors
        \item Motor controller is connected to X, Y, and Z limit switches
        \end{enumerate}
    \item Log into the Jetson
        \begin{enumerate}
        \item Username: dragonfly
        \item Password: alls33ingFly
        \end{enumerate}
    \item Launch the GUI
        \begin{enumerate}
        \item Open Terminal
        \item Run the following
            \begin{enumerate}
                \item cd ~/Desktop/refactored
                \item python gui.py
            \end{enumerate}
        \end{enumerate}
    \item Use the GUI to connect to establish a connection between the Jetson and the motor controller
        \begin{enumerate}
        \item Press the physical 'Reset' button on the motor controller
        \item Click the 'Serial Connect' button on the GUI and wait until the GUI is responsive again.
        \item Test that the connection has been made by clicking the arrows to manually jog the machine 
        \end{enumerate}
    \item Set Home
        \begin{enumerate}
        \item Confirm that there are no obstructions in the machine that will cause a collision
        \item Click the 'Set Home' button on the GUI. This will move the machine towards the limit switches until they are triggered
        \item *** If the machine doesn't move, repeat the previous step to establish a connection between the Jetson and motor controller ***
        \end{enumerate}
\end{enumerate}
\item Sample Setup
	\begin{enumerate}
	\item Sample levelling
		\begin{enumerate}
		\item Ensure that the the top and bottom face of the cookies are relatively level
        \item Take note of the sample name, id, etc on to a piece of paper
		\item Place the cookie on sampling table
		\item Place the bullseye level on top of the sample
        \item Rotate the sample until the bubble is biased towards the side of the levelling table with only one screw
        \item Loosen the lone screw until the bubble is centered on the bullseye
		\end{enumerate}
	\item Multiple samples
		\begin{enumerate}
		\item Level all samples and make sure to note their sample identifiers and relative positions on a piece of paper 
		\end{enumerate}
	\end{enumerate}
\item Capturing samples
    \begin{enumerate}
    \item Click the 'Fast' option in the jogging controls section of the GUI
    \item Set the jogging distance to be relatively large like 10 or 20mm
    \item Navigate with the jogging arrows to the approximate middle of a sample
    \item Jog in the Z-axis to get the sample in focus
        \begin{enumerate}
        \item I like to start by setting the jog to 5mm, and seeing the sample go into then out of focus
        \item To tune closer, I then change the jog to 1mm to get roughly to the focus
        \item Finally, I switch to 0.1mm and get as in focus as possible
        \item Note that it is pretty important to start off in focus. While the machine has a control algorithm to help keep the machine in focus, it performs best with a good start
        \end{enumerate}
    \item Fill in the 'Sample Height' and 'Sample Width' text boxes with an estimation of the size
    \item Click 'Test Sample Boundaries'
        \begin{enumerate}
        \item The purpose of testing the boundaries is to show the user the border of the sample. 
        \item The button will take the location of when you click the button, and traverse around the border of the sample according to what you defined as the height and width
        \item You should watch to verify that all of the sample that you want is within the border
        \item If you do not see all of the cookie, adjust either the starting location of the sample with the jogging arrows, the 'Sample Height' text box, the 'Sample Width' text box, or a combination of the three
        \end{enumerate}
    \end{enumerate}
\item Potential errors
    \begin{enumerate}
    \item Running out of memory
        \begin{enumerate}
        \item While this shouldn't be a problem, there is still a memory problem in the stitching implementation from when it was ported from Stitch2D. In theory this is fixable, but I haven't had the time to get to the root bug yet.
        \item If you need a higher resolution stitch than what the Jetson can do, you'll need to load the images onto a computer with more RAM, such as the Legion laptop, and stitch them on there. 
        \end{enumerate}
    \item Visible seams
        \begin{enumerate}
        \item This is an annoying artifact from what we have deduced to be a poor lens. If you want to maximize sampling speed, it is possible to take zoomed out photos with larger field of view. Although this 
        results in visible seams which have seemingly only been able to be fixed by cropping the images to a size which effectively cancels out the increased field of view.
        \item If you are going to be using a manual dating software like Coorecorder, then the seams may not be the end of the world. As long as you can differentiate the seams from the rings.
        \end{enumerate}
    \item 
    \end{enumerate}

\end{outline}
\end{document}

### 2024-10-01 docs/writing/operation_manual.tex
Adding figures and more nuance to instructions

@@
 \documentclass{article}
+\usepackage{graphicx}
 \usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
@@
 \setenumerate[4]{label=\alph*.}
+\title {Operation Manual}
 \begin{document}
-\section {Operation Manual}
+\maketitle
 \begin{outline}[enumerate]
 
-\1 Device Setup
+\1 {Device Setup}
 \begin{enumerate}
-	\item Turn on the Jetson by plugging the power port in
-	\item Turn on the motor controller with the switch
-	\item Turn on the power supply with the switch 
-	\item Ensure everything is plugged in correctly
+	\item Turn on the Jetson by plugging the power port in. See Figure \ref{fig:jetson1}.
+	\item Turn on the motor controller with the switch. See Figure \ref{fig:controller3}.
+	\item Turn on the power supply with the switch. See Figure \ref{fig:power1}.
+	\item Ensure everything is plugged in correctly. 
         \begin{enumerate}
-        \item USB ring light is powered
+        \item USB ring light is powered. It's best NOT to power it with the USB ports on the Jetson.
         \item Jetson is connected to motor controller with a USB-B to USB-A cable
@@
         \begin{enumerate}
-        \item Open Terminal
+        \item Open Terminal. See Figure \ref{fig:terminal1}
         \item Run the following
@@
                 \item cd ~/Desktop/refactored
+                \item sudo chmod 777 /dev/ttyUSB0
                 \item python gui.py
@@
         \end{enumerate}
-    \item Use the GUI to connect to establish a connection between the Jetson and the motor controller
+    \item Use the GUI to connect to establish a connection between the Jetson and the motor controller. See Figure \ref{fig:gui1}
         \begin{enumerate}
-        \item Press the physical 'Reset' button on the motor controller
-        \item Click the 'Serial Connect' button on the GUI and wait until the GUI is responsive again.
-        \item Test that the connection has been made by clicking the arrows to manually jog the machine 
+        \item Press the physical 'Reset' button on the motor controller. See Figure \ref{fig:controller3}
+        \item Click the 'Serial Connect' button on the GUI and wait until the GUI is responsive again.  See Figure \ref{fig:gui1}
+        \item Test that the connection has been made by clicking the arrows to manually jog the machine. See Figure \ref{fig:gui1}
         \end{enumerate}
@@
         \item Confirm that there are no obstructions in the machine that will cause a collision
-        \item Click the 'Set Home' button on the GUI. This will move the machine towards the limit switches until they are triggered
+        \item Click the 'Set Home' button on the GUI. This will move the machine towards the limit switches until they are triggered.  See Figure \ref{fig:gui1}
         \item *** If the machine doesn't move, repeat the previous step to establish a connection between the Jetson and motor controller ***
@@
 		\item Place the cookie on sampling table
-		\item Place the bullseye level on top of the sample
+		\item Place the bullseye level on top of the sample. See Figure \ref{fig:level1}
         \item Rotate the sample until the bubble is biased towards the side of the levelling table with only one screw
-        \item Loosen the lone screw until the bubble is centered on the bullseye
+        \item Loosen the lone screw until the bubble is centered on the bullseye. See Figure \ref{fig:level2}
 		\end{enumerate}
-	\item Multiple samples
-		\begin{enumerate}
-		\item Level all samples and make sure to note their sample identifiers and relative positions on a piece of paper 
-		\end{enumerate}
-	\end{enumerate}
-\item Capturing samples
+\item Capturing Samples
     \begin{enumerate}
@@
     \end{enumerate}
+    \item Click 'Add Cookie'
+        \begin{enumerate}
+            \item Once the boundaries are looking inclusive of the entire sample, add the cookie sample.
+            \item If you only want to do one cookie, now you can click 'Capture All Cookies'. If not, continue to Multiple Samples.
+        \end{enumerate}
+    \item Multiple samples
+		\begin{enumerate}
+		\item Repeat the sample levelling process for each cookie. Then repeat the Capturing Samples process. When you are ready click 'Capture All Cookies'
+        \item Note that it's possible for the device to fail when stitching cookies. If this happens it will nullify your work setting up the cookies unfortunately. 
+        \end{enumerate}
+    \item Cookies vs. Cores
+        \begin{enumerate}
+            \item The process remains mainly the same for capturing cores and cookies.
+            \item The main difference would be that if you can see the entire width of the core in one image, make the height OR width equal to zero.
+            \item Because a core holder is very close in height to the core itself, this can mess up the focusing control. 
+            \item Before doing a lot of cores, test out one at a time to dial in settings and confirm that the stitch is working alright.
+        \end{enumerate}
+	\end{enumerate}
 \item Potential errors
@@
         \end{enumerate}
-    \item 
+    \item Failing before capturing all samples
+        \begin{enumerate}
+            \item If this occurs, you'll have to check which cookies were completed and remove them from the table. Restart the adding samples step. 
+            \item If this happens a lot, let Adam know.
         \end{enumerate}
+    \item Unresponsive GUI
+        \begin{enumerate}
+            \item Close the terminal. This will force everything to close and end the process that is causing the delay. If this keeps happening let Adam know.
+        \end{enumerate}
+
+    \end{enumerate}
+\item Figures 
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/jetson.jpg}
+        \caption{Jetson Orin Nano with all the ports connected.}
+        \label{fig:jetson1}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/motor_controller.jpg}
+        \caption{Blackbox X32 Motor Controller from above.}
+        \label{fig:controller1}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/motor_controller_connections.jpg}
+        \caption{Blackbox X32 Motor Controller fully connected ports.}
+        \label{fig:controller2}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/motor_controller_buttons.jpg}
+        \caption{Blackbox X32 Motor Controller buttons.}
+        \label{fig:controller3}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/power_supply.jpg}
+        \caption{Power supply fully connected.}
+        \label{fig:power1}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/sample_before_level.jpg}
+        \caption{Cookie with the bubble of the level pointed directly towards the side of the table with one screw.}
+        \label{fig:level1}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/sample_levelled.jpg}
+        \caption{Cookie with the bubble of the level centered.}
+        \label{fig:level2}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/gui.jpg}
+        \caption{Default GUI appearance after launching.}
+        \label{fig:gui1}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/background.jpg}
+        \caption{Background after logging into the Jetson.}
+        \label{fig:background1}
+    \end{figure}
+
+    \begin{figure}
+        \includegraphics[width=\linewidth]{../content/terminal_commands.jpg}
+        \caption{Background after logging into the Jetson.}
+        \label{fig:terminal1}
+    \end{figure}
 

### 2024-10-04 docs/writing/paper.tex
Adding basic tex stub with bibliography

% !TeX root = paper.tex
\documentclass[a4paper,12pt]{article}

% Packages
\usepackage[utf8]{inputenc}     % For UTF-8 encoding
\usepackage{amsmath, amssymb}   % For math symbols
\usepackage{graphicx}           % For including graphics
\usepackage{geometry}           % For setting page margins
\usepackage{natbib}             % For citations
\usepackage{hyperref}           % For clickable links in references

% Page layout
\geometry{margin=1in}           % Set margins to 1 inch

% Title section
\title{Tina Paper In Development}
\author{Your Name}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
    This is the abstract of your paper. Briefly describe the purpose of the research, the main results, and the conclusions.
\end{abstract}

\section{Introduction}
This is the introduction section where you introduce the topic and cite some papers \citep{mohammadi_comparative_2024}.

\section{Methods}
Describe your methods here. You can use equations like this:

\section{Results}
Present your results here.

\section{Discussion}
Discuss the implications of your results here.

\section{Conclusion}
Summarize your key findings here.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}

### 2024-10-06 docs/writing/paper.tex
Adding paper boilerplate
Adding the beginnings of a diagram to show overlapping images and image focusing

@@
 
+\subsection{Hardware Design}
+
+\subsubsection{Mechanical Structure}
+The cartesian robot was purchased from an open source parts supplier, OpenBuilds. They sell various sets of gantry robots, motor controllers, and power supplies
+which can be conveniently purchased from one supplier. All metal components are cut and sized, and have very detailed assembly videos, an active community forum,
+and customer support to fall back on. The OpenBuilds ACRO system circumvents the need for most power tools and machining that would otherwise be necessary to build a robust robot.
+All other parts were designed using SolidWorks and printed on an FDM 3D printer using ABS material. Mechanical assembly can be completed entirely with common hand tools and metric fasteners.
+
+\subsubsection{Computer}
+The computer which runs the graphical user interface (GUI), gantry control, camera streaming, and image stitching is the NVIDIA Jetson Orin Nano Developer Kit (Orin Nano). Using a computer
+which is compatible with CSI cameras such as the Raspberry Pi HQ Camera was essential to maintain cost effectiveness. The Orin Nano, like others in the Jetson line of products, has hardware
+accelerators for common GPU tasks such as image streaming, which is vital to retain CPU power for GUI tasks and robot control.
+
+\subsubsection{Camera}
+Unlike the Gigapixel or CaptuRING, the camera is a machine vision camera rather than a handheld professional camera. Choosing the Raspberry Pi HQ Camera provided many lens options, significant
+forum support, and includes an Obsolescence Statement that the camera will be in production until January 2030.
+Combined with a 180mm C-mount microscope lens, the images can reach above 20,000 dots per inch (DPI).
+
+\subsection{Software Design}
+Software design here.
+
 \section{Results}
@@
 
+\subsection{Scans of Cookies and Cores}
+
+\subsection{Functional Limits}
+
 \section{Discussion}
 Discuss the implications of your results here.
+
+\subsection{Strengths and Opportunities}
+
+\subsection{Opportunities for Improvement}
 

### 2024-10-06 docs/writing/paper.tex
Adding the beginnings of a diagram to show overlapping images and image focusing

@@
 \subsection{Software Design}
-Software design here.
+All software was written in Python VERSION and a variety of packages for the GUI, image stitching, and calculations (Should I include all the packages in writing?). Care was taken to segment the
+software design into abstracted object oriented code.
+
+\subsubsection{Graphical User Interface}
+The GUI to control the machine launches from a Python script. Included are a viewer to see the live stream from the camera, buttons to jog the machine throughout the XYZ coordinate system,
+setting the size of samples, and running the process to capture and stitch images. The system was designed to allow for multiple samples to be prepared in a queue for bulk digitization.
+
+\subsubsection{Sample Digitization}
+The digitization of a sample can be reduced to a few processes - capturing a grid of overlapping images, image focusing, and image stitching.
 

### 2024-10-07 docs/writing/paper.tex
Continuing the software design portion

@@
 which can be conveniently purchased from one supplier. All metal components are cut and sized, and have very detailed assembly videos, an active community forum,
-and customer support to fall back on. The OpenBuilds ACRO system circumvents the need for most power tools and machining that would otherwise be necessary to build a robust robot.
+and customer support to fall back on.
+The OpenBuilds ACRO system circumvents the need for most power tools, machining, electrical connectors, and technical knowhow that would otherwise be necessary to build a robust robot.
 All other parts were designed using SolidWorks and printed on an FDM 3D printer using ABS material. Mechanical assembly can be completed entirely with common hand tools and metric fasteners.
@@
 
+Inside the GUI, the program is populated with the height and width of the sample as well as the height and width of the image. The user then navigates to the center of the sample and can choose 
+to begin imaging the sample or add the sample to a queue for bulk digitization. This provides the machine enough data to calculate the necessary rows and columns for traversing the entire surface
+area of the sample while maintaining a level of overlap between adjacent images. Image overlap is what allows for images to be stitched together as there is nonzero error when the robot translates 
+from one coordinate to another.
+
+To stitch adjacent images, feature based image stitching from a Python package Stitch2D was used. The underlying functions were based on the processing libraries
+OpenCV and NumPy.
+Multiple other image stitching packages were tested but none were able to successfully piece together our samples.
+(***Sounds kinda bad to not have data on a comparison as well as not having a metric to see how 'well' an image is stitched...
+but this is non trivial.
+Can we get away with it?***)
+
+One major difficulty with using the Raspberry Pi HQ Camera is the lack of a lens with auto-focus. An in focus image in this regime is dependent on the focal length of the lens set by the user and the
+distance from the sample.
+The microscope lens has a very small depth of field and goes from in-focus to out-of-focus in less than a 1mm step in the distance to the sample.
+Two cooperative solutions were implemented
+to achieve a focused image for the entire surface area of the sample - taking images from multiple distances from the sample and automatic control.
+
+\subsubsection{Image Stack Focusing}
+When the sample is placed on the table to be digitized, it is assumed that there will, at best, very slight variations in height across the surface area of the sample. A levelling table 
+assists in reducing this error although it will never be nonzero. At each location (X, Y) on the grid of overlapping images, 11 images are captured in equally distanced Z steps across a 1mm range. 
+Each of these images have a normalized variance score calculated - the maximum value represents the most in focus image \citep{sampat_extensive_2014}.
+
+\subsubsection{Automatic Control}
+
+$$\frac{1}{MN\mu} \sum\limits_{x=0}^{M-1} \sum\limits_{y=0}^{N-1}(f(x,y) - \mu)^2$$
+
 \section{Results}

### 2024-10-07 docs/writing/paper.tex
Adding some about automatic control

@@
 Each of these images have a normalized variance score calculated - the maximum value represents the most in focus image \citep{sampat_extensive_2014}.
+In practice, stopping and starting the motor causes
+enough vibration to require upwards to two seconds of pause time before the motion blur stops to take each photo. Instead of stopping and starting each image, an additional distance is added to the 
+Z-motion to allow the stepper to accelerate and reach constant velocity.
+At this speed, images are then taken with constant time steps instead of constant distance steps across the 1mm range.
+This eliminates
+the motion blur due to acceleration and dramatically reduces imaging time.
+
+$$\frac{1}{MN\mu} \sum\limits_{x=0}^{M-1} \sum\limits_{y=0}^{N-1}(f(x,y) - \mu)^2$$
 
 \subsubsection{Automatic Control}
-
-$$\frac{1}{MN\mu} \sum\limits_{x=0}^{M-1} \sum\limits_{y=0}^{N-1}(f(x,y) - \mu)^2$$
+While the image stack provides a basis to obtain an in focus image, it is statically limited to the 1mm range in Z-values. Instead of spending significant time guaranteeing that the sample
+is within the Z-value range across the entire surface area, a PID control algorithm takes the wheel. The information previously calculated in the image stack is sufficient 
+to inform the PID controller of how to adjust the initial Z-value for the next image stack. The controller is constantly trying to make the most in focus image be at the center of the image stack,
+index value 5. This allows for the most flexibility moving to the next location of the image grid.
 

### 2024-10-07 docs/writing/paper.tex
Adding PID controller paper

@@
 While the image stack provides a basis to obtain an in focus image, it is statically limited to the 1mm range in Z-values. Instead of spending significant time guaranteeing that the sample
-is within the Z-value range across the entire surface area, a PID control algorithm takes the wheel. The information previously calculated in the image stack is sufficient 
+is within the Z-value range across the entire surface area, a PID control algorithm takes the wheel \citep{odwyer_summary_2000}. The information previously calculated in the image stack is sufficient 
 to inform the PID controller of how to adjust the initial Z-value for the next image stack. The controller is constantly trying to make the most in focus image be at the center of the image stack,

### 2024-10-09 docs/writing/paper.tex
Slowly chipping away

--- 2024-10-07 docs/writing/paper.tex
@@
 \usepackage{hyperref}           % For clickable links in references
+\usepackage{subcaption}         % For side by side subfigures
 
@@
 \section{Introduction}
-This is the introduction section where you introduce the topic and cite some papers \citep{mohammadi_comparative_2024}.
+This is the introduction section where you introduce the topic and cite some papers.
 
 \section{Methods}
-Describe your methods here. You can use equations like this:
+The functionality of this system allows for the major steps in tree cookie and core digitization to be automated.
+Capturing images and stitching them together is handled without the need for user intervention.
+To do this, the system traverses the surface area of the sample after being given its dimensions.
+After the capturing process, the images are stitched and are stored on the device along with metadata of relevant
+settings and sample parameters.
 
@@
 \subsubsection{Automatic Control}
-While the image stack provides a basis to obtain an in focus image, it is statically limited to the 1mm range in Z-values. Instead of spending significant time guaranteeing that the sample
-is within the Z-value range across the entire surface area, a PID control algorithm takes the wheel. The information previously calculated in the image stack is sufficient 
-to inform the PID controller of how to adjust the initial Z-value for the next image stack. The controller is constantly trying to make the most in focus image be at the center of the image stack,
+While the image stack provides a basis to obtain an in focus image, it is statically limited to the 1mm range in the Z axis. Instead of spending significant time guaranteeing that the sample
+is within the the Z axis range across the entire surface area, a PID control algorithm takes the wheel \citep{odwyer_summary_2000}.
+The information previously calculated in the image stack is sufficient
+to inform the PID controller of how to adjust the initial the Z axis for the next image stack. The controller is constantly trying to make the most in focus image be at the center of the image stack,
 index value 5. This allows for the most flexibility moving to the next location of the image grid.
 
+\begin{figure}
+    \centering
+    \begin{subfigure}{.5\textwidth}
+      \centering
+      \includegraphics[height=0.75\linewidth]{../diagrams/sample_setup_ideal.png}
+      \caption{Ideal sample leveling}
+      \label{fig:sub1}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \centering
+      \includegraphics[height=0.75\linewidth]{../diagrams/sample_setup_realistic.png}
+      \caption{Realistic sample leveling}
+      \label{fig:sub2}
+    \end{subfigure}
+    \caption{Side view of the camera and sample on top of a leveling table.
+The ideal sample leveling shows a uniform distance d at all (x,y) coordinates on the sample.
+This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique (x,y) coordinates.}
+    \label{fig:test}
+\end{figure}
+
 \section{Results}
-Present your results here.
-
 \subsection{Scans of Cookies and Cores}
+The microscope lens greatly improved the maximum resolution of the digitization. Resolutions of up to 13,400 DPI were achieved, a large improvement when compared to 
+both high resolution flatbed scanners (Epson® Perfection v750 PRO) and CaptuRING.
 

### 2024-10-09 docs/writing/paper.tex
Adding a little bit in functional limits

@@
 \subsection{Functional Limits}
+With such high resolution, multiple logistics concerns arise. First is the file size. With such a high resolution, the size of the images can become extremely large. A cookie
+5 inches in diameter would result in an image with 67,000 pixels squared. This easily exceeds the 2.5 GB maximum file size in a TIFF - most software for viewing the file will also 
+be incompatible with any file this big as well.
 

### 2024-10-10 docs/writing/outline.tex
Changing itemize to be enumerate
Adding paper boilerplate
Adding more to the outline

@@
 \section {Paper Outline}
-\begin{outline}[itemize]
+\begin{outline}[enumerate]
 
 \1 Abstract
-\begin{itemize}
+\begin{enumerate}
 	\item Flatbed scanner alternative with improved resolution to digitizing tree cookies and cores
@@
 	\item Tina is open source and able to be assembled with 3D printed parts, OpenBuilds parts kits, and hand tools
-\end{itemize}
+\end{enumerate}
 \item Introduction
-	\begin{itemize}
+	\begin{enumerate}
 	\item Tree ring science overview
-		\begin{itemize}
+		\begin{enumerate}
 		\item **Lizzie add here**
@@
 		\item understanding tree growth mechanisms
-		\end{itemize}
+		\end{enumerate}
 	\item Development of scanning alternatives
-		\begin{itemize}
+		\begin{enumerate}
 		\item beginning with ATRICS
@@
 		\item DIY CaptuRING
-		\end{itemize}
+		\end{enumerate}
 	\item Main reasons to build this 
-		\begin{itemize}
+		\begin{enumerate}
 		\item further develop the affordable / open source alternatives for scientific equipment
@@
 		\item Digitizing wood samples for better data pooling / archiving
-		\end{itemize}
+		\end{enumerate}
 	\item Software can be used to help record annual growth 
-		\begin{itemize}
+		\begin{enumerate}
 		\item Coorecorder
 		\item Various ML based methods that Sandy has researched
-		\end{itemize}
-	\end{itemize}
+		\end{enumerate}
+	\end{enumerate}
 \item Materials and Methods
-		\begin{itemize}
+		\begin{enumerate}
 		\item Hardware Design
-			\begin{itemize}
+			\begin{enumerate}
 			\item Cartesian gantry robot kit from OpenBuilds (ACRO system)
 			\item Camera 
-				\begin{itemize}
+				\begin{enumerate}
 				\item chose Raspberry Pi HQ Camera for C-mount lens options, ability to stream, cost effectiveness, and clear obscelescence statement
-				\end{itemize}
+				\end{enumerate}
 			\item computer
-				\begin{itemize}
+				\begin{enumerate}
 				\item NVIDIA Jetson Orin Nano for processing power in a light package, can connect to Raspberry Pi camera. Powerful enough to power a GUI and monitor display as well
 				\item NVMM Hardware accelerated video streaming reduces CPU load 
-				\end{itemize}
+				\end{enumerate}
 			\item 3D printed components
-				\begin{itemize}
+				\begin{enumerate}
 				\item 3D printed parts from a Bambu X1C were designed to mount the lens and computer to the ACRO system
@@
 				\item levelling tables were designed to allow for the cookie samples to be arranged orthogonally to the lens. Imperative to capturing in focus images across the entire cookie
-				\end{itemize}
-			\end{itemize}
+				\end{enumerate}
+			\end{enumerate}
 		\item Software Design
-			\begin{itemize}
+			\begin{enumerate}
 			\item image focusing
-				\begin{itemize}
+				\begin{enumerate}
 				\item sample levelling is not perfect.
@@
 				\item automatic PID control was implemented to adjust the initial Z-coordinate for each set of images to increase the probability of having an in focus image at each image set
-				\end{itemize}
+				\end{enumerate}
 			\item grid traverse
-				\begin{itemize}
+				\begin{enumerate}
 				\item The software traverses the entire surface area of the sample. Each row and column is overlapping with its adjacent row and column. Allowing for feature based image sttiching.
-				\end{itemize}
+				\end{enumerate}
 			\item feature based image stitching
-				\begin{itemize}
+				\begin{enumerate}
 				\item An implementation, Stitch2D, of feature based OpenCV image stitching was used.
 Edits to the code were made to allow for more memory efficient methods of stitching with the use of NumPy memory maps
-				\end{itemize}
+				\end{enumerate}
 			\item user interface
-				\begin{itemize}
+				\begin{enumerate}
 				\item A GUI was made to navigate the machine to samples, set the size of the sample and its center. After samples are set up the operation is passive.
-				\end{itemize}
-			\end{itemize}
-		\end{itemize}
+				\end{enumerate}
+			\end{enumerate}
+		\end{enumerate}
 
 \item Results
-	\begin{itemize}
+	\begin{enumerate}
 	\item scans of cookies / cores
-		\begin{itemize}
+		\begin{enumerate}
 		\item ultra high resolution scans (DPI 15,000 +)
@@
 		\item large surface areas can produce files that are inconveniently large
-		\end{itemize}
+		\end{enumerate}
 	\item File-size results in functional limits 
-		\begin{itemize}
+		\begin{enumerate}
 		\item max filesize for TIFF files is 2.5 GB
@@
 		\item workaround with NumPy memory maps and cropped viewing 
-		\end{itemize}
-	\end{itemize}
+		\end{enumerate}
+	\end{enumerate}
 \item Discussion (for MEE Only)
-	\begin{itemize}
+	\begin{enumerate}
 	\item strengths and opportunities
-		\begin{itemize}
+		\begin{enumerate}
 		\item  Ability to capture multiple cores / cookies simultaneously
@@
 		\subitem although final quality is only as good as the sample preparation
-		\end{itemize}
+		\end{enumerate}
 	\item opportunities for improvements
 		\item Lenses
-			\begin{itemize}
+			\begin{enumerate}
 			\item troubles with fixed aperature / non autofocusing / poorly manufactured lens
-				\begin{itemize}
+				\begin{enumerate}
 				\item lack of control to increase sharpness
@@
 				\item poor lens manufacturing can lead to drastically different quality / lighting between the edges and center of an image
-				\end{itemize}
+				\end{enumerate}
 			\item good lens can increase sharpness and therefore stitching accuracy and ML inference
@@
 			\subitem poor lens quality seemed to be the culprit of producing visible seams when stitching with larger field of view images 
-			\end{itemize}
-
+			\end{enumerate}
+		\item tested on R-CNN
+			\begin {enumerate}
+			\item Cookie with a dpi of SMALL and dpi of 13500 were tested. With total pixel count of SMALL and BIG
+			\item tree rings were successfully measured with the metadata supplied from the machine but the SMALL took X minutes and big took Y minutes on a computer with Z specifications 
+			\end {enumerate}
+		\item tested on CooRecorder + CDendro
+			\begin {enumerate}
+			\item Maximum filesize 
+			\item Coordinates were registered on to the cookie scans 
+			\item Note to self: using CDendro for multiple samples from the same tree... (cookies) 
+			\end {enumerate}
 		\item focus stacking
@@
 		\item Autofocusing with hardware would significantly decrease the time to capture a cookie (using a professional camera)
-	\end{itemize}
+		\item Software support for BigTIFF 
+			\begin {enumerate}
+			\item Not all software can support this, especially if the computer is RAM constrained. Ring analysis tool which uses BigTIFF and loads only partial bits of the full image would be useful
+			\end {enumerate}
+	\end{enumerate}
 \item Conclusions (optional for MEE)
-	\begin{itemize}
+	\begin{enumerate}
 	\item
-	\end{itemize}
+	\end{enumerate}
 

### 2024-10-16 docs/writing/outline.tex
Refining outlien

@@
 	\item opportunities for improvements
+		\begin{enumerate}
+		\item focus stacking
+			\begin {enumerate}
+			\item when a sample is microtomed, the vessels are hollow. The autofocusing algorithm can be confused from this 
+			\end{enumerate}
 			\item Lenses
@@
 				\item troubles with fixed aperature / non autofocusing / poorly manufactured lens
-				\begin{enumerate}
 				\item lack of control to increase sharpness
@@
 				\item poor lens manufacturing can lead to drastically different quality / lighting between the edges and center of an image
-				\end{enumerate}
 				\item good lens can increase sharpness and therefore stitching accuracy and ML inference
 				\item obtain similar quality between the corner and center of the image
-			\subitem poor lens quality seemed to be the culprit of producing visible seams when stitching with larger field of view images 
+				\item Difficulty stitching with lower detail images, can't increase digitizing speed, must downscale high resolution images 
+				\end{enumerate}
+		\end{enumerate}
+		\begin{enumerate}
+		\item poor lens quality seemed to be the culprit of producing visible seams when stitching with larger field of view images 
 		\end{enumerate}
 	\item tested on R-CNN
-			\begin {enumerate}
+		\begin{enumerate}
 		\item Cookie with a dpi of SMALL and dpi of 13500 were tested. With total pixel count of SMALL and BIG
-			\item tree rings were successfully measured with the metadata supplied from the machine but the SMALL took X minutes and big took Y minutes on a computer with Z specifications 
-			\end {enumerate}
+			\begin{enumerate}
+			\item NEEDS specific: DPI, height and width of pixels, record runtime
+			\item NEEDS specifications of server computational power (GPU count/other metric for ANN power)
+			\end{enumerate}
+		\end{enumerate}
 	\item tested on CooRecorder + CDendro
-			\begin {enumerate}
+		\begin{enumerate}
 		\item Maximum filesize 
@@
 		\item Note to self: using CDendro for multiple samples from the same tree... (cookies) 
-			\end {enumerate}
-		\item focus stacking
-		\subitem when a sample is microtomed, the vessels are hollow. The autofocusing algorithm can be confused from this 
-		\item Difficulty stitching with lower detail images, can't increase digitizing speed, must downscale high resolution images 
-		\item Autofocusing with hardware would significantly decrease the time to capture a cookie (using a professional camera)
-		\item Software support for BigTIFF 
-			\begin {enumerate}
-			\item Not all software can support this, especially if the computer is RAM constrained. Ring analysis tool which uses BigTIFF and loads only partial bits of the full image would be useful
-			\end {enumerate}
+		\end{enumerate}
+	\item Software support for large files 
+		\begin{enumerate}
+		\item Not all software can support this, especially if the computer is RAM constrained.
+Ring analysis tool which uses BigTIFF, HD5, or memory maps and loads only partial portions of the full image would be useful
+		\end{enumerate}
 	\end{enumerate}
@@
 	\begin{enumerate}
-	\item
+	\item empty
 	\end{enumerate}

### 2024-10-18 code/mosaic.py
Adding batch size control to alignment, and feature extraction

@@ -1801,7 +1801,7 @@ class MemmapStructuredMosaic(MemmapMosaic):
                 tile.col = x
                 tile.grid = self.grid
 
-    def align(self, origin=None, limit=None, **kwargs):
+    def align(self, origin=None, limit=None, batch_size = 10, **kwargs):
         """Builds a mosaic outward from a single tile using feature matching
 
         Parameters
@@ -1841,18 +1841,25 @@ class MemmapStructuredMosaic(MemmapMosaic):
                 unique.setdefault(tile.id, tile)
                 for neighbor in tile.neighbors().values():
                     unique.setdefault(neighbor.id, neighbor)
+            
+             # Control batch size
             batch = [t for t in unique.values() if not t.features_detected]
-            self._batch_tile_method("detect_and_extract", batch=batch)
-
+            for i in range(0, len(batch), batch_size):
+                batch_slice = batch[i:i + batch_size]
+                self._batch_tile_method("detect_and_extract", batch=batch_slice)
+            
             # Batch align adjacent tiles
             unique = {}
             for tile in tiles:
                 for neighbor in tile.neighbors().values():
                     unique.setdefault(neighbor.id, (tile, neighbor))
+
+            # Control batch size
             batch = [(t, n) for t, n in unique.values() if not n.placed]
-            self._batch_tile_method(
-                "align_to", [t for t, _ in batch], batch=[n for _, n in batch], **kwargs
-            )
+            for i in range(0, len(batch), batch_size):
+                batch_slice = batch[i:i + batch_size]
+                self._batch_tile_method("align_to", [t for t, _ in batch_slice], batch=[n for _, n in batch_slice], **kwargs)
+
 
             # Otherwise re-run the loop with the next group of tiles
             tiles = [n for _, n in batch if n.placed]

### 2024-10-21 code/mosaic.py
Making super ugly changes to the Stitch2d architecture to heavily reduce RAM usage. No more storing data in each tile, load the data from a memmap each time you need it.

@@ -13,6 +13,7 @@ import numpy as np
 
 from tile import Tile, OpenCVTile, MemmapOpenCVTile, MemmapTile
 
+from memory_profiler import profile
 
 logger = logging.getLogger(__name__)
 
@@ -1789,7 +1790,7 @@ class MemmapStructuredMosaic(MemmapMosaic):
                 origin=origin,
                 direction=direction,
                 pattern=pattern,
-                fill_value=self.placeholder(tiles[0]),
+                # fill_value=self.placeholder(tiles[0]),
             )
 
         self._verify_tiles()
@@ -1848,6 +1849,10 @@ class MemmapStructuredMosaic(MemmapMosaic):
                 batch_slice = batch[i:i + batch_size]
                 self._batch_tile_method("detect_and_extract", batch=batch_slice)
             
+            # for id, ti in unique.items():
+            #     if ti.features_detected:
+            #         ti.refresh_memmap()
+
             # Batch align adjacent tiles
             unique = {}
             for tile in tiles:
@@ -1859,7 +1864,7 @@ class MemmapStructuredMosaic(MemmapMosaic):
             for i in range(0, len(batch), batch_size):
                 batch_slice = batch[i:i + batch_size]
                 self._batch_tile_method("align_to", [t for t, _ in batch_slice], batch=[n for _, n in batch_slice], **kwargs)
-
+                
 
             # Otherwise re-run the loop with the next group of tiles
             tiles = [n for _, n in batch if n.placed]

### 2024-10-21 code/stitcher.py

@@ -11,6 +11,8 @@ import glob
 import gc
 import imutils
 import logging as log
+from memory_profiler import profile
+import sys 
 
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
@@ -35,7 +37,7 @@ class Stitcher:
 
         self.load_metadata()
         
-        
+    @profile
     def write_dats(self, dats_path, resize = None):
         # Create dats directory 
         if not os.path.exists(dats_path):
@@ -43,26 +45,60 @@ class Stitcher:
 
         i = 0
         tile_paths = self.get_frames()
+        self._memmap_paths = []
+        first = True
         for tile_path in tile_paths:
             with rasterio.open(os.path.join(self._frame_dir, tile_path)) as dataset:
                 data = np.transpose(dataset.read(), (1, 2, 0))
                 if resize is not None and resize > 0 and resize < 1.0:
                     data = imutils.resize(data, width = int(resize * data.shape[1]))
-                memmap_array = np.memmap(os.path.join(dats_path, "{}_".format(i) + "memmap_array.dat"), dtype=dataset.dtypes[0], mode='w+', shape=(data.shape[0], data.shape[1], data.shape[2]))
+
+                self._memmap_shape = (data.shape[0], data.shape[1], data.shape[2])
+
+                # create a placeholder dat 
+                if first:
+                    memmap_path = os.path.join(dats_path, "placeholder_memmap_array.dat")
+                    memmap_array = np.memmap(memmap_path, dtype = dataset.dtypes[0], mode='w+', shape=(data.shape[0], data.shape[1], data.shape[2]))
+                    data = np.full((data.shape[0], data.shape[1], data.shape[2]), 0, dataset.dtypes[0])
+                    memmap_array[:] = data[:]
+                    memmap_array.flush()
+                    del memmap_array
+                    first = False
+
+                memmap_path = os.path.join(dats_path, "{}_".format(i) + "memmap_array.dat")
+                self._memmap_paths.append(memmap_path)
+                self._memmap_dtype = dataset.dtypes[0]
+                memmap_array = np.memmap(memmap_path, dtype=dataset.dtypes[0], mode='w+', shape=(data.shape[0], data.shape[1], data.shape[2]))
                 i+=1
-                memmap_array[:] = data
-                self._memmaps.append(memmap_array)
[…]

### 2024-10-21 code/tile.py

@@ -793,13 +793,15 @@ class MemmapTile:
     #: dict : maps strings to a subclass-specific feature matcher
     matchers = {}
 
-    def __init__(self, data, detector="sift"):
+    def __init__(self, path, shape, detector="sift"):
         """Initializes a mosaic from a list of tiles
 
         Parameters
         ----------
-        data : str or numpy.ndarray
+        data : str or numpy.memmap
             path to an image file or an array of image data
+        path: str
+            path to the memmap 
         detector : str
             name of the detector used to find/extract features. Currently
             only sift is supported.
@@ -807,11 +809,11 @@ class MemmapTile:
 
         self.id = uuid.uuid4()
 
-        self.source = data 
-        self.imdata = data if isinstance(self.source, np.memmap) else print("NOT np.memmap")
-
-        if self.imdata is None:
-            raise IOError(f"No image data found (source={self.source})")
+        self.source = path
+        self.path = path
+        self.shape = shape
+        # if self.imdata is None:
+        #     raise IOError(f"No image data found (source={self.source})")
 
         self.row = None
         self.col = None
@@ -936,21 +938,6 @@ class MemmapTile:
         return self.imdata.dtype
 
     @property
-    def shape(self):
-        """Gets the shape of the image"""
-        return self.imdata.shape
-
-    @property
-    def size(self):
-        """Gets the size of the image"""
-        return self.imdata.size
-
-    @property
-    def mp(self):
-        """Gets the size of the image in megapixels"""
-        return self.imdata.size / 1e6
-
-    @property
     def placed(self):
         """Whether the tile has been assigned coordinates in the mosaic"""
         return self.y is not None and self.x is not None
@@ -967,44 +954,10 @@ class MemmapTile:
             return self.source.copy()
         raise ValueError("Data must be a numpy array")
 
-    def copy(self):
-        """Creates a copy of the tile
-
-        Parameters
-        ----------
-        grid: list of lists
-            grid from the mosaic containing the tile
-
-        Returns
-        -------
-        Mosaic
-            copy of the tile
-        """
-        source = self.source
-        if not isinstance(self.source, str):
-            source = self.source.copy()
-
-        copy = self.__class__(source)
-
-        copy.imdata = self.imdata.copy()
-        copy.scale = self.scale
-
-        copy.row = self.row
-        copy.col = self.col
-
-        copy.y = self.y
[…]

### 2024-10-21 code/stitcher.py
Commenting out profiler

@@ -11,7 +11,7 @@ import glob
 import gc
 import imutils
 import logging as log
-from memory_profiler import profile
+# from memory_profiler import profile
 import sys 
 
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
@@ -37,7 +37,7 @@ class Stitcher:
 
         self.load_metadata()
         
-    @profile
+    # @profile
     def write_dats(self, dats_path, resize = None):
         # Create dats directory 
         if not os.path.exists(dats_path):
@@ -86,7 +86,7 @@ class Stitcher:
         gc.collect()
         print("Wrote dats")
 
-    @profile
+    # @profile
     def read_dats(self):
         del self._memmaps
 
@@ -171,7 +171,7 @@ class Stitcher:
             )
         
         return sorted_paths
-    @profile
+    # @profile
     def stitch(self, resize=None):
         if resize is not None and resize < 1.0 and resize > 0:
             path = os.path.join(self._frame_dir, "{}per".format(int(resize * 100)))

### 2024-10-21 code/mosaic.py
Reducing memory more by deleting the keypoints and descriptors of tiles that have had all of their neighbors placed. This did not account for tiles with less than 4 neighbors

@@ -1801,7 +1801,7 @@ class MemmapStructuredMosaic(MemmapMosaic):
                 tile.row = y
                 tile.col = x
                 tile.grid = self.grid
-
+    # @profile
     def align(self, origin=None, limit=None, batch_size = 10, **kwargs):
         """Builds a mosaic outward from a single tile using feature matching
 
@@ -1849,9 +1849,6 @@ class MemmapStructuredMosaic(MemmapMosaic):
                 batch_slice = batch[i:i + batch_size]
                 self._batch_tile_method("detect_and_extract", batch=batch_slice)
             
-            # for id, ti in unique.items():
-            #     if ti.features_detected:
-            #         ti.refresh_memmap()
 
             # Batch align adjacent tiles
             unique = {}
@@ -1868,6 +1865,17 @@ class MemmapStructuredMosaic(MemmapMosaic):
 
             # Otherwise re-run the loop with the next group of tiles
             tiles = [n for _, n in batch if n.placed]
+
+            # If all of the neighbors are placed, delete the keypoints and descriptors to save memory
+            for t, _ in batch:
+                neighbors = t.neighbors()
+
+                if ("top" in neighbors and "left" in neighbors and "bottom" in neighbors and "right" in neighbors):
+                    if (neighbors["top"].placed and neighbors["left"].placed and neighbors["bottom"].placed and neighbors["right"].placed):
+                            # del t.descriptors
+                            # del t.keypoints 
+                            t.descriptors = np.array([0])
+                            t.keypoints = ()
 
         else:
             if limit is not None:

### 2024-10-22 docs/writing/outline.tex
Edits to outline and gitignore

@@
 	\item Current alternatives are very expensive (\$70,000USD), lack a high level of detail (printer scanners), or cannot scan cookies (CaptuRING)
-	\item Tina is open source and able to be assembled with 3D printed parts, OpenBuilds parts kits, and hand tools
+	\item Tina is open source and able to be assembled with 3D printed parts, OpenBuilds parts kits, and hand tools % Can we add how much ours costs?
 \end{enumerate}
@@
 		\begin{enumerate}
-		\item **Lizzie add here**
+		\item tree rings provide critical data/insights across multiple fields
 		\item tree rings can help reconstruct past climates 
-		\item understanding tree growth mechanisms
+		\item understanding tree growth mechanisms (plant bio/ecology)
+		\end{enumerate}
+	% Consider adding this to below points (Development of scanning alternatives) as one paragraph on its own
+	\item Getting these data means measuring tree rings -- quick history
+		\begin{enumerate}
+		\item originally done on scopes
+		\item images important for several decades now (WinDendro?) 
+		\item But new methods coming up often to take advantage of better imaging technology ... 
+		% \item Could add here one sentence on the three methods you mention below depending on what you want to say ... 
 		\end{enumerate}
@@
 		\end{enumerate} 
+	% I suggest a more 'here we ....' paragraph next that includes the below points so something like: (comment this in if you like it and delete your section below
+	\iffalse
+	\item Here we tell you about our new device
+		\begin{enumerate}
+		\item We review hardware and software % Could add total estimated cost?
+		\item Then show how the method can be used to record annual growth through current common programs pr ...
+		\item Various ML based methods that Sandy has researched
+		\end{enumerate}
+	\fi
 	\item Software can be used to help record annual growth 
@@
 	\item strengths and opportunities
-		\begin{enumerate}
+		\begin{enumerate} % Usually I like a quick overview paragraph and I think you should still have one but as currently written I think the 'potential for automatic tree ring identification with iamge processing / ML ' and 'potential for vessel counts for an entire ring' connect too much to later sections, so you should move those sections up
 		\item  Ability to capture multiple cores / cookies simultaneously
 		\item cost effective
+		% I would move the section on  tested on CooRecorder + CDendro here then 
+		% Section on R-CNN here then ... a paragraph on potential for two things below (or you could have these points on potential as part of your conclusions)
 		\item potential for automatic tree ring identification with iamge processing / ML 
@@
 	\end{enumerate}
-\item Conclusions (optional for MEE)
+\item Conclusions (optional for MEE) % Usually I am anti-conclusions because you just end up repeating yourself, but I think this could work here, especially to end on a positive after the 'opportunities' section.
+I think you could add  the 'potential for automatic tree ring identification with iamge processing / ML ' and 'potential for vessel counts for an entire ring'  here perhaps.
 	\begin{enumerate}

### 2024-10-25 docs/writing/outline.tex
Tried to explain the 'why' before explaining the 'how' for the design decisions, haven't finished discussion yet

@@
 \usepackage{enumitem}
+\usepackage{graphicx}
 \setenumerate[1]{label=\Roman*.}
@@
 		\item But new methods coming up often to take advantage of better imaging technology ... 
+		\begin{enumerate}
+			\item vessel counting requires detail at the sub-vessel level % detail is only as good as sample prep
+		\end{enumerate}
 		% \item Could add here one sentence on the three methods you mention below depending on what you want to say ... 
@@
 		\end{enumerate} 
-	% I suggest a more 'here we ....' paragraph next that includes the below points so something like: (comment this in if you like it and delete your section below
-	\iffalse
 	\item Here we tell you about our new device
 		\begin{enumerate}
-		\item We review hardware and software % Could add total estimated cost?
-		\item Then show how the method can be used to record annual growth through current common programs pr ...
-		\item Various ML based methods that Sandy has researched
-		\end{enumerate}
-	\fi
-	\item Software can be used to help record annual growth 
-		\begin{enumerate}
-		\item Coorecorder
-		\item Various ML based methods that Sandy has researched
+		\item We review hardware and software 
+		\item total cost
+		\item Then show how the method can be used to record annual growth through current common programs 
+		\item allows for multiple cookies / cores to be added to a queue for sampling
+		\item stitching on device 
 		\end{enumerate}
@@
 		\begin{enumerate}
+		\item Guiding design considerations
+			\begin{enumerate}
+				\item Digitization of a tree cookie and core
+				\item Minimizing barriers to assemble (Cost effective, open source, open hardware, minimal use of specialized equipment/tools)
+				\item Reducing user interaction between sample prep and the final image
+				\item Achieving near microscopic detail with a field of view as large as the sample
+			\end{enumerate}
+		\item Considerations for Digitizing Cookies
+		\begin{enumerate}
+			\item Inconsistencies in cookie preparation, (different height, non parallel faces) (figure related) %inconsistent_cookies.jpg
+			\item Cookie's surface area requires movement in the X and Y direction
+			\item The inconsistencies require movement in the Z direction to maintain in focus images
+		\end{enumerate}
 		\item Hardware Design
 			\begin{enumerate}
-			\item Cartesian gantry robot kit from OpenBuilds (ACRO system)
-			\item Camera 
-				\begin{enumerate}
-				\item chose Raspberry Pi HQ Camera for C-mount lens options, ability to stream, cost effectiveness, and clear obscelescence statement
-				\end{enumerate}
-			\item computer
-				\begin{enumerate}
-				\item NVIDIA Jetson Orin Nano for processing power in a light package, can connect to Raspberry Pi camera. Powerful enough to power a GUI and monitor display as well
-				\item NVMM Hardware accelerated video streaming reduces CPU load 
-				\end{enumerate}
-			\item 3D printed components
-				\begin{enumerate}
-				\item 3D printed parts from a Bambu X1C were designed to mount the lens and computer to the ACRO system
-				\item other parts such as a drag chain for cable management were also 3D printed
-				\item levelling tables were designed to allow for the cookie samples to be arranged orthogonally to the lens. Imperative to capturing in focus images across the entire cookie
-				\end{enumerate}
+				\item Biggest cost in most alternatives is a professional camera, choosing a Raspberry Pi HQ camera avoids this. The camera also has a definitive obscelescence statement %camera 
+				\item But to use a CSI camera, we need special hardware.
+NVIDIA Jetson Orin Nano fits the bill and has enough computational power to handle the image processing tasks while powering a monitor and GUI %computer
+				\item Ideally all non-structural components could be purchased off the shelf and assembled.
+Fortunately complex geometries, dimensional accuracy, quick iteration speed, and cost effectiveness can be achieved with 3D printing  % 3d printing
+					\subitem Parts can easily be ordered from 3D printing shops but also has a low technical barrier for personal use
+				\item to combat sample inconsistencies, a levelling table and bullseye level were designed to make the prepared surface of the cookie closer to parallel to the XY plane
+				\item for the structure, the 1m x 1m OpenBuilds ACRO system was used.
+This kit is made of strong aluminum extrusion, has 4.5 micron resolution of accuracy, and is assembled quickly by following OpenBuild's detailed video instructions.
+% cartesian gantry machine
+				\item For the Z axis a lead screw linear actuator was attached to the mounting plate of the ACRO
 			\end{enumerate}
@@
 			\begin{enumerate}
+				\item Everything was designed in Python 
+				\item GUI to load samples and their identifiers to the queue to be digitized
 				\item image focusing
 					\begin{enumerate}
-				\item sample levelling is not perfect.
-At different points on the cookie, the Z-height of the gantry must vary to capture the in focus image.
-A routine was made to take multiple images at varied Z-heights.
+					\item A routine was made to take multiple images at varied Z-heights as a proxy for autofocusing.
 The image with the highest normalized variance score was chosen as the in focus image and is kept.
-				\item automatic PID control was implemented to adjust the initial Z-coordinate for each set of images to increase the probability of having an in focus image at each image set
+%reference
+					\item Rather than moving the Z-axis, taking a photo, then moving again, the images are taken when the Z-axis is moving at constant velocity and are triggered with a time delay.
+Stopping and starting the Z-axis results in significant vibration blur in the image, requiring 2 to 3 seconds of inactivity after coming to a stop before taking an image
+					\item It's possible for the height variation to exceed the range of the multiple images taken.
+A PID automatic control algorithm was used to make adjustments to the initial Z-height of the image stack.
+This is done by adjusting to keep the focused image in the center of the image stack % reference to PID
+					\item figure
 					\end{enumerate}
-			\item grid traverse
+				\item obtaining a large field of view with a grid of small field of view images 
 					\begin{enumerate}
-				\item The software traverses the entire surface area of the sample. Each row and column is overlapping with its adjacent row and column. Allowing for feature based image sttiching.
-				\end{enumerate}
-			\item feature based image stitching
-				\begin{enumerate}
-				\item An implementation, Stitch2D, of feature based OpenCV image stitching was used.
-Edits to the code were made to allow for more memory efficient methods of stitching with the use of NumPy memory maps
-				\end{enumerate}
-			\item user interface
-				\begin{enumerate}
-				\item A GUI was made to navigate the machine to samples, set the size of the sample and its center. After samples are set up the operation is passive.
+						\item When the sample is loaded into the machine, the machine traverses the user defined bounding box of the sample
+						\item It automatically takes a grid of images where each image has overlap with its adjacent neighbors
+						\item The overlap in the images allows for the images to be aligned and combined to their neighbors.
+Stitching an image to its neighbors is done by comparing calculated key points on each image and stitching them on their matches.
+						\item A python package Stitch2D was made to stitch together a grid from a set of structured images % should I talk about the changes I made to the package here or in the discussion?
+							\subitem We made changes to the package's stitching algorithm for memory efficiency as samples with thousands of images needed more RAM than what was available
+						\item figure
 					\end{enumerate}
@@
 		\end{enumerate}
-
 \item Results
@@
 		\begin{enumerate}
-		\item ultra high resolution scans (DPI 15,000 +)
+		\item ultra high resolution scans (DPI 13,500 +)
 		\item downscaled versions as high resolution is not necessary for all applications
-		\item Digitization time as a function of surface area. Large cookies can scan only a portion of the cookie including the center and half of the rings.
+		\item Digitization time and file size as a function of surface area. Large cookies can scan only a portion of the cookie including the center and half of the rings.
 		\item large surface areas can produce files that are inconveniently large
@@
 		\end{enumerate}
+	\item Images/tables	
+		\begin{enumerate}
+		\item Scale test image, proves the generated DPI calculation negates need to image with a ruler in frame
+		\item Sample size vs image capturing time graph
+		\item Hopefully sample size, dpi, automated tree ring time
+		\item Table with results of Equation for uncompressed filesize, emphasize lossy vs lossless compression, and true filesize %image filesize = DPI x 25.4 x sample-h-mm x sample-w-mm x 8bits x 3 colors
+		\end{enumerate}
 	\end{enumerate}
@@
 	\begin{enumerate}
+	\item Quick overview (TODO)
+		\begin{enumerate}
+			\item 
+		\end{enumerate}
 	\item strengths and opportunities
@@
 		\item cost effective
-		% I would move the section on  tested on CooRecorder + CDendro here then 
-		% Section on R-CNN here then ... a paragraph on potential for two things below (or you could have these points on potential as part of your conclusions)
-		\item potential for automatic tree ring identification with iamge processing / ML 
-		\item potential for vessel counts for an entire ring 
-		\subitem although final quality is only as good as the sample preparation
+		\item tested on CooRecorder + CDendro
+			\begin{enumerate}
+			\item Maximum filesize 
+			\item Coordinates were registered on to the cookie scans 
+			\item Note to self: using CDendro for multiple samples from the same tree... (cookies) 
+			\end{enumerate}
+		\item tested on R-CNN for deep learning tree ring identification
+			\begin{enumerate}
+			\item potential for integration with a more powerful computer to automatically identify rings while imaging
+			\item potential for final stitched images to have rings automatically counted
+			\item Cookie with a dpi of SMALL and dpi of 13500 were tested. With total pixel count of SMALL and BIG
+				\subitem NEEDS specific: DPI, height and width of pixels, record runtime
+				\subitem NEEDS specifications of server computational power (GPU count/other metric for ANN power)
+			\end{enumerate}
 		\end{enumerate}
@@
 		\end{enumerate}
-	\item tested on R-CNN
-		\begin{enumerate}
-		\item Cookie with a dpi of SMALL and dpi of 13500 were tested. With total pixel count of SMALL and BIG
-			\begin{enumerate}
-			\item NEEDS specific: DPI, height and width of pixels, record runtime
-			\item NEEDS specifications of server computational power (GPU count/other metric for ANN power)
-			\end{enumerate}
-		\end{enumerate}
-	\item tested on CooRecorder + CDendro
-		\begin{enumerate}
-		\item Maximum filesize 
-		\item Coordinates were registered on to the cookie scans 
-		\item Note to self: using CDendro for multiple samples from the same tree... (cookies) 
-		\end{enumerate}
+
 	\item Software support for large files 

### 2024-10-28 docs/writing/outline.tex
Incorporating comments to outline

@@
 			\begin{enumerate}
-			\item Inconsistencies in cookie preparation, (different height, non parallel faces) (figure related) %inconsistent_cookies.jpg
+			\item Inconsistencies in cookie preparation, (different height, non parallel faces)%inconsistent_cookies.jpg
+				\subitem figure for inconsistencies
 			\item Cookie's surface area requires movement in the X and Y direction
@@
 This is done by adjusting to keep the focused image in the center of the image stack % reference to PID
-					\item figure
+				\item figure for focusing 
+				\item figure for sample setup ideal/realistic
 				\end{enumerate}
@@
 					\subitem We made changes to the package's stitching algorithm for memory efficiency as samples with thousands of images needed more RAM than what was available
-						\item figure
+				\item figure for grid traverse
 				\end{enumerate}
@@
 		\begin{enumerate}
-			\item 
+		\item X,Y,Z imaging machine which acts as a scanner 
+		\item obtains a grid of in focus images and slightly overlapping images which get stitched together to form one large mosaic 
+		\item designed for other labs with minimal engineering experience or equipment to build this for themselves
 		\end{enumerate}
@@
 		\end{enumerate}
-	\item opportunities for improvements
+	\item opportunities for improvements %combine into strengths and opportunities? 
 		\begin{enumerate}
-		\item focus stacking
-			\begin {enumerate}
-			\item when a sample is microtomed, the vessels are hollow. The autofocusing algorithm can be confused from this 
-			\end{enumerate}
-			\item Lenses
-			\begin{enumerate}
-				\item troubles with fixed aperature / non autofocusing / poorly manufactured lens
-				\item lack of control to increase sharpness
-				\item autofocusing with Z-axis movement is time consuming to do without image blur
-				\item poor lens manufacturing can lead to drastically different quality / lighting between the edges and center of an image
-				\item good lens can increase sharpness and therefore stitching accuracy and ML inference
-				\item obtain similar quality between the corner and center of the image
-				\item Difficulty stitching with lower detail images, can't increase digitizing speed, must downscale high resolution images 
-				\end{enumerate}
-		\end{enumerate}
-		\begin{enumerate}
-		\item poor lens quality seemed to be the culprit of producing visible seams when stitching with larger field of view images 
-		\end{enumerate}
-
-	\item Software support for large files 
+		\item Software support for large files than 2.5GB >
 			\begin{enumerate}
@@
 			\end{enumerate}
+		\item Integration with automatic tree ring identification model
+			\begin{enumerate}
+			\item It would be interesting to analyze the images before stitching in a different thread to potentially stitch together a binary bitmap which highlight tree rings
+			\item This could potentially drop the need for a hypercomputer if inference could be made concurrently with the images being captured
+			\end{enumerate}
+		\item Identifying vessel counts in each growth year 
 		\end{enumerate}
@@
 
+\end{enumerate}
 \end{outline}

### 2024-10-29 docs/writing/outline.tex
Adding rough references

@@
 		\begin{enumerate}
-		\item tree rings provide critical data/insights across multiple fields
-		\item tree rings can help reconstruct past climates 
-		\item understanding tree growth mechanisms (plant bio/ecology)
+		\item tree rings provide critical data/insights across multiple fields % Guibal and Guiot
+		\item tree rings can help reconstruct past climates (dendroclimatology) % Sheppard
+		\item understanding tree growth mechanisms (plant bio/ecology) % Fritts and Swetnam 
 		\end{enumerate}

### 2024-10-30 docs/writing/outline.tex
Starting to write introduction and updating outline as I write and reword

@@
 		\begin{enumerate}
-		\item originally done on scopes
-		\item images important for several decades now (WinDendro?) 
-		\item But new methods coming up often to take advantage of better imaging technology ... 
+		\item originally done on scopes (stage micrometer) % Robinson and Evans, https://conservancy.umn.edu/server/api/core/bitstreams/30f91f03-c27b-44de-8659-5156220ce79b/content
+		\item eventually there was a shift to analyzing digital images of samples instead of stage micrometers % Guay et al. 1992 
+		\item digital copies inherently allow for contributing to data pools  % Rademacher​ et al.
+WIAD; also International Tree-Ring Data Bank (IRTB) https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring
+		\item Development of image acquisition in dendrochronology 
 			\begin{enumerate}
-			\item vessel counting requires detail at the sub-vessel level % detail is only as good as sample prep
+			\item two models: flatbed scanner or overlapping images and stitch
+			\item beginning with flatbed scanners % WinDendro
+				\subitem modern flatbed scanners can achieve high resolution (4200 dpi) but size of samples is limited % see CaptuRING paper for flatbed specs
+			\item overlapping images and stitching 
+				\subitem ATRICS, Gigapixel, CaptuRING 
+				\subitem takes advantage of advanced camera technology
 			\end{enumerate}
@@
 		\end{enumerate}
-	\item Development of scanning alternatives 
-		\begin{enumerate}
-		\item beginning with ATRICS
-		\item expensive Gigapixel
-		\item DIY CaptuRING
-		\end{enumerate}
+
 	\item Main reasons to build this 
@@
 		\item further develop the affordable / open source alternatives for scientific equipment
+		\item extend the capabilities of CaptuRING by allowing to stitch cookies and multiple samples in a queue
 		\item reduce the need for proprietary software such as PTGUI for image stitching
-		\item allow a method to digitize cookies to have higher quality data than cores alone
-		\item avoid size limitations to digitize tree cookies and cores 
+		\item increase the maximum size of a sample 
 		\item Digitizing wood samples for better data pooling / archiving
@@
 		\item cost effective
+		\item vessel counting requires detail at the sub-vessel level % https://roxas.wsl.ch/en/detail is only as good as sample prep
 		\item tested on CooRecorder + CDendro

### 2024-10-30 docs/writing/paper.tex
Starting to write introduction and updating outline as I write and reword

@@
 \section{Introduction}
-This is the introduction section where you introduce the topic and cite some papers.
+The study of tree rings has proven useful across multiple fields, proving to be a reliable subject for reconstructing past climates of regional and local environments, as well as a mechanism to understand tree growth response \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
+To obtain these data from the tree rings, it is necessary to measure tree rings from a tree cookie or core.
+The first high precision tool for this purpose was a stage micrometer, involving a trained technician to incrementally shift a tree core under the objective of a microscope - informing a computer when a new ring is encountered \citep{robinson_microcomputer_nodate}.
+While this method has very high precision, the data is only as accurate as the experience and knowledge of the technician at the time of recording \citep{levanic_atrics_2007}.
+The desire to remove repetition of errors in sampling and sampling bias across to individual technicians led researchers to an alternative - image analysis. 
+\par
+The first step in measuring tree ring width from images requires the digitization of the sample.
+There are currently two competing models to digitize a sample.
+The first model is to use a flatbed scanner or similar tool, providing the benefit to
+
+The first approach involved flatbed scanning technology which digitized the surface of a prepared sample \citep{guay_new_1992}. 
+Flatbed scanners eventually encountered a resolution insufficiency when analyzing very narrow tree rings on the scale of a few cells.
+A new method introduced with the ATRICS system increased the resolution of samples by stitching together multiple images from a digital microscope camera \citep{levanic_atrics_2007}.
+ATRICS set an example of the process which has been implemented in two other image acquisition tools, Gigapixel and CaptuRING \citep{griffin_gigapixel_2021, garcia-hidalgo_capturing_2022}.
+Both of these tools use professional cameras to take individual images of cores, although CaptuRING requires the technician to transfer the images to a second device and use the proprietary image stitching software PTGUI.
+A budget friendly image acquisition tool with integrated image stitching for both cores and cookies has yet to be released.
+
+% Potentially should have a section on tree ring width measurement software
+% The images were then fed to software such as WinDENDRO to have technicians click tree rings on a graphical user interface (GUI) \citep{guay_new_1992}.
+% More modern implementations of measurement software now exists which includes width measurement, cross dating, and autodetection of rings as well \citep{maxwell_measuring_2021}.
 

### 2024-10-30 docs/writing/paper.tex
I think I figured out the intro of the outline now. Writing in the paper is almost caught up

@@
 There are currently two competing models to digitize a sample.
-The first model is to use a flatbed scanner or similar tool, providing the benefit to
+The first model is to use a flatbed scanner or similar tool, to digitize the entire sample at once \citep{guay_new_1992}.
+Eventually flatbed scanners have reached a maximum resolution, introducing the second model spawning from ATRICS \citep{levanic_atrics_2007}.
+A high resolution camera takes multiple images across the surface of the sample, and uses image stitching techniques to combine them into one ultra high resolution image \citep{muhlich_stitching_2022}.
+This methodology is often seen in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}.
 
-The first approach involved flatbed scanning technology which digitized the surface of a prepared sample \citep{guay_new_1992}. 
-Flatbed scanners eventually encountered a resolution insufficiency when analyzing very narrow tree rings on the scale of a few cells.
-A new method introduced with the ATRICS system increased the resolution of samples by stitching together multiple images from a digital microscope camera \citep{levanic_atrics_2007}.
-ATRICS set an example of the process which has been implemented in two other image acquisition tools, Gigapixel and CaptuRING \citep{griffin_gigapixel_2021, garcia-hidalgo_capturing_2022}.
 Both of these tools use professional cameras to take individual images of cores, although CaptuRING requires the technician to transfer the images to a second device and use the proprietary image stitching software PTGUI.
 A budget friendly image acquisition tool with integrated image stitching for both cores and cookies has yet to be released.
-
-% Potentially should have a section on tree ring width measurement software
-% The images were then fed to software such as WinDENDRO to have technicians click tree rings on a graphical user interface (GUI) \citep{guay_new_1992}.
-% More modern implementations of measurement software now exists which includes width measurement, cross dating, and autodetection of rings as well \citep{maxwell_measuring_2021}.
 

### 2024-10-31 docs/writing/paper.tex
Adding more to the intro

@@
 \par
-The first step in measuring tree ring width from images requires the digitization of the sample.
-There are currently two competing models to digitize a sample.
-The first model is to use a flatbed scanner or similar tool, to digitize the entire sample at once \citep{guay_new_1992}.
-Eventually flatbed scanners have reached a maximum resolution, introducing the second model spawning from ATRICS \citep{levanic_atrics_2007}.
-A high resolution camera takes multiple images across the surface of the sample, and uses image stitching techniques to combine them into one ultra high resolution image \citep{muhlich_stitching_2022}.
+The first step in measuring tree ring width from images requires the digitization of the sample from one of two major methods.  
+The original technique was as a flatbed scanner which can digitize the entire sample at once \citep{guay_new_1992}. 
+With a top of the line scanner, like the Epson Perfection V850 Pro, it's possible to scan at maximum resolutions of 4800 dpi. 
+The resolution ceiling and a limited scan area of 8.5" x 11.7" make scanners unfit for ultra high resolution analysis or larger samples.
+
+During the time it took for scanners to reach high resolution, the second digitization model was introduced with ATRICS \citep{levanic_atrics_2007}. 
+Rather than scanning a whole sample at once, a high resolution camera takes multiple images across the surface of the sample and uses image stitching software techniques to combine them into one ultra high resolution image \citep{muhlich_stitching_2022}.
 This methodology is often seen in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}. 
+This method requires either the camera objective to move relative to the sample, or the sample move underneath a stationary camera. 
+For ATRICS and a more modern do-it-yourself alternative, CaptuRING, the sample is moved relative to the camera \citep{garcia-hidalgo_capturing_2022}. 
+Gigapixel takes a different approach by moving the camera relative to the sample, allowing for multiple samples to be recorded in sequence.
 

### 2024-10-31 docs/writing/paper.tex
Introduction draft done

@@
 The desire to remove repetition of errors in sampling and sampling bias across to individual technicians led researchers to an alternative - image analysis.
-\par
+
 The first step in measuring tree ring width from images requires the digitization of the sample from one of two major methods.  
 The original technique was as a flatbed scanner which can digitize the entire sample at once \citep{guay_new_1992}. 
-With a top of the line scanner, like the Epson Perfection V850 Pro, it's possible to scan at maximum resolutions of 4800 dpi. 
-The resolution ceiling and a limited scan area of 8.5" x 11.7" make scanners unfit for ultra high resolution analysis or larger samples.
+With a top of the line scanner, like the Epson Perfection V850 Pro, it's possible to scan at maximum resolutions of 4800 dpi and scan an area of up to 8.5" x 11.7".
+Analysis which relies on higher resolution larger samples require a different digitization approach.
 
-During the time it took for scanners to reach high resolution, the second digitization model was introduced with ATRICS \citep{levanic_atrics_2007}. 
+The second digitization model was introduced with ATRICS \citep{levanic_atrics_2007}. 
 Rather than scanning a whole sample at once, a high resolution camera takes multiple images across the surface of the sample and uses image stitching software techniques to combine them into one ultra high resolution image \citep{muhlich_stitching_2022}.
@@
 Gigapixel takes a different approach by moving the camera relative to the sample, allowing for multiple samples to be recorded in sequence. 
+While these machines can all digitize cores, none have been shown to digitize cookies.
 
-Both of these tools use professional cameras to take individual images of cores, although CaptuRING requires the technician to transfer the images to a second device and use the proprietary image stitching software PTGUI.
-A budget friendly image acquisition tool with integrated image stitching for both cores and cookies has yet to be released.
+Tina was made to combine the defining features of the previously mentioned machines into one while making the code open-source and chassis open-hardware device. 
+We designed Tina to digitize both cookies and cores, extend the maximum sample length, perform image stitching without user intervention, while minimizing cost. 
+The only specialized piece of equipment needed to build Tina is a 3D printer, but the parts can be readily ordered through 3D print shops if preferred.
+Excluding 3D printed parts, the total cost of the machine is approximately \$2,200 USD compared to the \$70,000 USD of the Gigapixel \citep{griffin_gigapixel_2021}.
+The total cost of the machine is almost comparable in price or less to many professional camera and macro lens combinations.
+Additional savings can also be made when factoring in the cost of a professional stitching software license such as PTGui.
 
@@
 To do this, the system traverses the surface area of the sample after being given its dimensions.
-After the capturing process, the images are stitched and are stored on the device along with metadata of relevant
+After the capturing process, the images are stitched and are stored on the device along with metadata of relevant machine
 settings and sample parameters.
 
-\subsection{Hardware Design}
+%% Previous Methods is commented out so I can reorder things 
+% \subsection{Hardware Design}
 
-\subsubsection{Mechanical Structure}
-The cartesian robot was purchased from an open source parts supplier, OpenBuilds. They sell various sets of gantry robots, motor controllers, and power supplies
-which can be conveniently purchased from one supplier. All metal components are cut and sized, and have very detailed assembly videos, an active community forum,
-and customer support to fall back on.
+% \subsubsection{Frame}
+% The cartesian gantry machine was purchased from an open source parts supplier, OpenBuilds. They sell various sets of gantry robots, motor controllers, and power supplies
+% which can be conveniently purchased from one supplier. All metal components are cut and sized, and have very detailed assembly videos, an active community forum,
+% and customer support to fall back on.
 The OpenBuilds ACRO system circumvents the need for most power tools, machining, electrical connectors, and technical knowhow that would otherwise be necessary to build a robust robot.
-All other parts were designed using SolidWorks and printed on an FDM 3D printer using ABS material. Mechanical assembly can be completed entirely with common hand tools and metric fasteners.
+% All other parts were designed using SolidWorks and printed on an FDM 3D printer using ABS material. Mechanical assembly can be completed entirely with common hand tools and metric fasteners.
 
-\subsubsection{Computer}
-The computer which runs the graphical user interface (GUI), gantry control, camera streaming, and image stitching is the NVIDIA Jetson Orin Nano Developer Kit (Orin Nano). Using a computer
-which is compatible with CSI cameras such as the Raspberry Pi HQ Camera was essential to maintain cost effectiveness. The Orin Nano, like others in the Jetson line of products, has hardware
-accelerators for common GPU tasks such as image streaming, which is vital to retain CPU power for GUI tasks and robot control.
+% \subsubsection{Computer}
+% The computer which runs the graphical user interface (GUI), gantry control, camera streaming, and image stitching is the NVIDIA Jetson Orin Nano Developer Kit (Orin Nano). Using a computer
+% which is compatible with CSI cameras such as the Raspberry Pi HQ Camera was essential to maintain cost effectiveness. The Orin Nano, like others in the Jetson line of products, has hardware
+% accelerators for common GPU tasks such as image streaming, which is vital to retain CPU power for GUI tasks and robot control.
 
-\subsubsection{Camera}
-Unlike the Gigapixel or CaptuRING, the camera is a machine vision camera rather than a handheld professional camera. Choosing the Raspberry Pi HQ Camera provided many lens options, significant
-forum support, and includes an Obsolescence Statement that the camera will be in production until January 2030.
+% \subsubsection{Camera}
+% Unlike the Gigapixel or CaptuRING, the camera is a machine vision camera rather than a handheld professional camera. Choosing the Raspberry Pi HQ Camera provided many lens options, significant
+% forum support, and includes an Obsolescence Statement that the camera will be in production until January 2030.
 Combined with a 180mm C-mount microscope lens, the images can reach above 20,000 dots per inch (DPI).
 
-\subsection{Software Design}
-All software was written in Python VERSION and a variety of packages for the GUI, image stitching, and calculations (Should I include all the packages in writing?). Care was taken to segment the
-software design into abstracted object oriented code.
+% \subsection{Software Design}
+% All software was written in Python VERSION and a variety of packages for the GUI, image stitching, and calculations (Should I include all the packages in writing?). Care was taken to segment the
+% software design into abstracted object oriented code.
 
-\subsubsection{Graphical User Interface}
-The GUI to control the machine launches from a Python script. Included are a viewer to see the live stream from the camera, buttons to jog the machine throughout the XYZ coordinate system,
-setting the size of samples, and running the process to capture and stitch images. The system was designed to allow for multiple samples to be prepared in a queue for bulk digitization.
+% \subsubsection{Graphical User Interface}
+% The GUI to control the machine launches from a Python script. Included are a viewer to see the live stream from the camera, buttons to jog the machine throughout the XYZ coordinate system,
+% setting the size of samples, and running the process to capture and stitch images. The system was designed to allow for multiple samples to be prepared in a queue for bulk digitization.
 
-\subsubsection{Sample Digitization}
-The digitization of a sample can be reduced to a few processes - capturing a grid of overlapping images, image focusing, and image stitching.
+% \subsubsection{Sample Digitization}
+% The digitization of a sample can be reduced to a few processes - capturing a grid of overlapping images, image focusing, and image stitching.
 
-Inside the GUI, the program is populated with the height and width of the sample as well as the height and width of the image. The user then navigates to the center of the sample and can choose 
-to begin imaging the sample or add the sample to a queue for bulk digitization. This provides the machine enough data to calculate the necessary rows and columns for traversing the entire surface
-area of the sample while maintaining a level of overlap between adjacent images. Image overlap is what allows for images to be stitched together as there is nonzero error when the robot translates 
-from one coordinate to another.
+% Inside the GUI, the program is populated with the height and width of the sample as well as the height and width of the image. The user then navigates to the center of the sample and can choose 
+% to begin imaging the sample or add the sample to a queue for bulk digitization. This provides the machine enough data to calculate the necessary rows and columns for traversing the entire surface
+% area of the sample while maintaining a level of overlap between adjacent images. Image overlap is what allows for images to be stitched together as there is nonzero error when the robot translates 
+% from one coordinate to another.
 
-To stitch adjacent images, feature based image stitching from a Python package Stitch2D was used. The underlying functions were based on the processing libraries
-OpenCV and NumPy.
+% To stitch adjacent images, feature based image stitching from a Python package Stitch2D was used. The underlying functions were based on the processing libraries
+% OpenCV and NumPy.
 Multiple other image stitching packages were tested but none were able to successfully piece together our samples.
@@
 
-One major difficulty with using the Raspberry Pi HQ Camera is the lack of a lens with auto-focus. An in focus image in this regime is dependent on the focal length of the lens set by the user and the
-distance from the sample.
+% One major difficulty with using the Raspberry Pi HQ Camera is the lack of a lens with auto-focus.
+An in focus image in this regime is dependent on the focal length of the lens set by the user and the
+% distance from the sample.
 The microscope lens has a very small depth of field and goes from in-focus to out-of-focus in less than a 1mm step in the distance to the sample.
 Two cooperative solutions were implemented
-to achieve a focused image for the entire surface area of the sample - taking images from multiple distances from the sample and automatic control.
+% to achieve a focused image for the entire surface area of the sample - taking images from multiple distances from the sample and automatic control.
 
-\subsubsection{Image Stack Focusing}
-When the sample is placed on the table to be digitized, it is assumed that there will, at best, very slight variations in height across the surface area of the sample. A levelling table 
-assists in reducing this error although it will never be nonzero. At each location (X, Y) on the grid of overlapping images, 11 images are captured in equally distanced Z steps across a 1mm range. 
-Each of these images have a normalized variance score calculated - the maximum value represents the most in focus image \citep{sampat_extensive_2014}.
+% \subsubsection{Image Stack Focusing}
+% When the sample is placed on the table to be digitized, it is assumed that there will, at best, very slight variations in height across the surface area of the sample. A levelling table 
+% assists in reducing this error although it will never be nonzero. At each location (X, Y) on the grid of overlapping images, 11 images are captured in equally distanced Z steps across a 1mm range. 
+% Each of these images have a normalized variance score calculated - the maximum value represents the most in focus image \citep{sampat_extensive_2014}.
 In practice, stopping and starting the motor causes
-enough vibration to require upwards to two seconds of pause time before the motion blur stops to take each photo. Instead of stopping and starting each image, an additional distance is added to the 
-Z-motion to allow the stepper to accelerate and reach constant velocity.
+% enough vibration to require upwards to two seconds of pause time before the motion blur stops to take each photo. Instead of stopping and starting each image, an additional distance is added to the 
+% Z-motion to allow the stepper to accelerate and reach constant velocity.
 At this speed, images are then taken with constant time steps instead of constant distance steps across the 1mm range.
 This eliminates
-the motion blur due to acceleration and dramatically reduces imaging time.
+% the motion blur due to acceleration and dramatically reduces imaging time.
 

### 2024-10-31 docs/writing/outline.tex
I think I figured out the intro of the outline now. Writing in the paper is almost caught up
Starting to make methods more clear to read. Suddenly Latex workshop doesn't want to recognize references.

@@
 				\subitem modern flatbed scanners can achieve high resolution (4200 dpi) but size of samples is limited % see CaptuRING paper for flatbed specs
+				\subitem can scan both cookies and cores	
 			\item overlapping images and stitching 
 				\subitem ATRICS, Gigapixel, CaptuRING 
+				\subitem so far can only capture cores
+				\subitem budget friendly implementations require user intervention between image acquisition and stitching
 				\subitem takes advantage of advanced camera technology
@@
 		\begin{enumerate}
-		\item Guiding design considerations
-			\begin{enumerate}
-			\item Digitization of a tree cookie and core
-			\item Minimizing barriers to assemble (Cost effective, open source, open hardware, minimal use of specialized equipment/tools)
-			\item Reducing user interaction between sample prep and the final image
-			\item Achieving near microscopic detail with a field of view as large as the sample
-			\end{enumerate}
+		% 
+		% \item Guiding design considerations
+		% 	\begin{enumerate}
+		% 	\item Digitization of a tree cookie and core
+		% 	\item Minimizing barriers to assemble (Cost effective, open source, open hardware, minimal use of specialized equipment/tools)
+		% 	\item Reducing user interaction between sample prep and the final image
+		% 	\item Achieving near microscopic detail with a field of view as large as the sample
+		% 	\end{enumerate}
 		\item Considerations for Digitizing Cookies

### 2024-10-31 docs/writing/paper.tex
Starting to make methods more clear to read. Suddenly Latex workshop doesn't want to recognize references.

@@
 
+The frame, camera, computer, software, and sample leveling tables, work cohesively to address the nuances of digitizing both cookies and cores. 
+Each of these subsystems will be reviewed along with the design considerations involved to make a complete machine. 
+Considering Tina is an imaging machine, we will start off by looking at the camera and move outward.
+
+\subsection{System Overview}
+Camera, computer, software, XYZ frame, adapters.
+
+\subsection{Camera}
+In the Gigapixel and CaptuRING, a professional handheld camera was used to capture images.
+While this benefits from a wide variety of lens and camera choices - and principally autofocus - they come at a cost.
+Beyond its literal price tag, a professional camera also carries a significant weight, demanding a more robust and therefore more costly machine frame. 
+These cameras are also optimized for handheld use, not for integration into computer driven machines. This results in clunky access to images from software. 
+Designing around a machine vision camera such as the 12MP Raspberry Pi HQ Camera drastically reduced the camera cost while improving the access to the data stream coming from the camera sensor.
+Pairing the Raspberry Pi HQ Camera with a high magnification C-Mount lens allows for images with very high resolution in dots-per-inch (DPI) to be captured.
+(THIS IS AN ESTIMATE, GET THE REAL VALUES FROM TINA) If the field of view of the image
+could be optically zoomed to 5mm x 3mm, the resolution can reach a theoretical 32,000 DPI.
+
+\subsubsection{Focusing}
+Normalized variance process goes here.
+
+Achieving a focused image across the entire surface area of the sample is the basis for an accurate stitch and effective analysis.
+By saving costs on a fixed focus camera and lens, it was necessary to implement a method to focus the images.
+Focus is solely a function of distance to the subject in the regime.
+With this high magnification lens, having an image go from in-focus to out-of-focus was sensitive to sub millimeter variation in distance to the lens.
+This requires cores and especially cookies to be as close to orthogonally aligned to the lens as possible.
+A 3D printed levelling table and bullseye level were used to correct plane misalignment of up to 10 degrees in cookies (CALCULATE THIS?).
+Slight sample misalignments are corrected within a closed loop automatic control PID algorithm.
+
 %% Previous Methods is commented out so I can reorder things 
@@
 
-$$\frac{1}{MN\mu} \sum\limits_{x=0}^{M-1} \sum\limits_{y=0}^{N-1}(f(x,y) - \mu)^2$$
+% $$\frac{1}{MN\mu} \sum\limits_{x=0}^{M-1} \sum\limits_{y=0}^{N-1}(f(x,y) - \mu)^2$$
 
-\subsubsection{Automatic Control}
-While the image stack provides a basis to obtain an in focus image, it is statically limited to the 1mm range in the Z axis. Instead of spending significant time guaranteeing that the sample
-is within the the Z axis range across the entire surface area, a PID control algorithm takes the wheel \citep{odwyer_summary_2000}.
+% \subsubsection{Automatic Control}
+% While the image stack provides a basis to obtain an in focus image, it is statically limited to the 1mm range in the Z axis. Instead of spending significant time guaranteeing that the sample
+% is within the the Z axis range across the entire surface area, a PID control algorithm takes the wheel \citep{odwyer_summary_2000}.
 The information previously calculated in the image stack is sufficient
-to inform the PID controller of how to adjust the initial the Z axis for the next image stack. The controller is constantly trying to make the most in focus image be at the center of the image stack,
-index value 5. This allows for the most flexibility moving to the next location of the image grid.
+% to inform the PID controller of how to adjust the initial the Z axis for the next image stack.
+The controller is constantly trying to make the most in focus image be at the center of the image stack,
+% index value 5. This allows for the most flexibility moving to the next location of the image grid.
 
-\begin{figure}
-    \centering
-    \begin{subfigure}{.5\textwidth}
-      \centering
-      \includegraphics[height=0.75\linewidth]{../diagrams/sample_setup_ideal.png}
-      \caption{Ideal sample leveling}
-      \label{fig:sub1}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \centering
-      \includegraphics[height=0.75\linewidth]{../diagrams/sample_setup_realistic.png}
-      \caption{Realistic sample leveling}
-      \label{fig:sub2}
-    \end{subfigure}
-    \caption{Side view of the camera and sample on top of a leveling table.
+% \begin{figure}
+%     \centering
+%     \begin{subfigure}{.5\textwidth}
+%       \centering
+%       \includegraphics[height=0.75\linewidth]{../diagrams/sample_setup_ideal.png}
+%       \caption{Ideal sample leveling}
+%       \label{fig:sub1}
+%     \end{subfigure}%
+%     \begin{subfigure}{.5\textwidth}
+%       \centering
+%       \includegraphics[height=0.75\linewidth]{../diagrams/sample_setup_realistic.png}
+%       \caption{Realistic sample leveling}
+%       \label{fig:sub2}
+%     \end{subfigure}
+%     \caption{Side view of the camera and sample on top of a leveling table.
 The ideal sample leveling shows a uniform distance d at all (x,y) coordinates on the sample.
 This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique (x,y) coordinates.}
-    \label{fig:test}
-\end{figure}
+%     \label{fig:test}
+% \end{figure}
 

### 2024-11-06 docs/writing/paper.tex
Adding some math equations to methods regarding image focusing

@@
 \usepackage{geometry}           % For setting page margins
-\usepackage{natbib}             % For citations
+\usepackage[numbers]{natbib}             % For citations
 \usepackage{hyperref}           % For clickable links in references
 \usepackage{subcaption}         % For side by side subfigures
-
 % Page layout
@@
 Tina was made to combine the defining features of the previously mentioned machines into one while making the code open-source and chassis open-hardware device. 
-We designed Tina to digitize both cookies and cores, extend the maximum sample length, perform image stitching without user intervention, while minimizing cost. 
+We designed Tina to digitize both cookies and cores, extend the maximum sample length, perform image stitching without operator intervention, while minimizing cost. 
+Successful functionality of this system is dependent on the major steps in tree cookie and core digitization to be automated. 
 The only specialized piece of equipment needed to build Tina is a 3D printer, but the parts can be readily ordered through 3D print shops if preferred.
@@
 
-\section{Methods}
-The functionality of this system allows for the major steps in tree cookie and core digitization to be automated.
-Capturing images and stitching them together is handled without the need for user intervention.
-To do this, the system traverses the surface area of the sample after being given its dimensions.
+\section{System Overview} % this is so dry fix later
+
+Tina can be thought of as a combination of multiple subsystems: the camera, computer, and gantry machine. The first system is the camera and computer. 
+A 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens is connected to an NVIDIA Jetson Orin Nano (Jetson) through a CSI cable. 
+The Jetson is a powerful SoC edge computer which drives a computer monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for automatic control, and performs calculations to stitch individual images into one mosaic.
+The gantry machine is built using the extruded aluminum and premade adapters from the OpenBuilds ACRO kit. 
+Movement of the machine is originates from NEMA 17 stepper motors and is transferred to
+linear motion in the X and Y direction using timing belts and timing pulleys. 
+Motion in the Z-direction is controlled by a NEMA 17 lead screw based linear actuator also supplied by OpenBuilds.
+Custom 3D models were designed in SolidWorks to create adapters to mount the camera to the Z axis as well as mounting the Z axis to the XY plate of the ACRO. 
+The OpenBuilds X32 Motor Controller running GRBL firmware converts G-Code commands received via serial from the Jetson.
+GRBL then converts them to electrical signals to spin the lead screw and timing belts.
+
+Interacting with the Jetson is similar to working with any Ubuntu based Linux distribution.
+
+\section{Sample Digitization} % Methods
+The subsystems can be best understood by following the path from sample setup through the computing process to create the stitched image.
+
+To achieve this with a fixed focus camera, the samples must be nearly orthogonal to the camera lens.
+Misalignments between the sample and lens can be corrected by using the 3D printed sample levelling table we designed.
+Once the sample is level, the operator interacts with the machine through the GUI to navigate the camera to the center of the sample, and initializes focus to be sharp.
+The height and width dimensions are then entered in the GUI
+to be saved along with the center XYZ coordinate of the sample. This procedure can be repeated to create a queue of samples to digitize.
+
+Sample height, width, identifiers, and centering an in focus sample is sufficient information to digitize.
+
+\subsection{Image Capturing}
+Once prompted the system begins to exhaustively traverse the surface area of the sample. 
+The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors - the basis of image stitching. 
+Digitizing cores can often be done without needing to traverse in both the X and Y axis as the field of view in an image can capture the core's entire axial diameter in its field of view when moving along the core's length.
+% confusing wording fix later
+Spanning the surface area of a cookie is more complicated as it needs move in both the X and Y axis (see Figure).
+
+By centering the design around a fixed focus camera and lens, it is necessary to implement some sense of automatic control to focus the images. 
+In this system, the only way to control focus is by changing the lens's distance to the sample. 
+And with the microscope lens, the depth of field of the image is so sensitive that a perturbation of less half of a millimeter can move an entire image out of focus.
+Solving this in a time efficient manner involves two stages. 
+The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
+This allows the initial $Z_0$ value from $(X_0, Y_0, Z_0)$ to be a fantastic initial guess as to what value would have an in focus image across the entire surface. 
+Sample surface misalignment and a non flat sample (bumpy) surfaces require more than one image to be taken at each image $(X, Y)$. 
+From the first $(X_0, Y_0)$ coordinate captured, 11 images are taken at different $Z$ values in $\boldsymbol{Z_{\text{focus}}}$. 
+So long as $Z_{sample}$ is within this set, an in focus image exists.
+
+\[
+Z_{sample} \in
+\boldsymbol{Z_{\text{focus}}} = 
+Z_0 + 
+\begin{bmatrix}
+-0.5 \\
+-0.4 \\
+\vdots \\
+0.4 \\
+0.5 \\
+\end{bmatrix}
+\exists 
+Z_{focused}
+\] % unsure if this makes sense
+
+Beginning at 0.5 mm above $Z_0$ and the last image finishing at 0.5 mm below $Z_0$.
+To reduce motion blur in the images, this $\boldsymbol{Z_{\text{focus}}}$ is traversed at constant velocity and images are captured without stopping. 
+The 11 images then have their normalized variance calculated in a separate thread to measure the image sharpness \citep{sampat_extensive_2014}.
+The image with the maximum score is saved while the rest are deleted from storage.
+
+\[
+    i_{\max} = \arg\max_{i} \boldsymbol{Z_{\text{focus}}}
+\]
+
+This focusing procedure works well alone when the sample alignment has a difference in height no greater than 1 mm. 
+Greater alignment error can be managed by having the control over adjusting $Z_{0,k}$ for $(X_k,Y_k)$.
+The likelihood of an adjacent $(X_{k+1}, Y_{k+1})$ containing an in focus image is highest when the current $i_{{\max,k}}$ is at the middle index of $\boldsymbol{Z_{focus}}$. 
+A PID control algorithm with a process variable of $i_{max}$ and control variable of $Z_{0,k}$ allows a negative feedback loop to improve focusing across the entire sample \citep{odwyer_summary_2000}.
+
 After the capturing process, the images are stitched and are stored on the device along with metadata of relevant machine
-settings and sample parameters.
+settings and sample information.
 
@@
 
-\subsection{System Overview}
-Camera, computer, software, XYZ frame, adapters.
-
 \subsection{Camera}
@@
 Beyond its literal price tag, a professional camera also carries a significant weight, demanding a more robust and therefore more costly machine frame. 
-These cameras are also optimized for handheld use, not for integration into computer driven machines. This results in clunky access to images from software. 
+These cameras are also optimized for handheld use, not for integration into computer driven machines. 
+This results in clunky access to images from software. 
 Designing around a machine vision camera such as the 12MP Raspberry Pi HQ Camera drastically reduced the camera cost while improving the access to the data stream coming from the camera sensor.
@@
 (THIS IS AN ESTIMATE, GET THE REAL VALUES FROM TINA) If the field of view of the image
-could be optically zoomed to 5mm x 3mm, the resolution can reach a theoretical 32,000 DPI.
 

### 2024-11-08 code/camera.py
Adding config file to change hard coded values

@@ -2,6 +2,8 @@ from threading import Thread
 import time
 import logging as log
 import gi
+import utils
+
 gi.require_version('Gst', '1.0')
 from gi.repository import Gst, GObject, GLib
 
@@ -11,10 +13,12 @@ class Camera:
 
     def __init__(self, quiet = True):
         # Adding hardcoded image size, will need to update to update DPI
-        CROP_W = 500
-        CROP_H = 100
-        W_PIXELS = 3840 - 2 * CROP_W
-        H_PIXELS = 2160 - 2 * CROP_H
+        self.config = utils.load_config()
+        
+        CROP_W = self.config["camera"]["CROP_W"]
+        CROP_H = self.config["camera"]["CROP_H"]
+        W_PIXELS = self.config["camera"]["W_PIXELS_NO_CROP"] - 2 * CROP_W
+        H_PIXELS = self.config["camera"]["H_PIXELS_NO_CROP"] - 2 * CROP_H
 
         Gst.init(None)
         # Create the pipeline with both display and save frame functionality

### 2024-11-08 code/controller.py

@@ -13,6 +13,7 @@ from multiprocessing import Process
 from pathlib import Path
 import json
 from typing import Callable
+import utils
 
 
 class Controller:
@@ -24,9 +25,18 @@ class Controller:
             image_height_mm (float): Initial image height
         """
         #Settings for capturing images from multiple distances
-        self.n_images = 9 #make sure you're not going faster than the frame rate of the GStreamer feed... 
-        self.height_range = 1
-        
+        # self.n_images = 11 #make sure you're not going faster than the frame rate of the GStreamer feed... 
+        self.config = utils.load_config()
+        self.n_images = self.config["controller"]["N_IMAGES_MULTIPLE_DISTANCES"] #make sure you're not going faster than the frame rate of the GStreamer feed... 
+        # self.height_range = 1
+        self.height_range = self.config["controller"]["HEIGHT_RANGE_MM"]
+        self.acceleration_buffer = self.config["controller"]["ACCELERATION_BUFFER_MM"]
+        self.stitch_sizes = self.config["controller"]["STITCH_SIZES"]
+        self.slow_feed_rate_xy = self.config["controller"]["SLOW_FEED_RATE_XY"]
+        self.slow_feed_rate_z = self.config["controller"]["SLOW_FEED_RATE_Z"]
+        self.fast_feed_rate_xy = self.config["controller"]["FAST_FEED_RATE_XY"]
+        self.fast_feed_rate_z = self.config["controller"]["FAST_FEED_RATE_Z"]
+    
         #Objects
         self.cookies = []
         self._gantry = gantry.Gantry()
@@ -152,7 +162,8 @@ class Controller:
         Args:
             frame_dir (str): Directory of where the frames are. 
         """
-        sizes = [0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0]
+        # sizes = [0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0]
+        sizes = self.config["controller"]["STITCH_PERCENTAGES"]
         for size in sizes:
             st = stitcher.Stitcher(frame_dir) 
             #st.stitch(resize=size)
@@ -207,16 +218,16 @@ class Controller:
                     current_col = None
                     if row % 2 == 1:
                         current_col = cols - col - 1
-                        coordinates[row][current_col] = self._gantry.get_xyz()
[…]

### 2024-11-08 code/cookie.py

@@ -44,11 +44,11 @@ class Cookie:
         self.image_width_mm = image_width_mm
         self.image_height_mm = image_height_mm
         self.rows, self.cols, self.x_step_size, self.y_step_size = self.calculate_grid()
-        self.background = np.zeros((self.rows, self.cols))
-        self.background_std = np.zeros((self.rows, self.cols))
-        self.coordinates = np.zeros((self.rows, self.cols, 3))
-        self.focus_index = np.zeros((self.rows, self.cols))
-        self.nvar = np.zeros((self.rows, self.cols, 9))
+        self.background = []
+        self.background_std = []
+        self.coordinates = []
+        self.focus_index = []
+        self.nvar = []
 
     def calculate_grid(self):
         """Calculates the amount of rows and columns necessary to traverse the cookie.

### 2024-11-08 code/focus.py

@@ -5,18 +5,26 @@ import numpy as np
 import os
 from pathlib import Path
 from PID import AsynchronousPID 
+import utils
 
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
 class Focus:
 
     def __init__(self, delete_flag, setpoint):
+        self.config = utils.load_config()
+
         self.DELETE_FLAG = delete_flag
         self.scale_factor = 0.05
-        self.PID = AsynchronousPID(Kp=1.0, Ki=0.1, Kd=0.05, setpoint=setpoint) 
+
+        self.kp = self.config["focus"]["Kp"]        
+        self.ki = self.config["focus"]["Ki"]        
+        self.kd = self.config["focus"]["Kd"]     
+        self.background_std_threshold = self.config["focus"]["BACKGROUND_STD_THRESHOLD"]  # low zoom = 0.015, high zoom = 0.125
+        self.PID = AsynchronousPID(Kp=self.kp, Ki=self.ki, Kd=self.kd, setpoint=setpoint) 
         self.TESTINGLOG = []
 
-    def find_focus(self, focus_queue, pid_queue, pid_lock, directory, nvars: np.array, index:np.array, background:np.array, background_std:np.array):
+    def find_focus(self, focus_queue, pid_queue, pid_lock, directory, nvars: list, index:list, background:list, background_std:list):
         while True:
             image_files = focus_queue.get()
             pid_lock.acquire()
@@ -37,10 +45,10 @@ class Focus:
                 Path("{}/focused_images".format(self.directory)).mkdir(exist_ok=True)
                 shutil.copy(focused_image_name, "{}/focused_images/{}".format(directory,filename))
             
-            background_std[row][col] = std
-            background[row][col] = self.is_background(std)
-            index[row][col] = stack_number
-            nvars[row][col][:] = nv
+            background_std.append(std)
+            background.append(self.is_background(std))
+            index.append(stack_number)
+            nvars.append(nv)
 
             if not self.is_background(std):
                 control_variable = self.PID.update(int(stack_number))
@@ -76,7 +84,7 @@ class Focus:
                 if var > best_var:
                     best_image_filepath = image_name
                     best_var = var
-
+            
         std = np.std(vars)
 
         return best_image_filepath, std, vars
@@ -87,7 +95,7 @@ class Focus:
     
     def is_background(self, std):
         # High zoom, 0.125 seemed to work. Low zoom, 0.015 seems to work
-        if std > 0.015:
[…]

### 2024-11-08 code/gantry.py

@@ -3,12 +3,14 @@ import serial
 import time
 from threading import Thread, Lock
 import re
+import utils
 
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
 class Gantry:
     def __init__(self, serial_port = "/dev/ttyUSB0", quiet=True):
-        
+        self.config = utils.load_config()
+
 
         self.quiet = quiet
         self._serial_port = serial_port # windows should be a "COM[X]" port which will vary per device
@@ -16,8 +18,8 @@ class Gantry:
         # machine settings
         # fast z is 100, slow z is 15
         # fast xy is 500, slow xy is 200
-        self.feed_rate_z = 300
-        self.feed_rate_xy = 300 
+        self.feed_rate_z = self.config["gantry"]["FEED_RATE_DEFAULT_Z"]
+        self.feed_rate_xy = self.config["gantry"]["FEED_RATE_DEFAULT_XY"] 
 
         # sample information
         self.cookie_samples = []

### 2024-11-08 code/gui.py

@@ -7,14 +7,17 @@ import logging as log
 import controller
 from datetime import datetime
 import time
+import utils
 
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
 class App(Gtk.Window):
     def __init__(self):
         super().__init__(title="Cookie Capture")
-        self.set_default_size(600, 400)
-        self.set_size_request(600, 400)
+        self.config = utils.load_config()
+
+        self.set_default_size(self.config["gui"]["DEFAULT_WINDOW_SIZE"][0], self.config["gui"]["DEFAULT_WINDOW_SIZE"][1])
+        self.set_size_request(self.config["gui"]["DEFAULT_WINDOW_SIZE"][0], self.config["gui"]["DEFAULT_WINDOW_SIZE"][1])
         self.connect("destroy", self.quit_program)
         
         self.controller = controller.Controller(3, 2)
@@ -120,7 +123,7 @@ class App(Gtk.Window):
         label_height_cookie = Gtk.Label(label="Enter Sample Height (mm):   ")
         box.pack_start(label_height_cookie, True, True, 0)
         self.entry_height_cookie = Gtk.Entry()
-        self.entry_height_cookie.set_text("45")
+        self.entry_height_cookie.set_text("{}".format(self.config["gui"]["DEFAULT_SAMPLE_HEIGHT_MM"]))
         box.pack_start(self.entry_height_cookie, True, True, 0)
         self.entry_height_cookie.connect('focus-out-event', self.print_cookie_height_entry)
 
@@ -128,7 +131,7 @@ class App(Gtk.Window):
         label_width_cookie = Gtk.Label(label="Enter Sample Width (mm):   ")
         box.pack_start(label_width_cookie, True, True, 0)
         self.entry_width_cookie = Gtk.Entry()
-        self.entry_width_cookie.set_text("45")
+        self.entry_width_cookie.set_text("{}".format(self.config["gui"]["DEFAULT_SAMPLE_WIDTH_MM"]))
         box.pack_start(self.entry_width_cookie, True, True, 0)
         self.entry_width_cookie.connect('focus-out-event', self.print_cookie_width_entry)
     
@@ -147,7 +150,7 @@ class App(Gtk.Window):
         self.zoom_combo.pack_start(renderer_text, True)
         self.zoom_combo.add_attribute(renderer_text, "text", 0)
 
[…]

### 2024-11-08 code/stitcher.py

@@ -13,6 +13,8 @@ import imutils
 import logging as log
 # from memory_profiler import profile
 import sys 
+import time
+import utils
 
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
@@ -25,15 +27,13 @@ class MaxFileSizeException(Exception):
 
 class Stitcher:
     def __init__(self, frame_dir):
+        config = utils.load_config()
         self._memmaps = []
         self._tiles = []
         self._mosaic = None
-        self._sizes = [1.0, 0.4, 0.3, 0.2, 0.1, 0.05, 0.01] # percentage of full resolution
-        self._full_resolution = round(3840 * 2160 / 1e6, 2)
         self._frame_dir = frame_dir
-        self._resolution = round(3840 * 2160 / 1e6, 2)        
         self._metadata = None
-        self._max_file_size = 10000 # 10 GBs
+        self._max_file_size = config["stitcher"]["MAX_FILE_SIZE_GB"] * 1000 # GBs
 
         self.load_metadata()
         
@@ -129,7 +129,7 @@ class Stitcher:
             self._metadata = j
             f.close()
 
-    def write_metadata(self, new_dir, shape):
+    def write_metadata(self, new_dir, shape, stitch_time):
         new_metadata = {}
         new_metadata["species"]=self._metadata["species"]
         new_metadata["id1"]=self._metadata["id1"]
@@ -143,7 +143,8 @@ class Stitcher:
         new_metadata["pixels_h"] = shape[0]
         new_metadata["pixels_w"] = shape[1]
         new_metadata["depth"] = shape[2]
-        
+        new_metadata["stitch_time"] = stitch_time
+         
 
         with open(os.path.join(new_dir, 'metadata.json'), 'w', encoding='utf-8') as f:
             json.dump(new_metadata, f, ensure_ascii=False, indent=4)
@@ -173,6 +174,8 @@ class Stitcher:
         return sorted_paths
     # @profile
     def stitch(self, resize=None):
+        start_time = time.time()
+
         if resize is not None and resize < 1.0 and resize > 0:
             path = os.path.join(self._frame_dir, "{}per".format(int(resize * 100)))
             mosaic_dat_path = os.path.join(path, "mosaic_{}per.dat".format(int(resize *100)))
@@ -198,7 +201,9 @@ class Stitcher:
 
         shape = self._mosaic.save(mosaic_dat_path)
 
-        self.write_metadata(path, shape)
[…]

### 2024-11-08 code/utils.py

@@ -1,0 +1,10 @@
+import json
+import os
+
+def load_config(path = "."):
+    config_dir = os.path.join(path, "config.json")
+    f = open(config_dir)
+    config = json.load(f)
+    f.close()
+
+    return config["config"]

### 2024-11-08 code/camera.py
Running fine but we're not using crop... string interpolation seems weird for Height and Width

@@ -24,7 +24,7 @@ class Camera:
         # Create the pipeline with both display and save frame functionality
         
         self.pipeline = Gst.parse_launch(
-            "nvarguscamerasrc wbmode=1 ee-mode=2 ee-strength=0.5 exposurecompensation=0.5 exposuretimerange='680000000 600000000'  aelock=true ! video/x-raw(memory:NVMM),width={},height={},framerate=30/1 ! videorate ! video/x-raw(memory:NVMM),width=3840,height=2160,framerate=15/1 !".format(W_PIXELS + (2*CROP_W), H_PIXELS + (2*CROP_H)) + #683709000
+            "nvarguscamerasrc wbmode=1 ee-mode=2 ee-strength=0.5 exposurecompensation=0.5 exposuretimerange='680000000 600000000'  aelock=true ! video/x-raw(memory:NVMM),width={},height={},framerate=30/1 ! videorate ! video/x-raw(memory:NVMM),width=3840,height=2160,framerate=15/1 !".format(W_PIXELS + (2*CROP_W), H_PIXELS + (2*CROP_H)) + #683709000 # AM I SURE THIS IS WHERE THE IMAGE SIZE SHOULD CHANGE?
             "nvvideoconvert src-crop='{}:{}:{}:{}' flip-method=2 ! video/x-raw,width={},height={},framerate=15/1 ! videobalance contrast=1.25 ! queue max-size-buffers=1 leaky=2 ! tee name=t ".format( CROP_W, CROP_H, W_PIXELS, H_PIXELS, W_PIXELS, H_PIXELS) +
             "t. ! queue max-size-buffers=1 leaky=2 ! autovideosink "
             "t. ! queue max-size-buffers=1 leaky=2 ! avenc_tiff ! tee name=t_bin ! fakesink"

### 2024-11-11 code/controller.py
Making changes to the way commands are sent to move gantry. XY movements is controlled by precalculated values for the grid, the idea is that there should be better PID control if we always image outward from the center.

@@ -14,10 +14,11 @@ from pathlib import Path
 import json
 from typing import Callable
 import utils
+import numpy as np
 
 
 class Controller:
-    def __init__(self, image_width_mm: float, image_height_mm: float):
+    def __init__(self):
         """Abstraction of the controller which moves the gantry, gets information from the GUI, operates the camera, and determines when to stitch.
 
         Args:
@@ -41,11 +42,11 @@ class Controller:
         self.cookies = []
         self._gantry = gantry.Gantry()
         self.camera = camera.Camera()
-        self.focus = focus.Focus(delete_flag=True, setpoint=round(self.n_images / 2))
+        self.focus = focus.Focus(delete_flag=True, setpoint=math.floor(self.n_images / 2))
 
         #attributes
-        self.image_height_mm = image_height_mm
-        self.image_width_mm = image_width_mm
+        self.image_height_mm = self.config["gui"]["DEFAULT_IMAGE_HEIGHT_MM"]
+        self.image_width_mm = self.config["gui"]["DEFAULT_IMAGE_WIDTH_MM"]
         self.directory = "."
 
 
@@ -108,10 +109,8 @@ class Controller:
             self.set_directory(cookie.directory)
 
             start_time = time.time()
-
-            _, _, z = cookie.get_center_location()
         
-            gantry_thread = Thread(target=self.capture_grid_photos, args=(cookie.coordinates, cookie.directory, focus_queue, pid_queue, pid_lock, cookie.rows, cookie.cols, cookie.y_step_size, cookie.x_step_size, z, self.n_images, self.height_range, progress_callback, stop_capture))
+            gantry_thread = Thread(target=self.capture_grid_photos, args=(cookie.coordinates, cookie.directory, cookie.targets_top, cookie.targets_bot, focus_queue, pid_queue, pid_lock, cookie.rows, cookie.cols, self.n_images, self.height_range, progress_callback, stop_capture))
             focus_thread = Thread(target=self.focus.find_focus, args=(focus_queue, pid_queue, pid_lock, cookie.directory, cookie.nvar, cookie.focus_index, cookie.background, cookie.background_std))
             gantry_thread.start()
             focus_thread.start()
@@ -141,9 +140,9 @@ class Controller:
                 width_est_pixels = cookie.width / cookie.image_width_mm * self.camera.w_pixels 
                 height_est_pixels = cookie.height / cookie.image_height_mm * self.camera.h_pixels
[…]

### 2024-11-11 code/cookie.py

@@ -26,7 +26,7 @@ class Cookie:
         self.width = cookie_width_mm
         self.height = cookie_height_mm
         self.percent_overlap = percent_overlap
-        self._center = (x,y,z)
+        self._center = (round(x, 4),round(y, 4),round(z, 4))
         tl_x = x - (self.width/2)
         tl_y = y + (self.height/2)
         tl_z = z
@@ -43,19 +43,20 @@ class Cookie:
 
         self.image_width_mm = image_width_mm
         self.image_height_mm = image_height_mm
-        self.rows, self.cols, self.x_step_size, self.y_step_size = self.calculate_grid()
+        self.rows, self.cols, self.x_step_size, self.y_step_size = self.calculate_grid_params()
+        self.targets_top, self.targets_bot = self.calculate_image_locations(self.rows, self.cols, self.x_step_size, self.y_step_size) # Each index of the (-1, 5) shape array is (X, Y, Z, row, col)
         self.background = []
         self.background_std = []
         self.coordinates = []
         self.focus_index = []
         self.nvar = []
 
-    def calculate_grid(self):
+    def calculate_grid_params(self):
         """Calculates the amount of rows and columns necessary to traverse the cookie.
 
         Returns:
-            int: Count of rows
-            int: Count of cols
+            int: Count of rows, always odd
+            int: Count of cols, always odd
             float: step size for translation in x direction 
             float: step size for translation in y direction 
         """
@@ -65,10 +66,105 @@ class Cookie:
         x_step_size = self.image_width_mm - overlap_x
         y_step_size = self.image_height_mm - overlap_y 
 
-        cols = math.ceil(self.width / x_step_size)
-        rows = math.ceil(self.height / y_step_size)
+        if x_step_size == 0:
+            cols = 1
+        else:
+            cols = math.ceil(self.width / x_step_size)
+
+        if y_step_size == 0:
+            rows = 1
+        else:
+            rows = math.ceil(self.height / y_step_size)
+
+        # Verify that there are an odd number of rows and columns to make grid calculation easier
+        if cols % 2 != 1:
+            cols += 1
+        
+        if rows % 2 != 1:
+            rows += 1
 
         return rows, cols, x_step_size, y_step_size
+
[…]

### 2024-11-11 code/gui.py

@@ -20,7 +20,7 @@ class App(Gtk.Window):
         self.set_size_request(self.config["gui"]["DEFAULT_WINDOW_SIZE"][0], self.config["gui"]["DEFAULT_WINDOW_SIZE"][1])
         self.connect("destroy", self.quit_program)
         
-        self.controller = controller.Controller(3, 2)
+        self.controller = controller.Controller()
 
         grid = Gtk.Grid()
         grid.set_row_homogeneous(False)

### 2024-11-12 docs/writing/outline.tex
Adding some math equations to methods regarding image focusing
Updating some paper and outline things

@@
 		\end{enumerate}
+	\end{enumerate}
+\item System Overview
+	\begin{enumerate}
+		\item Camera
+		\item computer
+		\item software
+		\item XYZ frame
+		\item 3D Printed adapters
+		\item levelling table
 	\end{enumerate}
@@
 		\item large surface areas can produce files that are inconveniently large
+		\item FIGURE: Sample size vs image capturing time graph
+		\item FIGURE: Scale test image, proves the generated DPI calculation negates need to image with a ruler in frame
 		\end{enumerate}
@@
 		\end{enumerate}
-	\item Images/tables	
+	\item RCNN
 		\begin{enumerate}
-		\item Scale test image, proves the generated DPI calculation negates need to image with a ruler in frame
-		\item Sample size vs image capturing time graph
-		\item Hopefully sample size, dpi, automated tree ring time
-		\item Table with results of Equation for uncompressed filesize, emphasize lossy vs lossless compression, and true filesize %image filesize = DPI x 25.4 x sample-h-mm x sample-w-mm x 8bits x 3 colors
+		\item TABLE: RCNN results, sample height and width, dpi, automated tree ring time
 		\end{enumerate}
@@
 		\begin{enumerate}
+		\item TABLE: with results of Equation for uncompressed filesize, emphasize lossy vs lossless compression, and true filesize %image filesize = DPI x 25.4 x sample-h-mm x sample-w-mm x 8bits x 3 colors
 		\item Software support for large files than 2.5GB >

### 2024-11-12 docs/writing/paper.tex
Updating some paper and outline things

--- 2024-10-31 docs/writing/paper.tex
@@
 \section{Methods}
-The functionality of this system allows for the major steps in tree cookie and core digitization to be automated.
-Capturing images and stitching them together is handled without the need for user intervention.
-To do this, the system traverses the surface area of the sample after being given its dimensions.
-After the capturing process, the images are stitched and are stored on the device along with metadata of relevant machine
-settings and sample parameters.
 
-%% Previous Methods is commented out so I can reorder things 
-% \subsection{Hardware Design}
+\subsection{System Overview} % this is so dry fix later
 
-% \subsubsection{Frame}
-% The cartesian gantry machine was purchased from an open source parts supplier, OpenBuilds. They sell various sets of gantry robots, motor controllers, and power supplies
-% which can be conveniently purchased from one supplier. All metal components are cut and sized, and have very detailed assembly videos, an active community forum,
-% and customer support to fall back on.
-The OpenBuilds ACRO system circumvents the need for most power tools, machining, electrical connectors, and technical knowhow that would otherwise be necessary to build a robust robot.
-% All other parts were designed using SolidWorks and printed on an FDM 3D printer using ABS material. Mechanical assembly can be completed entirely with common hand tools and metric fasteners.
+Tina can be thought of as a combination of multiple subsystems: the camera, computer, and gantry machine. 
+Cartesian movement in the X, Y and Z directions is a result of two kits and a motor controller from OpenBuilds - the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
+The ACRO provides build plates for custom attachments, we made an adapter to combine the linear actuator to the build plate to add a third degree of freedom.
+On the linear actuator's build plate, adapters were made to hold a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens connected to an NVIDIA Jetson Orin Nano (Jetson) through a CSI cable.
+By choosing this combination, we were able to reduce the weight and cost of the camera significantly - trickling the budget into an efficient computer which can handle intensive image processing. 
+The Jetson is a powerful SoC edge computer which drives a computer monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for automatic control, and performs calculations to stitch individual images into one mosaic.
+Despite the weight reductions, torsion on the gantry arm still resulted in a non-zero torsional deflection. An torsion correcting adapter was designed to compensate.
 
-% \subsubsection{Computer}
-% The computer which runs the graphical user interface (GUI), gantry control, camera streaming, and image stitching is the NVIDIA Jetson Orin Nano Developer Kit (Orin Nano). Using a computer
-% which is compatible with CSI cameras such as the Raspberry Pi HQ Camera was essential to maintain cost effectiveness. The Orin Nano, like others in the Jetson line of products, has hardware
-% accelerators for common GPU tasks such as image streaming, which is vital to retain CPU power for GUI tasks and robot control.
+\subsection{Sample Digitization} % Methods
+The subsystems can be best understood by following the path from sample setup through the computing process to create the stitched image.
 
-% \subsubsection{Camera}
-% Unlike the Gigapixel or CaptuRING, the camera is a machine vision camera rather than a handheld professional camera. Choosing the Raspberry Pi HQ Camera provided many lens options, significant
-% forum support, and includes an Obsolescence Statement that the camera will be in production until January 2030.
-Combined with a 180mm C-mount microscope lens, the images can reach above 20,000 dots per inch (DPI).
+To achieve this with a fixed focus camera, the samples must be nearly orthogonal to the camera lens.
+Misalignments between the sample and lens up to 10 degrees can be corrected by using the 3D printed sample levelling table we designed (See Figure \ref{fig:ideal_levelling}).
+Once the sample is level, the operator interacts with the machine through the GUI to navigate the camera to the center of the sample, and initializes focus to be sharp.
+The height and width dimensions are then entered in the GUI
+to be saved along with the center XYZ coordinate of the sample. This procedure can be repeated to create a queue of samples to digitize.
 
-% \subsection{Software Design}
-% All software was written in Python VERSION and a variety of packages for the GUI, image stitching, and calculations (Should I include all the packages in writing?). Care was taken to segment the
-% software design into abstracted object oriented code.
-
-% \subsubsection{Graphical User Interface}
-% The GUI to control the machine launches from a Python script. Included are a viewer to see the live stream from the camera, buttons to jog the machine throughout the XYZ coordinate system,
-% setting the size of samples, and running the process to capture and stitch images. The system was designed to allow for multiple samples to be prepared in a queue for bulk digitization.
-
-% \subsubsection{Sample Digitization}
-% The digitization of a sample can be reduced to a few processes - capturing a grid of overlapping images, image focusing, and image stitching.
-
-% Inside the GUI, the program is populated with the height and width of the sample as well as the height and width of the image. The user then navigates to the center of the sample and can choose 
-% to begin imaging the sample or add the sample to a queue for bulk digitization. This provides the machine enough data to calculate the necessary rows and columns for traversing the entire surface
-% area of the sample while maintaining a level of overlap between adjacent images. Image overlap is what allows for images to be stitched together as there is nonzero error when the robot translates 
-% from one coordinate to another.
-
-% To stitch adjacent images, feature based image stitching from a Python package Stitch2D was used. The underlying functions were based on the processing libraries
-% OpenCV and NumPy.
-Multiple other image stitching packages were tested but none were able to successfully piece together our samples.
-(***Sounds kinda bad to not have data on a comparison as well as not having a metric to see how 'well' an image is stitched...
-but this is non trivial.
-Can we get away with it?***)
-
-% One major difficulty with using the Raspberry Pi HQ Camera is the lack of a lens with auto-focus.
-An in focus image in this regime is dependent on the focal length of the lens set by the user and the
-% distance from the sample.
-The microscope lens has a very small depth of field and goes from in-focus to out-of-focus in less than a 1mm step in the distance to the sample.
-Two cooperative solutions were implemented
-% to achieve a focused image for the entire surface area of the sample - taking images from multiple distances from the sample and automatic control.
-
-% \subsubsection{Image Stack Focusing}
-% When the sample is placed on the table to be digitized, it is assumed that there will, at best, very slight variations in height across the surface area of the sample. A levelling table 
-% assists in reducing this error although it will never be nonzero. At each location (X, Y) on the grid of overlapping images, 11 images are captured in equally distanced Z steps across a 1mm range. 
-% Each of these images have a normalized variance score calculated - the maximum value represents the most in focus image \citep{sampat_extensive_2014}.
-In practice, stopping and starting the motor causes
-% enough vibration to require upwards to two seconds of pause time before the motion blur stops to take each photo. Instead of stopping and starting each image, an additional distance is added to the 
-% Z-motion to allow the stepper to accelerate and reach constant velocity.
-At this speed, images are then taken with constant time steps instead of constant distance steps across the 1mm range.
-This eliminates
-% the motion blur due to acceleration and dramatically reduces imaging time.
-
-$$\frac{1}{MN\mu} \sum\limits_{x=0}^{M-1} \sum\limits_{y=0}^{N-1}(f(x,y) - \mu)^2$$
-
-\subsubsection{Automatic Control}
-While the image stack provides a basis to obtain an in focus image, it is statically limited to the 1mm range in the Z axis. Instead of spending significant time guaranteeing that the sample
-is within the the Z axis range across the entire surface area, a PID control algorithm takes the wheel \citep{odwyer_summary_2000}.
-The information previously calculated in the image stack is sufficient
-to inform the PID controller of how to adjust the initial the Z axis for the next image stack. The controller is constantly trying to make the most in focus image be at the center of the image stack,
-index value 5. This allows for the most flexibility moving to the next location of the image grid.
+Sample height, width, identifiers, and centering an in focus sample is sufficient information to digitize.
 
@@
       \centering
-      \includegraphics[height=0.75\linewidth]{../diagrams/sample_setup_ideal.png}
+      \includegraphics[height=0.5\linewidth]{../diagrams/sample_setup_ideal.png}
       \caption{Ideal sample leveling}
-      \label{fig:sub1}
+      \label{fig:ideal_levelling}
     \end{subfigure}%
@@
       \centering
-      \includegraphics[height=0.75\linewidth]{../diagrams/sample_setup_realistic.png}
+      \includegraphics[height=0.5\linewidth]{../diagrams/sample_setup_realistic.png}
       \caption{Realistic sample leveling}
-      \label{fig:sub2}
+      \label{fig:realistic_levelling}
     \end{subfigure}
     \caption{Side view of the camera and sample on top of a leveling table.
-The ideal sample leveling shows a uniform distance d at all (x,y) coordinates on the sample.
-This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique (x,y) coordinates.}
-    \label{fig:test}
+The ideal sample leveling shows a uniform distance d at all $(X,Y)$ coordinates on the sample.
+This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique $(X_i, Y_i)$.}
+    \label{fig:sample_levelling}
 \end{figure}
+
+\subsection{Image Capturing}
+Once prompted the system begins to exhaustively traverse the surface area of the sample. 
+The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors - the basis of image stitching. 
+Digitizing cores can often be done without needing to traverse in both the X and Y axis as the field of view in an image can capture the core's entire axial diameter in its field of view when moving along the core's length.
+% confusing wording fix later
+Spanning the surface area of a cookie is more complicated as it needs move in both the X and Y axis (see Figure).
+
+By centering the design around a fixed focus camera and lens, it is necessary to implement some sense of automatic control to focus the images. 
+In this system, the only way to control focus is by changing the lens's distance to the sample. 
+And with the microscope lens, the depth of field of the image is so sensitive that a perturbation of less half of a millimeter can move an entire image out of focus.
+Solving this in a time efficient manner involves two stages. 
+The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
+This allows the initial $Z_0$ value from $(X_0, Y_0, Z_0)$ to be a fantastic initial guess as to what value would have an in focus image across the entire surface. 
+Sample surface misalignment and a non flat sample (bumpy) surfaces require more than one image to be taken at each image $(X, Y)$. 
+From the first $(X_0, Y_0)$ coordinate captured, 11 images are taken at different $Z$ values in $\boldsymbol{Z_{\text{focus}}}$. 
+So long as $Z_{sample}$ is within this set, an in focus image exists.
+
+\[
+Z_{sample} \in
+\boldsymbol{Z_{\text{focus}}} = 
+Z_0 + 
+\begin{bmatrix}
+-0.5 \\
+-0.4 \\
+\vdots \\
+0.4 \\
+0.5 \\
+\end{bmatrix}
+\exists 
+Z_{focused}
+\] % unsure if this makes sense
+
+Beginning at 0.5 mm above $Z_0$ and the last image finishing at 0.5 mm below $Z_0$.
+To reduce motion blur in the images, this $\boldsymbol{Z_{\text{focus}}}$ is traversed at constant velocity and images are captured without stopping. 
+The 11 images then have their normalized variance, $NV$, calculated in a separate thread to measure the image sharpness \citep{sampat_extensive_2014}.
+The image with the maximum score is saved while the rest are deleted from storage.
+
+\[
+    i_{\max} = \arg\max_{i} NV(image(\boldsymbol{{Z_\text{focus}})})
+\]
+
+This focusing procedure works well alone when the sample alignment has a difference in height no greater than 0.5 mm, $d_1 - d_2 < 0.5 mm $ (See Figure \ref{fig:realistic_levelling}). 
+And while 1mm may sound like a small range, it is important to note that increases in time at each $(X_k, Y_k)$ are repeated $k$ times.
+
+Rather than adjusting this range, a greater alignment error can be managed by controlling the center of the range - $Z_{0,k}$ for $(X_k,Y_k)$. 
+The likelihood of an adjacent $(X_{k+1}, Y_{k+1})$ containing an in focus image is highest when the current $i_{{\max,k}}$ is at the middle index of $\boldsymbol{Z_{focus}}$. 
+A PID control algorithm with a process variable of $i_{max}$ and control variable of $Z_{0,k}$ allows a negative feedback loop to improve focusing across the entire sample \citep{odwyer_summary_2000}.
+Instead of $(Z_{max} - Z_{min}) < 0.5mm$, now the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which much more forgiving.
+
+After the capturing overlapping images across the entire surface area, the images are ready to be stitched.
+
+\subsection{Stitching}
+
+Image stitching is a well explored field, ranging from panoramic images taken on most smart phones to highly tuned microscopy slide stitching. 
+But the basis of stitching requires adjacent images to have a region of overlap. 
+Of the stitching tools with a software API that we tested, only the python package Stitch2D was able to stitch our images successfully into a grid. % need to reference this
+This package wraps OpenCV functions for finding distinctive image features from SIFT and feature matching \citep{lowe_distinctive_2004}. 
+The default implementation of the package works very well but has a very high memory space complexity and fails due to out-of-memory errors when stitching more than a thousand images.
+%is this wording correct?, I'm trying to kindly say it was memory inefficient
+With a few key memory conscious changes to the algorithm, the package was able to run on the Jetson without a problem - confirming the ability to stitch images with filesizes greater than total RAM.
 

### 2024-11-13 code/focus_stack.py
Removing unnecessary file

@@ -1,313 +1,0 @@
-## Code inspired by https://github.com/cmcguinness/focusstack/tree/master
-
-import numpy as np
-import cv2
-import os
-import logging as log
-import pandas as pd
-
-log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
-
-class FocusStack:
-
-    def __init__(self):
-        pass
-
-    def findHomography(self, image_1_kp, image_2_kp, matches):
-        image_1_points = np.zeros((len(matches), 1, 2), dtype=np.float32)
-        image_2_points = np.zeros((len(matches), 1, 2), dtype=np.float32)
-
-        for i in range(0,len(matches)):
-            image_1_points[i] = image_1_kp[matches[i].queryIdx].pt
-            image_2_points[i] = image_2_kp[matches[i].trainIdx].pt
-
-
-        homography, mask = cv2.findHomography(image_1_points, image_2_points, cv2.RANSAC, ransacReprojThreshold=2.0)
-
-        return homography
-
-    def align_fft(self, images):    
-        ref_img = self.best_laplacian(images)
-
-        ref_img = cv2.cvtColor(ref_img,cv2.COLOR_BGR2GRAY).astype("float32")
-
-        for target in images:
-            target = cv2.cvtColor(target, cv2.COLOR_BGR2GRAY)
-            tar_form = target.astype("float32")
-            shift = cv2.phaseCorrelate(ref_img, tar_form)
-            print(shift)
-
-    def best_laplacian(self, images):
-        best_lap = 0
-        best_img = []
-
-        for img in images:
-            lap = self.compute_laplacian(img)
-            if lap.var() > best_lap:
-                best_lap = lap.var()
-                best_img = img
-
-        return best_img
-
-    def align_images(self, images):
-
-        outimages = []
-
-        detector = cv2.SIFT_create()#contrastThreshold = 0.05, edgeThreshold = 100)
-        
-
-        ref_img = self.best_laplacian(images)
-
-        outimages.append(ref_img)
-        image1gray = cv2.cvtColor(ref_img,cv2.COLOR_BGR2GRAY)
-        image_1_kp, image_1_desc = detector.detectAndCompute(image1gray, None)
-
-        # img=cv2.drawKeypoints(image1gray,image_1_kp,ref_img,flags=cv2.DRAW_MATCHES_FLAGS_DRAW_RICH_KEYPOINTS)
-        # cv2.imwrite('sift_keypoints.jpg',img)
[…]

### 2024-11-14 code/rcode/renamingStichedImages.R
quick code to change names of the files

@@ -1,0 +1,27 @@
+# Set the main directory path (adjust this to your actual main directory location)
+main_directory <- "/Volumes/PortableSSD/cookies"
+
+# Get all .tif files in the directory and subdirectories
+all_files <- dir_ls(main_directory, recurse = TRUE, glob = "*.tif")
+
+# Loop through each file!
+for (file in all_files) {
+  
+  # Grab the path of the folder one level up from the .tif file
+  parent_folder_path <- path_dir(path_dir(file))
+  
+  # Extract the folder name to use as the ID
+  folder_id <- path_file(parent_folder_path)
+  
+  # Construct the new file name by combining the folder ID with "-30per.tif"
+  new_file_name <- paste0(folder_id, "-30per.tif")
+  
+  # Create the full path for the new file name
+  new_file_path <- path(path_dir(file), new_file_name)
+  
+  # Rename the file
+  file_move(file, new_file_path)
+  
+  # Print the old and new file paths for confirmation
+  message("Renamed ", file, " to ", new_file_path)
+}

### 2024-11-20 docs/writing/paper.tex
Having a hard time transitioning well into the results/discussion

@@
 
+Depending on the analysis, it is not necessary to digitize samples at a very high DPI. To address this, a parameter in the machine configuration file 
+allows the user to downsize images before they get stitched together to avoid unnecessary sample detail.
+
 \section{Results}
 \subsection{Scans of Cookies and Cores}
-The microscope lens greatly improved the maximum resolution of the digitization. Resolutions of up to 13,400 DPI were achieved, a large improvement when compared to 
-both high resolution flatbed scanners (Epson® Perfection v750 PRO) and CaptuRING.
+While the theoretical maximum resolution of the system can achieve 32,000 DPI smaller resolutions were preferred in our digitizations to decrease digitization time and filesize.
+This is a significant improvement when compared to the scanning resolution of flatbed scanners (Epson® Perfection v750 PRO), CaptuRING, and Gigapixel: 6,000 DPI, 4,200 DPI, and 19,812 DPI respectively.
+
+So far we have used this machine to digitize approximately 90 tree cookies which diameter's varied between 1 to 12cm.
+While the machine has a work plane of approximately 60cm x 60cm, the limiting factor of a sample is its surface area. 
+This is more of a concern for cookies than for cores, as even a 60cm length core will have less surface area than a 12cm x 12cm cookie. 
+See FIGURE for sample size, DPI, and filesize graph.
+
+A strong feature between this and alternatives is the ability to setup multiple samples with their identifiers into a queue to be digitized. 
+This conveniently makes interactions with the machine in bursts rather than having to time when a sample will be done to begin setting up another. 
+We were able to successfully queue 10 cookie samples to be digitized although we did not stress test this further.
+
+See FIGURE for the digitization and stitching times.
 
 \subsection{Functional Limits}
-With such high resolution, multiple logistics concerns arise. First is the file size. With such a high resolution, the size of the images can become extremely large. A cookie
-5 inches in diameter would result in an image with 67,000 pixels squared. This easily exceeds the 2.5 GB maximum file size in a TIFF - most software for viewing the file will also 
-be incompatible with any file this big as well.
+With such high resolution, file size limitations can arise. One of the most common lossless image filetypes, TIFF, is limited to 2.5GB.
+But even before we reach this limit, software such as ImageJ, CooRecorder, and other image viewers/analysis tools are not designed to handle larger filesizes nearing 2GB.
+This is important to note as this machine is capable of stitching files greater than 2.5GB but the results are stored in binary NumPy memory map files rather than TIFF's. 
+Storing data this way allows the user to either downsize stitched image or view portions of it with their own script.
 
 \section{Discussion}
-Discuss the implications of your results here.
+By taking advantage of a common 3 degree of freedom cartesian machine design, edge computing device, and a microscope camera we were able to design a cost effective and high resolution digitization tool for wood samples.
+We significantly increased the maximum sample length and maximum resolution compared to common alternatives in the field. 
+In addition, the machine design was intended to be readily replicated by labs with minimal engineering experience and equipment.
 
 \subsection{Strengths and Opportunities}
+Our design has tried to minimize the barriers to building this machine in smaller labs. After investing a few hours in the sourcing and assembly of parts, 
+the rest of the implementation has been completed. Digitized samples from this machine have been able to have tree rings registered onto their images using CooRecorder.
+They have also been ran through a deep learning tree ring identification R-CNN model, although this took an extremely long time even with a hypercomputer.
 
 \subsection{Opportunities for Improvement}
+Integration into a cloud data storage system would be a great addition to remove the need for an operator to move the final data with a physical external drive. 
+Further testing with vessel counting models is also an interesting extension to take advantage of high resolution scans of well prepared samples.
 
 \section{Conclusion}
-Summarize your key findings here.
 

### 2025-01-06 docs/writing/paper.tex
Adding some wording changes

@@
 Rather than scanning a whole sample at once, a high resolution camera takes multiple images across the surface of the sample and uses image stitching software techniques to combine them into one ultra high resolution image \citep{muhlich_stitching_2022}.
-This methodology is often seen in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}. 
+Stitching multiple images into one mosaic has been applied in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}. 
 This method requires either the camera objective to move relative to the sample, or the sample move underneath a stationary camera. 
@@
 
-Tina was made to combine the defining features of the previously mentioned machines into one while making the code open-source and chassis open-hardware device. 
-We designed Tina to digitize both cookies and cores, extend the maximum sample length, perform image stitching without operator intervention, while minimizing cost. 
+Tina was made to combine the defining features of the previously mentioned machines into one while making the project open-source and open-hardware. 
+We designed Tina to digitize both cookies and cores, increase the maximum sample length, perform image stitching without operator intervention, all while minimizing cost. 
 Successful functionality of this system is dependent on the major steps in tree cookie and core digitization to be automated. 
@@
 The total cost of the machine is almost comparable in price or less to many professional camera and macro lens combinations.
-Additional savings can also be made when factoring in the cost of a professional stitching software license such as PTGui.
+Additional savings can also be had when factoring in the cost of a professional stitching software license such as PTGui.
 
@@
 Tina can be thought of as a combination of multiple subsystems: the camera, computer, and gantry machine. 
-Cartesian movement in the X, Y and Z directions is a result of two kits and a motor controller from OpenBuilds - the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
-The ACRO provides build plates for custom attachments, we made an adapter to combine the linear actuator to the build plate to add a third degree of freedom.
+Cartesian movement in the X, Y and Z directions is a result of two machine kits and a motor controller from OpenBuilds - the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
+Building on top of kits allowed for quick assembly and prefabricated parts saving development time. 
+To fit on the build plates of the ACRO system, we made an adapter to connect the linear actuator to the X and Y axis, thus creating motion in the Z axis.
 On the linear actuator's build plate, adapters were made to hold a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens connected to an NVIDIA Jetson Orin Nano (Jetson) through a CSI cable.
 By choosing this combination, we were able to reduce the weight and cost of the camera significantly - trickling the budget into an efficient computer which can handle intensive image processing. 
-The Jetson is a powerful SoC edge computer which drives a computer monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for automatic control, and performs calculations to stitch individual images into one mosaic.
-Despite the weight reductions, torsion on the gantry arm still resulted in a non-zero torsional deflection. An torsion correcting adapter was designed to compensate.
+The Jetson is a powerful edge computer which drives a computer monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for automatic control, and performs calculations to stitch individual images into one mosaic.
+Despite the weight reductions, torsion on the gantry arm still resulted in a non-zero torsional deflection. A torsion correcting adapter was designed to counteract this rotation and level the lens.
 
 \subsection{Sample Digitization} % Methods
-The subsystems can be best understood by following the path from sample setup through the computing process to create the stitched image.
+The subsystems can be best understood by following the process from sample setup through the computing process to create the stitched image.
 
@@
 Misalignments between the sample and lens up to 10 degrees can be corrected by using the 3D printed sample levelling table we designed (See Figure \ref{fig:ideal_levelling}).
-Once the sample is level, the operator interacts with the machine through the GUI to navigate the camera to the center of the sample, and initializes focus to be sharp.
+Once the sample is level, the operator interacts with the machine through the GUI to navigate the camera to the center of the sample, and focuses the preview image to be sharp.
 The height and width dimensions are then entered in the GUI
-to be saved along with the center XYZ coordinate of the sample. This procedure can be repeated to create a queue of samples to digitize.
+to be saved along with the detected center coordinate of the sample. This procedure can be repeated to create a queue of samples to digitize.
 
-Sample height, width, identifiers, and centering an in focus sample is sufficient information to digitize.
+The height, width, identifiers, and centering of the sample is sufficient information to digitize.
 
@@
 The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
-This allows the initial $Z_0$ value from $(X_0, Y_0, Z_0)$ to be a fantastic initial guess as to what value would have an in focus image across the entire surface. 
+This allows the initial $Z_0$ value from $(X_0, Y_0, Z_0)$ to be an informed initial guess as to what value would have an in focus image across the entire surface. 
 Sample surface misalignment and a non flat sample (bumpy) surfaces require more than one image to be taken at each image $(X, Y)$. 

### 2025-01-27 code/plots/digitization_time_and_size.py
Adding times collected from metadata

@@ -1,17 +1,18 @@
 import matplotlib.pyplot as plt
 
 # Fake data points
-time_minutes = [0, 10, 20, 30, 40]  # Time in minutes
-sample_size_mm2 = [5, 15, 30, 50, 80]  # Sample size in mm^2
+time_minutes = [11, 43.6, 94.5, 120.3, 145.3, 149.6, 325.4]  # Time in minutes
+sample_size_mm2 = [660, 1225, 1225, 1517, 1849, 1980, 4200]  # Sample size in mm^2
+stitch_time_minutes = [0.4, 1.29, 2.98, 3.65, 4.32, 5.14, 10.28]
 
 # Create the plot
 plt.figure(figsize=(8, 6))
-plt.plot(time_minutes, sample_size_mm2, marker='o', linestyle='-', color='b')
-
+plt.plot(time_minutes, sample_size_mm2, marker='o', linestyle='-', color='b', label = "Imaging Time")
+plt.plot(stitch_time_minutes, sample_size_mm2, marker = "x", color = 'r', label = "30% Downscale Stitching Time")
 # Add labels and title
 plt.xlabel('Time (minutes)', fontsize=12)
 plt.ylabel('Sample Size (mm$^2$)', fontsize=12)
-plt.title('Time vs Sample Size', fontsize=14)
+plt.title('Time to Digitize Samples by Surface Area', fontsize=14)
 
 # Add grid and legend
 plt.grid(True, linestyle='--', alpha=0.7)

### 2025-01-29 docs/writing/outline.tex
Adding updates to paper and outline that I kept locally for too long

@@
 		\begin{enumerate}
-		\item ultra high resolution scans (DPI 13,500 +)
-		\item downscaled versions as high resolution is not necessary for all applications
+		\item ultra high resolution scans (DPI 22,000+)
+		%% TODO: Measure DPI at 4.5 zoom as well but explain that capturing time and is inconveniently long and files are huge 
+		\item DPI calculated with a slide microscopy scale with 0.01mm precision
+		%% TODO: Figure showing the scale and the tool to measure DPI. 
+		\item Can set the machine to computationally downscale images to different DPI values without changing zoom
 		\item Digitization time and file size as a function of surface area. Large cookies can scan only a portion of the cookie including the center and half of the rings.
@@
 		\item FIGURE: Sample size vs image capturing time graph
+		%% TODO: get all the JSON files from scanning together in one folder for me to analyze
+		%% TODO: imaging _time = num_images * mu AND num_images = sample_height / (image_height * (1 - percent_overlap)) + sample_width / (image_width * (1 - percent_overlap))
+		%%  	 SEE IF ABOVE MODEL IS CORRECT WITH THE METADATA
 		\item FIGURE: Scale test image, proves the generated DPI calculation negates need to image with a ruler in frame
@@
 		\item max filesize for TIFF files is 2.5 GB
+		\item the final file is compressed with lossless LZW compression
 		\item other lossless filetypes for larger images are not compatible with standard image viewers
-		\item workaround with NumPy memory maps and cropped viewing 
+		\item workaround with NumPy memory maps and cropped viewing but there isn't a solid large image viewer 
+		%% TODO: create a python script which opens the range of the image file but skips a lot of rows and columns 
+		%% 		 when opening a .dat file so that you can preview 
 		\end{enumerate}

### 2025-01-29 docs/writing/paper.tex
Adding updates to paper and outline that I kept locally for too long

@@
 % Packages
-\usepackage[utf8]{inputenc}     % For UTF-8 encoding
-\usepackage{amsmath, amssymb}   % For math symbols
-\usepackage{graphicx}           % For including graphics
-\usepackage{geometry}           % For setting page margins
-\usepackage[numbers]{natbib}             % For citations
-\usepackage{hyperref}           % For clickable links in references
-\usepackage{subcaption}         % For side by side subfigures
+\usepackage[utf8]{inputenc}     
+\usepackage{amsmath, amssymb} 
+\usepackage{graphicx}          
+\usepackage{geometry}           
+\usepackage[numbers]{natbib}            
+\usepackage{hyperref}          
+\usepackage{subcaption}         
 % Page layout
-\geometry{margin=1in}           % Set margins to 1 inch
+\geometry{margin=1in}
 
 % Title section
-\title{Tina Paper In Development}
+\title{TIM: Tree Imaging Machine}
 \author{Your Name}
@@
 \begin{abstract}
-    This is the abstract of your paper. Briefly describe the purpose of the research, the main results, and the conclusions.
+The Tree Imaging Machine (TIM) is a do-it-yourself scanning tool made to digitize tree cookies and cores.
+TIM takes partially overlapping microscopic images of samples and stitches the individual images together to form
+a mosaic.
+These scans can produce multiple gigabyte stitched images which can be zoomed in to see distinct features on the scale of 0.01mm.
+With scans of up to 21,140 DPI, TIM produces one of the highest resolution images among similar tools.
+We designed TIM to have a large working plane to allow the digitization of individual cores over 50cm and to allow the operator to set a batch of samples to be digitized without intervention. 
+All of the 3D printed parts are available for download on the NIH 3D printing repository and the software is 
+open source. Instructions to recreate this tool can be found on the code repository. 
 \end{abstract}
@@
 For ATRICS and a more modern do-it-yourself alternative, CaptuRING, the sample is moved relative to the camera \citep{garcia-hidalgo_capturing_2022}. 
-Gigapixel takes a different approach by moving the camera relative to the sample, allowing for multiple samples to be recorded in sequence. 
+Gigapixel takes a different approach by moving the camera relative to the sample, allowing for multiple samples to be recorded in sequence \citep{griffin_gigapixel_2021}. 
 While these machines can all digitize cores, none have been shown to digitize cookies.
 
-Tina was made to combine the defining features of the previously mentioned machines into one while making the project open-source and open-hardware. 
-We designed Tina to digitize both cookies and cores, increase the maximum sample length, perform image stitching without operator intervention, all while minimizing cost. 
+TIM was made to combine the defining features of the previously mentioned machines into one while making the project open-source and open-hardware. 
+We designed TIM to digitize both cookies and cores, increase the maximum sample length, perform image stitching without operator intervention, all while minimizing cost. 
 Successful functionality of this system is dependent on the major steps in tree cookie and core digitization to be automated. 
-The only specialized piece of equipment needed to build Tina is a 3D printer, but the parts can be readily ordered through 3D print shops if preferred.
+The only specialized piece of equipment needed to build TIM is a 3D printer, but the parts can be readily ordered through 3D print shops if preferred.
 Excluding 3D printed parts, the total cost of the machine is approximately \$2,200 USD compared to the \$70,000 USD of the Gigapixel \citep{griffin_gigapixel_2021}.
@@
 
-Tina can be thought of as a combination of multiple subsystems: the camera, computer, and gantry machine. 
+TIM can be thought of as a combination of multiple subsystems: the camera, computer, and gantry machine. 
 Cartesian movement in the X, Y and Z directions is a result of two machine kits and a motor controller from OpenBuilds - the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
-Building on top of kits allowed for quick assembly and prefabricated parts saving development time. 
+Building on top of kits allowed for quick assembly with sound instructions and saved development time. 
 To fit on the build plates of the ACRO system, we made an adapter to connect the linear actuator to the X and Y axis, thus creating motion in the Z axis.
@@
 
+\subsection{Preparing Cookies and Cores}
+TIM can be used to scan both tree cookies and tree cores. Sample preparation for both are equally important but require different approaches.
+The scanning surface of the cookies and cores need to be sanded with incrementally increasing sand paper grit.
+Our cookies were sanded using a belt sander from 60 to 400 grit and polished with 1200 grit by hand.
+Sanding cores was a mix of belt sander and a small orbital sander using the same grit sequence.
+
+Special sanding considerations need to be taken for each sample type. 
+For cookies, it is important to have the top and bottom surfaces be nearly parallel.
+Small differences in plane angle can be corrected using the levelling table (See Figure \ref{fig:ideal_levelling}).
+For cores, care should be taken to maximize the scanning surface. This means removing material to be coincident to the center of the cross section of the core. 
+Cores also need to be aligned to be parallel to the X axis to be digitized, a squaring tool helps guarantee this. Any warping in core mounts need to be counteracted with the use of clamps.
+
 \subsection{Sample Digitization} % Methods
 The subsystems can be best understood by following the process from sample setup through the computing process to create the stitched image. 
-
 To achieve this with a fixed focus camera, the samples must be nearly orthogonal to the camera lens.
 Misalignments between the sample and lens up to 10 degrees can be corrected by using the 3D printed sample levelling table we designed (See Figure \ref{fig:ideal_levelling}).
-Once the sample is level, the operator interacts with the machine through the GUI to navigate the camera to the center of the sample, and focuses the preview image to be sharp.
+Once the sample is level, the operator interacts with the machine through the GUI to navigate the camera to the center of the sample, and focuses the preview image to be sharp by moving in the Z axis.
 The height and width dimensions are then entered in the GUI
 to be saved along with the detected center coordinate of the sample. This procedure can be repeated to create a queue of samples to digitize.
-
 The height, width, identifiers, and centering of the sample is sufficient information to digitize.
@@
 \subsection{Image Capturing}
-Once prompted the system begins to exhaustively traverse the surface area of the sample. 
+In the GUI the operator indicates the height and width of the field of view in an individual image from the camera. 
+This varies on the focal length of the lens, but we have kept the field of view to be 3mm x 5mm. 
+Once prompted the system begins to exhaustively traverse the surface area of the sample, using the sample height and width entered by the operator.  
 The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors - the basis of image stitching. 
-Digitizing cores can often be done without needing to traverse in both the X and Y axis as the field of view in an image can capture the core's entire axial diameter in its field of view when moving along the core's length.
-% confusing wording fix later
-Spanning the surface area of a cookie is more complicated as it needs move in both the X and Y axis (see Figure).
+Digitizing cores can be done without needing to traverse in both the X and Y axis. 
+The optimal scanning path for a core is to ignore the Y axis and traverse the cores length.
+Spanning the surface area of a cookie requires movement in both the X and Y axis.
 
 By centering the design around a fixed focus camera and lens, it is necessary to implement some sense of automatic control to focus the images. 
-In this system, the only way to control focus is by changing the lens's distance to the sample. 
+The only way to control the focus of a fixed focus lens is by changing the lens's distance to the sample. 
 And with the microscope lens, the depth of field of the image is so sensitive that a perturbation of less half of a millimeter can move an entire image out of focus.
@@
 The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
-This allows the initial $Z_0$ value from $(X_0, Y_0, Z_0)$ to be an informed initial guess as to what value would have an in focus image across the entire surface. 
-Sample surface misalignment and a non flat sample (bumpy) surfaces require more than one image to be taken at each image $(X, Y)$. 
-From the first $(X_0, Y_0)$ coordinate captured, 11 images are taken at different $Z$ values in $\boldsymbol{Z_{\text{focus}}}$. 
-So long as $Z_{sample}$ is within this set, an in focus image exists.
+This allows the initial $Z_0$ value from $(X_0, Y_0, Z_0)$ to be an informed guess as to what value would have an in focus image across the entire surface. 
+From the first $(X_0, Y_0)$ coordinate captured, 11 images are taken at different $Z$ values in $\boldsymbol{Z_{\text{samples}}}$. 
+So long as $Z_{focused}$ is within this set, an in focus image exists.
 
 \[
-Z_{sample} \in
-\boldsymbol{Z_{\text{focus}}} = 
+Z_{focused} \in
+\boldsymbol{Z_{\text{samples}}} = 
 Z_0 + 
@@
 \end{bmatrix}
-\exists 
-Z_{focused}
+% \exists 
+% Z_{focused}
 \] % unsure if this makes sense
@@
 Beginning at 0.5 mm above $Z_0$ and the last image finishing at 0.5 mm below $Z_0$.
-To reduce motion blur in the images, this $\boldsymbol{Z_{\text{focus}}}$ is traversed at constant velocity and images are captured without stopping. 
+To reduce motion blur in the images, the range of $\boldsymbol{Z_{\text{samples}}}$ is traversed at constant velocity and images are captured without stopping. 
 The 11 images then have their normalized variance, $NV$, calculated in a separate thread to measure the image sharpness \citep{sampat_extensive_2014}.
@@
 This focusing procedure works well alone when the sample alignment has a difference in height no greater than 0.5 mm, $d_1 - d_2 < 0.5 mm $ (See Figure \ref{fig:realistic_levelling}). 
-And while 1mm may sound like a small range, it is important to note that increases in time at each $(X_k, Y_k)$ are repeated $k$ times.
-
-Rather than adjusting this range, a greater alignment error can be managed by controlling the center of the range - $Z_{0,k}$ for $(X_k,Y_k)$. 
-The likelihood of an adjacent $(X_{k+1}, Y_{k+1})$ containing an in focus image is highest when the current $i_{{\max,k}}$ is at the middle index of $\boldsymbol{Z_{focus}}$. 
+% And while 1mm may sound like a small range, it is important to note that increases in time at each $(X_k, Y_k)$ are repeated $k$ times. 
+Rather than adjusting this range, a greater alignment error can be managed by controlling the center of the range, $Z_{0,k}$, for $(X_k,Y_k)$. 
+The likelihood of an adjacent $(X_{k+1}, Y_{k+1})$ containing an in focus image is highest when the current $i_{{\max,k}}$ is at the middle index of $\boldsymbol{Z_{samples}}$. 
 A PID control algorithm with a process variable of $i_{max}$ and control variable of $Z_{0,k}$ allows a negative feedback loop to improve focusing across the entire sample \citep{odwyer_summary_2000}.
-Instead of $(Z_{max} - Z_{min}) < 0.5mm$, now the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which much more forgiving.
-
+Instead of $(Z_{max} - Z_{min}) < 0.5mm$, now the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which allows for a more forgiving sample setup.
 After the capturing overlapping images across the entire surface area, the images are ready to be stitched.
@@
 But the basis of stitching requires adjacent images to have a region of overlap. 
-Of the stitching tools with a software API that we tested, only the python package Stitch2D was able to stitch our images successfully into a grid. % need to reference this
-This package wraps OpenCV functions for finding distinctive image features from SIFT and feature matching \citep{lowe_distinctive_2004}. 
-The default implementation of the package works very well but has a very high memory space complexity and fails due to out-of-memory errors when stitching more than a thousand images.
+Of the stitching tools with a software API that we tested, only the Python package Stitch2D was able to stitch our images successfully into a grid. % need to reference this
+This package wraps OpenCV functions for finding distinctive image features from the SIFT feature detector and brute force feature matching \citep{lowe_distinctive_2004}. 
+The default implementation of the package works very well but has an $O(n^2)$ space complexity and fails due to out-of-memory errors when stitching hundreds of images.
 %is this wording correct?, I'm trying to kindly say it was memory inefficient
-With a few key memory conscious changes to the algorithm, the package was able to run on the Jetson without a problem - confirming the ability to stitch images with filesizes greater than total RAM.
+With a few key memory conscious changes to the algorithm, the space complexity was reduced to approximately $O(\log{n})$ and the package was able to run on the Jetson without a problem.
 
-Depending on the analysis, it is not necessary to digitize samples at a very high DPI. To address this, a parameter in the machine configuration file 
-allows the user to downsize images before they get stitched together to avoid unnecessary sample detail.
+Depending on the analysis, it is unnecessary to digitize samples at a very high DPI. To address this, a parameter in the machine configuration file 
+allows the user to downsize images before they get stitched together to avoid unnecessary sample detail and decrease the size of the files.
+With the same images, multiple final stitched resolutions can be made.
 
 \section{Results}
-\subsection{Scans of Cookies and Cores}
-While the theoretical maximum resolution of the system can achieve 32,000 DPI smaller resolutions were preferred in our digitizations to decrease digitization time and filesize.
-This is a significant improvement when compared to the scanning resolution of flatbed scanners (Epson® Perfection v750 PRO), CaptuRING, and Gigapixel: 6,000 DPI, 4,200 DPI, and 19,812 DPI respectively.
 
-So far we have used this machine to digitize approximately 90 tree cookies which diameter's varied between 1 to 12cm.
-While the machine has a work plane of approximately 60cm x 60cm, the limiting factor of a sample is its surface area. 
-This is more of a concern for cookies than for cores, as even a 60cm length core will have less surface area than a 12cm x 12cm cookie. 
-See FIGURE for sample size, DPI, and filesize graph.
+\begin{figure}
+  \centering
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_whole.png}
+    \caption{Scale Bar for DPI measurements}
+    \label{fig:dpi_measurement_whole}
+  \end{subfigure}%
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_zoomed_2.png}
+    \caption{Zoomed in Scale Bar}
+    \label{fig:dpi_measurement_zoomed}
+  \end{subfigure}
+  \caption{Example of a single DPI measurement using a 0.01mm slide scale.
+The DPI was measured at multiple locations in the field of view both vertically and horizontally.
+They converged on the same value.}
+  \label{fig:dpi_measurements}
+\end{figure}
 
-A strong feature between this and alternatives is the ability to setup multiple samples with their identifiers into a queue to be digitized. 
-This conveniently makes interactions with the machine in bursts rather than having to time when a sample will be done to begin setting up another. 
-We were able to successfully queue 10 cookie samples to be digitized although we did not stress test this further.
+TIM is capable of producing extremely high resolution scans.
+A microscopy slide scale with 0.01mm graduations was used to calculate the dots per inch (DPI) of individual images captured from the system (Figure \ref{fig:dpi_measurements}).  
+Using this scale we calculated the maximum DPI to be 21,140.
+This calculation can be replicated using the Python tool we created as necessary for varied lens focal lengths.
+When downscaling the final stitch,
+DPI scales linearly with the downscale percentage defined in the machine parameters.
 
-See FIGURE for the digitization and stitching times.
+TIM has been used to digitize 90 cookies and is in the process of finishing 1,200 cores. 
+Digitizing samples takes significantly longer for cookies than it does for cores (Figure \ref{fig:digitization_time}).
+This is due to the surface area of cookies requiring more images as their surface area has a square relationship to sample diameter.
+For large cookies, the operator can choose to selectively scan a portion of the cookie. Scanning cookies at maximum resolution is unrealistic as the max cookie size is very small. 
+To fit into the maximum 2.5 GB limits of a TIFF file, the maximum diameter of a cookie scanned at 30\% of maximum resolution is limited to approximately 130 mm.
+For cores at maximum resolution, this file size limit
+constrains samples to approximately 1,500 mm in length.
+Note that a larger ACRO frame would be needed to support this sample length.
+Samples that are larger than this file size limit are still possible to be scanned but they are no longer stored as a TIFF file.
+An uncompressed binary NumPy memory-mapped array file is produced.
+This implies that the operator knows how to splice into these files and is comfortable working with NumPy memory-mapped arrays in Python.
 
-\subsection{Functional Limits}
-With such high resolution, file size limitations can arise. One of the most common lossless image filetypes, TIFF, is limited to 2.5GB.
-But even before we reach this limit, software such as ImageJ, CooRecorder, and other image viewers/analysis tools are not designed to handle larger filesizes nearing 2GB.
-This is important to note as this machine is capable of stitching files greater than 2.5GB but the results are stored in binary NumPy memory map files rather than TIFF's. 
-Storing data this way allows the user to either downsize stitched image or view portions of it with their own script.
+\begin{figure}
+    \centering
+    \includegraphics[height=0.5\linewidth]{../../code/plots/time_and_area.png}
+    \caption{Time to digitize a sample is dependent on the sample's surface area and the desired final resolution.
+    Cores benefit from linear surface area to sample length.
+The range of sample included are from a 3mm x 220mm core to a 75mm x 56mm cookie.
+True sampling times are drastically influenced by configurable machine parameters.}
+    \label{fig:digitization_time}
+\end{figure}
 
 \section{Discussion}
-By taking advantage of a common 3 degree of freedom cartesian machine design, edge computing device, and a microscope camera we were able to design a cost effective and high resolution digitization tool for wood samples.
+By taking advantage of a common 3 degree of freedom cartesian machine design, powerful edge computing, and a microscope camera we were able to design a cost effective and high resolution digitization tool for wood samples.
 We significantly increased the maximum sample length and maximum resolution compared to common alternatives in the field. 
@@
 the rest of the implementation has been completed. Digitized samples from this machine have been able to have tree rings registered onto their images using CooRecorder.
-They have also been ran through a deep learning tree ring identification R-CNN model, although this took an extremely long time even with a hypercomputer.
-
-\subsection{Opportunities for Improvement}
 Integration into a cloud data storage system would be a great addition to remove the need for an operator to move the final data with a physical external drive. 
-Further testing with vessel counting models is also an interesting extension to take advantage of high resolution scans of well prepared samples.
-
-\section{Conclusion}
+Further testing with vessel counting models or ring identification models would be an interesting extension to take advantage of high resolution scans.
 

### 2025-01-30 code/controller.py
Stub for multiple x's to find best image

@@ -20,10 +20,6 @@ import numpy as np
 class Controller:
     def __init__(self):
         """Abstraction of the controller which moves the gantry, gets information from the GUI, operates the camera, and determines when to stitch.
-
-        Args:
-            image_width_mm (float): Initial image width
-            image_height_mm (float): Initial image height
         """
         #Settings for capturing images from multiple distances
         # self.n_images = 11 #make sure you're not going faster than the frame rate of the GStreamer feed... 
@@ -229,94 +225,19 @@ class Controller:
 
                 self._gantry.block_for_jog()
                 coordinates.append(self._gantry.get_xyz())
-                img_filenames = self.capture_images_multiple_distances(d, n_images, self._gantry.feed_rate_z, height_range, self.acceleration_buffer, row, col)
+                img_filenames = self.capture_images_multiple_z(d, n_images, self._gantry.feed_rate_z, height_range, self.acceleration_buffer, row, col)
                 focus_queue.put(img_filenames)
                 img_num += 1
 
                 elapsed_time = time.time() - start_stack
                 progress_callback((elapsed_time, img_num, rows*cols))
             
-
             pid_queue.task_done() # pretty sure we don't need this 
             focus_queue.put([-1])
             break
         
-            
-            # for row in range(rows):
-            #     if stop_capture.is_set():
-            #         break
-            #     # for last column, we only want to take photo, not move.
-            #     for col in range(cols - 1):
-            #         start_stack = time.time()
-            #         # Odd rows go left
-            #         if stop_capture.is_set():
-            #             break
-                    
-            #         current_col = None
-            #         if row % 2 == 1:
-            #             current_col = cols - col - 1
-            #             coordinates.append(self._gantry.get_xyz())
-            #             imgs = self.capture_images_multiple_distances(d, n_images, self._gantry.feed_rate_z, height_range, self.acceleration_buffer, row, current_col, z_start)
-            #             pid_lock.acquire()
-            #             self.jog_relative_x(-x_dist)
-            #             pid_lock.release()
-            #         # Even rows go right
-            #         else:
-            #             current_col = col
-            #             coordinates.append(self._gantry.get_xyz())
-            #             imgs = self.capture_images_multiple_distances(d, n_images, self._gantry.feed_rate_z, height_range, self.acceleration_buffer, row, current_col, z_start)
[…]

### 2025-01-30 code/focus.py
Making images be image_filenames for readibility

@@ -60,8 +60,7 @@ class Focus:
                 log.info("Background detected, ignore focus index score")
             focus_queue.task_done()
             pid_lock.release()
-	    
-        
+    
     def compute_variance(self, image):
         # adapted from macro info at https://imagejdocu.list.lu/macro/normalized_variance
         mean = np.mean(image)
@@ -71,12 +70,12 @@ class Focus:
         normVar = b/(height * width * mean)
         return normVar
 
-    def best_focused_image(self, images):
+    def best_focused_image(self, image_filenames):
         best_image_filepath = None
         best_var = 0
         vars = []
 
-        for image_name in images:
+        for image_name in image_filenames:
             image = cv2.imread(image_name, cv2.IMREAD_GRAYSCALE)
             if type(image) == np.ndarray:
                 var = self.compute_variance(image)

### 2025-01-31 code/controller.py
Functional core centering for vertically aligned cores. It could use some tuning but honestly it worked pretty good as is on 20 cores.

@@ -157,8 +157,6 @@ class Controller:
         Args:
             frame_dir (str): Directory of where the frames are. 
         """
-        # sizes = [0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0]
-        # sizes = self.config["controller"]["STITCH_SIZES"]
         for size in self.stitch_sizes:
             st = stitcher.Stitcher(frame_dir) 
             #st.stitch(resize=size)
@@ -176,7 +174,7 @@ class Controller:
                 st.delete_dats()
                 del st
     
-    def capture_grid_photos(self, coordinates: list, d: Path, targets_top:np.array, targets_bot: np.array, focus_queue: queue.Queue, pid_queue: queue.Queue, pid_lock, rows:int, cols:int, n_images: int, height_range: float, progress_callback: Callable, stop_capture: Event):
+    def capture_grid_photos(self, coordinates: list, d: Path, targets_top:np.array, targets_bot: np.array, focus_queue: queue.Queue, pid_queue: queue.Queue, pid_lock, rows:int, cols:int, n_images: int, height_range: float, progress_callback: Callable, stop_capture: Event, is_core: bool = True, is_vertical: bool = True):
         """Command to traverse the sample and capture an image at each location. 
 
         Args:
@@ -196,6 +194,8 @@ class Controller:
             height_range (float): The range between the maximum and minimum z distance when trying to find an in focus image
             progress_callback (Callable): GUI widget to update progress bar
             stop_capture (Event): Event to stop capturing when possible
+            is_core (bool): Is the sample a core? 
+            is_vertical (bool): Only matters if is_core == True. Is the core vertically aligned?
         """
         # for loop capture
         # Change feed rate back to being slow
@@ -203,6 +203,7 @@ class Controller:
         while True and not stop_capture.is_set():
             img_num = 0
 
+            # Targets are XYZ coordinates to jog to to capture an image.
             targets = np.vstack((targets_top, targets_bot))
             for target in targets:
                 start_stack = time.time()
@@ -211,10 +212,31 @@ class Controller:
 
                 x, y, z, row, col = target[0], target[1], target[2], int(target[3]), int(target[4])
 
-                if img_num == 0 or img_num == len(targets_top):
+                # Jog to the origin of the core/cookie
+                if img_num == 0:
                     self._gantry.jog_absolute_xyz(x, y, z)
+
[…]

### 2025-01-31 code/focus.py

@@ -70,7 +70,7 @@ class Focus:
         normVar = b/(height * width * mean)
         return normVar
 
-    def best_focused_image(self, image_filenames):
+    def best_focused_image(self, image_filenames, delete = False):
         best_image_filepath = None
         best_var = 0
         vars = []
@@ -86,7 +86,12 @@ class Focus:
             
         std = np.std(vars)
 
+        log.info(image_filenames)
+
+        if delete:
+            self.delete_unfocused(image_filenames)
         return best_image_filepath, std, vars
+        
 
     def delete_unfocused(self, images_to_delete):
         for file_name in images_to_delete:
@@ -109,3 +114,75 @@ class Focus:
 
     ''' pid moment:
         need to update from controller, methods go in focus? '''
+
+
+if __name__ == "__main__":
+
+    import glob
+    import matplotlib.pyplot as plt
+
+    filename_grep_1 = "C:\\code\\TreeRings\\code\\images_1\\image_*"
+    filename_grep_2 = "C:\\code\\TreeRings\\code\\images_2\\image_*"
+    filename_grep_3 = "C:\\code\\TreeRings\\code\\images_3\\image_*"
+    filename_grep_4 = "C:\\code\\TreeRings\\code\\images_4\\image_*"
+    filename_grep_5 = "C:\\code\\TreeRings\\code\\images_5\\image_*"
+
+    filenames = glob.glob(filename_grep_2)
+    # filenames.extend(glob.glob(filename_grep_2))
+    # filenames.extend(glob.glob(filename_grep_3))
+    # filenames.extend(glob.glob(filename_grep_4))
+    # filenames.extend(glob.glob(filename_grep_5))
+
+
+    def compute_variance(image):
+        # adapted from macro info at https://imagejdocu.list.lu/macro/normalized_variance
+        mean = np.mean(image)
+        width, height = image.shape
+        square_diff = (image - mean)**2
+        b = np.sum(square_diff)
+        normVar = b/(height * width * mean)
+        return normVar
+    
+    def horizontal_focus_distribution(filename, n_cols):
+        """Split a single image into n_cols and calculate the normalized variance for each column.
+
+        Args:
+            filename (_type_): Filename of the image to analyze.
+            n_cols (_type_): Number of columns to split the image into. Must be even to split correctly.
+        """
+        if n_cols % 2 == 1:
+            n_cols += 1
+
+        image = cv2.imread(filename, cv2.IMREAD_GRAYSCALE)
[…]

### 2025-01-31 code/gantry.py

@@ -116,55 +116,88 @@ class Gantry:
         self.s.write(str.encode("{}\n".format(cmd))) # Send g-code block to grbl
         return read_response(self.s)
 
-    def jog_absolute_xyz(self, x, y, z) -> None:
-        cmd = "$J=G90 G21 X{} F{}".format(x, self.feed_rate_xy)
+    def jog_absolute_xyz(self, x, y, z, feed = None) -> None:
+        if feed is not None:
+            feed_rate = feed
+        else:
+            feed_rate = self.feed_rate_xy
+
+        cmd = "$J=G90 G21 X{} F{}".format(x, feed_rate)
         self._send_command(cmd)
-        cmd = "$J=G90 G21 Y{} F{}".format(y, self.feed_rate_xy)
+        cmd = "$J=G90 G21 Y{} F{}".format(y, feed_rate)
         self._send_command(cmd)
-        cmd = "$J=G90 G21 Z{} F{}".format(z, self.feed_rate_xy)
+        cmd = "$J=G90 G21 Z{} F{}".format(z, feed_rate)
         self._send_command(cmd)
         
-    def jog_absolute_xy(self, x, y) -> None:
-        cmd = "$J=G90 G21 X{} Y{} F{}".format(x, y, self.feed_rate_xy)
+    def jog_absolute_xy(self, x, y, feed = None) -> None:
+        if feed is not None:
+            feed_rate = feed
+        else:
+            feed_rate = self.feed_rate_xy
+        cmd = "$J=G90 G21 X{} Y{} F{}".format(x, y, feed_rate)
         self._send_command(cmd)
 
-    def jog_absolute_x(self, pos) -> None:
-        cmd = "$J=G90 G21 X{} F{}".format(pos, self.feed_rate_xy)
+    def jog_absolute_x(self, pos, feed = None) -> None:
+        if feed is not None:
+            feed_rate = feed
+        else:
+            feed_rate = self.feed_rate_xy
+        cmd = "$J=G90 G21 X{} F{}".format(pos, feed_rate)
         self._send_command(cmd)
 
-    def jog_absolute_y(self, pos) -> None:
-        cmd = "$J=G90 G21 Y{} F{}".format(pos, self.feed_rate_xy)
+    def jog_absolute_y(self, pos, feed = None) -> None:
+        if feed is not None:
+            feed_rate = feed
+        else:
+            feed_rate = self.feed_rate_xy
[…]

### 2025-01-31 code/controller.py
Changing the wording so that we are capturing samples which can either be a core or cookie instead of only cookies

@@ -1,6 +1,6 @@
 import gantry
 import focus
-import cookie
+import sample
 import camera
 import stitcher
 import logging as log
@@ -35,7 +35,7 @@ class Controller:
         self.fast_feed_rate_z = self.config["controller"]["FAST_FEED_RATE_Z"]
     
         #Objects
-        self.cookies = []
+        self.samples = []
         self._gantry = gantry.Gantry()
         self.camera = camera.Camera()
         self.focus = focus.Focus(delete_flag=True, setpoint=math.floor(self.n_images / 2))
@@ -80,12 +80,12 @@ class Controller:
 
     #### SERPENTINE METHODS ####
     
-    def capture_cookie(self, cookie: cookie.Cookie, progress_callback: Callable, stop_capture: Event):
+    def capture_sample(self, sample: sample.Sample, progress_callback: Callable, stop_capture: Event):
         """Abstraction to execute a capture sequence. This involves moving the the top left of the sample, traversing in a serpentining pattern 
         across the dimensions of the sample. At each step in the grid, multiple images are taken and only the most in focus is kept. 
 
         Args:
-            cookie (cookie.Cookie): Instance of a sample to be captured
+            sample (sample.Sample): Instance of a sample to be captured
             progress_callback (Callable): GUI Callback to update progress bar 
             stop_capture (Event): Event to stop capture after the current image sequence 
         """
@@ -98,16 +98,16 @@ class Controller:
             self.focus.set_setpoint(round(self.n_images/2))
 
             #set directories
-            species = cookie.species
-            id1 = cookie.id1
-            id2 = cookie.id2
+            species = sample.species
+            id1 = sample.id1
+            id2 = sample.id2
 
-            self.set_directory(cookie.directory)
+            self.set_directory(sample.directory)
 
             start_time = time.time()
         
-            gantry_thread = Thread(target=self.capture_grid_photos, args=(cookie.coordinates, cookie.directory, cookie.targets_top, cookie.targets_bot, focus_queue, pid_queue, pid_lock, cookie.rows, cookie.cols, self.n_images, self.height_range, progress_callback, stop_capture))
-            focus_thread = Thread(target=self.focus.find_focus, args=(focus_queue, pid_queue, pid_lock, cookie.directory, cookie.nvar, cookie.focus_index, cookie.background, cookie.background_std))
[…]

### 2025-01-31 code/gantry.py

@@ -21,10 +21,6 @@ class Gantry:
         self.feed_rate_z = self.config["gantry"]["FEED_RATE_DEFAULT_Z"]
         self.feed_rate_xy = self.config["gantry"]["FEED_RATE_DEFAULT_XY"] 
 
-        # sample information
-        self.cookie_samples = []
-        self.core_samples = []
-
         self.s = None
         self.stop_threads = False
 

### 2025-01-31 code/gui.py

@@ -13,7 +13,7 @@ log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
 class App(Gtk.Window):
     def __init__(self):
-        super().__init__(title="Cookie Capture")
+        super().__init__(title="Capture")
         self.config = utils.load_config()
 
         self.set_default_size(self.config["gui"]["DEFAULT_WINDOW_SIZE"][0], self.config["gui"]["DEFAULT_WINDOW_SIZE"][1])
@@ -40,19 +40,19 @@ class App(Gtk.Window):
     def create_entries(self, grid):
         ## Sample
         
-        frame_entry_cookie = Gtk.Frame(label="Sample Entries")
-        frame_entry_cookie.set_size_request(-1,-1)
-        frame_entry_cookie.set_hexpand(True)
-        frame_entry_cookie.set_vexpand(True)
-        frame_entry_cookie.set_halign(Gtk.Align.FILL)
-        frame_entry_cookie.set_valign(Gtk.Align.FILL)
-        grid.attach(frame_entry_cookie, 1, 0, 1, 1)
+        frame_entry_sample = Gtk.Frame(label="Sample Entries")
+        frame_entry_sample.set_size_request(-1,-1)
+        frame_entry_sample.set_hexpand(True)
+        frame_entry_sample.set_vexpand(True)
+        frame_entry_sample.set_halign(Gtk.Align.FILL)
+        frame_entry_sample.set_valign(Gtk.Align.FILL)
+        grid.attach(frame_entry_sample, 1, 0, 1, 1)
         
-        box_cookie = Gtk.Box(orientation=Gtk.Orientation.VERTICAL, spacing=7)
-        frame_entry_cookie.add(box_cookie)
+        box_sample = Gtk.Box(orientation=Gtk.Orientation.VERTICAL, spacing=7)
+        frame_entry_sample.add(box_sample)
         
-        self.create_cookie_height_entry(box_cookie)
-        self.create_cookie_width_entry(box_cookie)
+        self.create_sample_height_entry(box_sample)
+        self.create_sample_width_entry(box_sample)
         
         ## Machine 
         
@@ -100,10 +100,10 @@ class App(Gtk.Window):
         frame_buttons1.add(box_buttons1)
 
         
-        self.create_capture_all_cookies_button(box_buttons1)
-        self.create_add_cookie_dialog_button(box_buttons1)
+        self.create_capture_all_samples_button(box_buttons1)
[…]

### 2025-01-31 code/sample.py

@@ -1,0 +1,196 @@
+import cv2
+import time
+import numpy as np
+import math
+from pathlib import Path
+from datetime import datetime
+
+class Sample:
+    def __init__(self, sample_width_mm: int, sample_height_mm: int, species:str, id1:str, id2:str, notes:str, image_width_mm:float, image_height_mm:float, is_core:bool,  percent_overlap:int = 50, is_vertical:bool = True, x:float = None, y:float = None, z:float = None):
+        """Class which contains the necessary information for each sample that needs to be digitized. Data comes from user inputs from GUI. Data is saved in metadata.json files alongside stitched images
+
+        Args:
+            sample_width_mm (int): Width of minimum bounding rectangle around edges of the sample
+            sample_height_mm (int): Height of minimum bounding rectangle around the edges of the sample
+            species (str): Species descriptor
+            id1 (str): Additional species ID1
+            id2 (str): Additional specied ID 2
+            notes (str): Additional notes for ambiguity in sample
+            image_width_mm (float): Width in mm of the field of view of a single image. Helpful to print a 1mm x 1mm grid to measure
+            image_height_mm (float): Height in mm of the field of view of a single image. Helpful to print a 1mm x 1mm grid to measure
+            is_core (bool): Is the sample a core? 
+            percent_overlap (int, optional): Desired overlap between two images. Defaults to 50.
+            is_vertical (bool, optional): If is_core, is the length of the core aligned vertically? Defaults to True.
+            x (float, optional): X position of the center of the sample in the gantry coordinate system. Defaults to None.
+            y (float, optional): Y position of the center of the sample in the gantry coordinate system. Defaults to None.
+            z (float, optional): Z position of the center of the sample in the gantry coordinate system. Defaults to None.
+        """
+        self.width = sample_width_mm
+        self.height = sample_height_mm
+        self.percent_overlap = percent_overlap
+        self._center = (round(x, 4),round(y, 4),round(z, 4))
+        tl_x = x - (self.width/2)
+        tl_y = y + (self.height/2)
+        tl_z = z
+        self._top_left = (tl_x, tl_y, tl_z)
+        self.species = species
+        self.id1 = id1
+        self.id2 = id2
[…]

### 2025-01-31 code/gui.py
Adding core to metadata dialog box

@@ -389,7 +389,7 @@ class App(Gtk.Window):
     def cb_add_sample_dialog(self, widget):
         width = int(self.entry_width_sample.get_text())
         height = int(self.entry_height_sample.get_text())
-        overlap, species, id1, id2, notes = self.show_metadata_dialog()
+        overlap, species, id1, id2, notes, is_core = self.show_metadata_dialog()
         if species == False:
             return
         if overlap == '':
@@ -399,7 +399,7 @@ class App(Gtk.Window):
             species = species.replace(" ", "_")
             id1 = id1.replace(" ", "_")
             id2 = id2.replace(" ", "_")
-        self.controller.add_sample(width, height, overlap, species, id1, id2, notes)
+        self.controller.add_sample(width, height, overlap, species, id1, id2, notes, is_core)
         log.info("Adding Sample \nW: {}\nH: {}\nO: {}\nS:  {}\nID1:  {}\nID2:  {}\nNotes:  {}\n".format(width, height, overlap, species, id1, id2, notes))
     
     def show_metadata_dialog(self):
@@ -420,6 +420,7 @@ class App(Gtk.Window):
         id2_label = Gtk.Label(label="ID 2")
         id2_entry = Gtk.Entry()
         notes_label = Gtk.Label(label="Notes")
+        is_core_button = Gtk.CheckButton(label = "Core Sample")
         
         #Create larger text box for notes 
         text_view = Gtk.TextView()
@@ -440,6 +441,7 @@ class App(Gtk.Window):
         box.add(overlap_entry)
         box.add(notes_label)
         box.add(notes_window)
+        box.add(is_core_button)
 
         species_label.show()
         species_entry.show()
@@ -451,9 +453,10 @@ class App(Gtk.Window):
         overlap_entry.show()
         notes_label.show()
         notes_window.show_all()
+        is_core_button.show()
 
-        overlap_entry.set_text("50")
-
+        overlap_entry.set_text("{}".format(self.config["gui"]["DEFAULT_PERCENT_OVERLAP"]))
+        is_core_button.set_active(True) #default to core
         # Run the dialog and capture the response
         response = dialog.run()
 
@@ -464,6 +467,7 @@ class App(Gtk.Window):
             id2 = id2_entry.get_text()
             buffer = text_view.get_buffer()
             notes = buffer.get_text(buffer.get_start_iter(), buffer.get_end_iter(), True)
+            is_core = is_core_button.get_active()
             log.info("Adding Sample")
         else:
[…]

### 2025-01-31 code/cookie.py
Changing from cookie to sample

@@ -1,192 +1,0 @@
-import cv2
-import time
-import numpy as np
-import math
-from pathlib import Path
-from datetime import datetime
-
-class Cookie:
-    def __init__(self, cookie_width_mm: int, cookie_height_mm: int, species:str, id1:str, id2:str, notes:str, image_width_mm:float, image_height_mm:float, percent_overlap:int = 50, x:float = None, y:float = None, z:float = None):
-        """Class which contains the necessary information for each sample that needs to be digitized. Data comes from user inputs from GUI. Data is saved in metadata.json files alongside stitched images
-
-        Args:
-            cookie_width_mm (int): Width of minimum bounding rectangle around edges of the sample
-            cookie_height_mm (int): Height of minimum bounding rectangle around the edges of the sample
-            species (str): Species descriptor
-            id1 (str): Additional species ID1
-            id2 (str): Additional specied ID 2
-            notes (str): Additional notes for ambiguity in sample
-            image_width_mm (float): Width in mm of the field of view of a single image. Helpful to print a 1mm x 1mm grid to measure
-            image_height_mm (float): Height in mm of the field of view of a single image. Helpful to print a 1mm x 1mm grid to measure
-            percent_overlap (int, optional): Desired overlap between two images. Defaults to 50.
-            x (float, optional): X position of the center of the sample in the gantry coordinate system. Defaults to None.
-            y (float, optional): Y position of the center of the sample in the gantry coordinate system. Defaults to None.
-            z (float, optional): Z position of the center of the sample in the gantry coordinate system. Defaults to None.
-        """
-        self.width = cookie_width_mm
-        self.height = cookie_height_mm
-        self.percent_overlap = percent_overlap
-        self._center = (round(x, 4),round(y, 4),round(z, 4))
-        tl_x = x - (self.width/2)
-        tl_y = y + (self.height/2)
-        tl_z = z
-        self._top_left = (tl_x, tl_y, tl_z)
-        self.species = species
-        self.id1 = id1
-        self.id2 = id2
-        self.notes = notes
-
-        dirtime = datetime.now().strftime("%H_%M_%S")
-        directory = "./{}_{}_{}_{}".format(species, id1, id2, dirtime)
-        Path(directory).mkdir()
[…]

### 2025-01-31 code/controller.py
Adding config for core centering and adding logs

@@ -106,7 +106,7 @@ class Controller:
 
             start_time = time.time()
         
-            gantry_thread = Thread(target=self.capture_grid_photos, args=(sample.coordinates, sample.directory, sample.targets_top, sample.targets_bot, focus_queue, pid_queue, pid_lock, sample.rows, sample.cols, self.n_images, self.height_range, progress_callback, stop_capture))
+            gantry_thread = Thread(target=self.capture_grid_photos, args=(sample.coordinates, sample.directory, sample.targets_top, sample.targets_bot, focus_queue, pid_queue, pid_lock, sample.rows, sample.cols, self.n_images, self.height_range, progress_callback, stop_capture, sample.is_core))
             focus_thread = Thread(target=self.focus.find_focus, args=(focus_queue, pid_queue, pid_lock, sample.directory, sample.nvar, sample.focus_index, sample.background, sample.background_std))
             gantry_thread.start()
             focus_thread.start()
@@ -218,9 +218,11 @@ class Controller:
 
                     # If the sample is a vertically aligned core, try to center the core in the FOV on the first 
                     if is_core and is_vertical:
+                        log.info("Core centering procedure start.")
                         self._gantry.block_for_jog()
-                        r = 5
-                        filenames = self.capture_images_multiple_x(d, n_images, self._gantry.feed_rate_z, r, self.acceleration_buffer)
+                        r = self.config["controller"]["CORE_CENTERING_RANGE"] #5
+                        n_images_centering = self.config["controller"]["N_IMAGES_CORE_CENTERING"]
+                        filenames = self.capture_images_multiple_x(d, n_images_centering, self._gantry.feed_rate_z, r, self.acceleration_buffer)
                         self.recenter_core_naive(filenames, r, self._gantry.feed_rate_z)
 
                 # Jog x and y and allow PID to handle the Z
@@ -350,6 +352,7 @@ class Controller:
         self._gantry.block_for_jog()
         # Return to original location
         self.jog_relative_x(-1 * (r / 2 + acceleration_buffer), feed=feed_rate)
+        log.info("Jog to original location.")
         self._gantry.block_for_jog()        
 
         return image_filenames
@@ -367,8 +370,9 @@ class Controller:
         i = filenames.index(focused_filename)
         i_middle = len(filenames) // 2 # guaranteed to be middle because n_images must be odd
         damper = 0.75
[…]

### 2025-02-12 docs/writing/paper.tex
Adding updates to paper

@@
 a mosaic.
-These scans can produce multiple gigabyte stitched images which can be zoomed in to see distinct features on the scale of 0.01mm.
-With scans of up to 21,140 DPI, TIM produces one of the highest resolution images among similar tools.
+These mosaics can be zoomed in to see distinct features on the scale of 0.01mm.
+With scans of up to 21,140 DPI, TIM produces one of the highest resolution images among similar tools for the competitive price of less than \$3,000 USD.
 We designed TIM to have a large working plane to allow the digitization of individual cores over 50cm and to allow the operator to set a batch of samples to be digitized without intervention. 
@@
 \section{Introduction}
-The study of tree rings has proven useful across multiple fields, proving to be a reliable subject for reconstructing past climates of regional and local environments, as well as a mechanism to understand tree growth response \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
-To obtain these data from the tree rings, it is necessary to measure tree rings from a tree cookie or core.
+The study of tree rings is useful across multiple fields, proving to be a reliable method for reconstructing past climates of regional and local environments, as well as a mechanism to understand tree growth response \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
+To obtain these data from the tree rings, it is necessary to measure tree rings widths from a tree cookie or core.
 The first high precision tool for this purpose was a stage micrometer, involving a trained technician to incrementally shift a tree core under the objective of a microscope - informing a computer when a new ring is encountered \citep{robinson_microcomputer_nodate}.
@@
 The original technique was as a flatbed scanner which can digitize the entire sample at once \citep{guay_new_1992}. 
-With a top of the line scanner, like the Epson Perfection V850 Pro, it's possible to scan at maximum resolutions of 4800 dpi and scan an area of up to 8.5" x 11.7".
-Analysis which relies on higher resolution larger samples require a different digitization approach.
+With a top of the line scanner, like the Epson Perfection V850 Pro, it's possible to scan at maximum resolutions of 4800 dpi and scan an area of up to 8.5" x 11.7" (21.6cm x 29.7cm).
+Analysis which relies on higher resolution or of larger samples require a different digitization approach.
 
@@
 TIM was made to combine the defining features of the previously mentioned machines into one while making the project open-source and open-hardware. 
-We designed TIM to digitize both cookies and cores, increase the maximum sample length, perform image stitching without operator intervention, all while minimizing cost. 
-Successful functionality of this system is dependent on the major steps in tree cookie and core digitization to be automated. 
+We designed TIM to digitize both cookies and cores, increase the maximum sample length, perform image stitching without operator intervention, and capture batches of samples in a queue - all while minimizing cost.
 The only specialized piece of equipment needed to build TIM is a 3D printer, but the parts can be readily ordered through 3D print shops if preferred.
@@
 Cartesian movement in the X, Y and Z directions is a result of two machine kits and a motor controller from OpenBuilds - the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
+Smaller and larger working planes easily be achieved by purchasing a different sized ACRO kit.
 Building on top of kits allowed for quick assembly with sound instructions and saved development time. 
@@
 By choosing this combination, we were able to reduce the weight and cost of the camera significantly - trickling the budget into an efficient computer which can handle intensive image processing. 
-The Jetson is a powerful edge computer which drives a computer monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for automatic control, and performs calculations to stitch individual images into one mosaic.
+The Jetson is a powerful edge computer which drives a monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for automatic control, and performs calculations to stitch individual images into one mosaic.
 Despite the weight reductions, torsion on the gantry arm still resulted in a non-zero torsional deflection. A torsion correcting adapter was designed to counteract this rotation and level the lens.
@@
 TIM can be used to scan both tree cookies and tree cores. Sample preparation for both are equally important but require different approaches.
-The scanning surface of the cookies and cores need to be sanded with incrementally increasing sand paper grit.
-Our cookies were sanded using a belt sander from 60 to 400 grit and polished with 1200 grit by hand.
-Sanding cores was a mix of belt sander and a small orbital sander using the same grit sequence.
+The scanning surface of the cookies and cores need to be sanded with incrementally increasing sand paper grit. Our cookies were sanded using an orbital sander with mesh sandpaper from 60 to 800 grit.
+We used a microscope to visually inspect the quality of the sanding and when vessels were easily identifiable, we considered the sample to be sufficiently prepared.
 
@@
 For cookies, it is important to have the top and bottom surfaces be nearly parallel.
-Small differences in plane angle can be corrected using the levelling table (See Figure \ref{fig:ideal_levelling}).
+Small differences in plane angle can be corrected using the 3D printed levelling table we designed (See Figure \ref{fig:ideal_levelling}).
 For cores, care should be taken to maximize the scanning surface. This means removing material to be coincident to the center of the cross section of the core. 
-Cores also need to be aligned to be parallel to the X axis to be digitized, a squaring tool helps guarantee this. Any warping in core mounts need to be counteracted with the use of clamps.
+Similarly to cookies, care should be given to achieving a  scanning surface as parallel to the XY-plane as possible.
+Cores also need to be aligned to be parallel to the Y axis to be digitized, using squaring tool or similar to help with alignment.
+Any warping in core mounts need to be counteracted with the use of trigger clamps.
 
 \subsection{Sample Digitization} % Methods
-The subsystems can be best understood by following the process from sample setup through the computing process to create the stitched image. 
+The subsystems can be best understood by following the process from sample setup through through obtaining a stitched image. 
 To achieve this with a fixed focus camera, the samples must be nearly orthogonal to the camera lens. 
-Misalignments between the sample and lens up to 10 degrees can be corrected by using the 3D printed sample levelling table we designed (See Figure \ref{fig:ideal_levelling}).
 Once the sample is level, the operator interacts with the machine through the GUI to navigate the camera to the center of the sample, and focuses the preview image to be sharp by moving in the Z axis.
 The height and width dimensions are then entered in the GUI
-to be saved along with the detected center coordinate of the sample. This procedure can be repeated to create a queue of samples to digitize.
-The height, width, identifiers, and centering of the sample is sufficient information to digitize.
+to be saved along with the detected center coordinate of the sample. This procedure can be repeated to create a queue of samples to digitize as a batch.
+The height, width, identifiers, sample type, and centering of the sample is sufficient information to digitize.
 
@@
 
-By centering the design around a fixed focus camera and lens, it is necessary to implement some sense of automatic control to focus the images. 
+By centering the design around a fixed focus camera and lens, it is necessary to implement automatic control to focus the images. 
 The only way to control the focus of a fixed focus lens is by changing the lens's distance to the sample. 
-And with the microscope lens, the depth of field of the image is so sensitive that a perturbation of less half of a millimeter can move an entire image out of focus.
+And with the microscope lens, the depth of field of the image is so sensitive that sub millimeter heights can move an entire image out of focus.
 Solving this in a time efficient manner involves two stages. 
 The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
-This allows the initial $Z_0$ value from $(X_0, Y_0, Z_0)$ to be an informed guess as to what value would have an in focus image across the entire surface. 
+This allows the initial $Z_0$ value ,captured when adding a sample, to be an informed guess as to what value would have an in focus image across the entire surface. 
 From the first $(X_0, Y_0)$ coordinate captured, 11 images are taken at different $Z$ values in $\boldsymbol{Z_{\text{samples}}}$. 
@@
 
-Beginning at 0.5 mm above $Z_0$ and the last image finishing at 0.5 mm below $Z_0$.
-To reduce motion blur in the images, the range of $\boldsymbol{Z_{\text{samples}}}$ is traversed at constant velocity and images are captured without stopping. 
+Our focusing procedure begins at 0.5 mm above $Z_0$ and finishes with the last image at 0.5 mm below $Z_0$.
 The 11 images then have their normalized variance, $NV$, calculated in a separate thread to measure the image sharpness \citep{sampat_extensive_2014}.
-The image with the maximum score is saved while the rest are deleted from storage.
+The image with the maximum score is saved and can be considered the most in focus image, while the rest are deleted from storage.
 
@@
 
+To reduce motion blur in the images, the range of $\boldsymbol{Z_{\text{samples}}}$ is traversed at constant velocity and images are captured without stopping. 
+Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to improve speed and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
+This is a variation of the standard approach used in 3D printing and CNC control methods.
+
+\subsection{Accounting for Translation Error Between Core Samples}
+
+Difficulties arise when more than one core are added to the digitization queue.
+The ACRO has a theoretical 4.5 micron translational accuracy, but is hard to achieve.
+These inaccuracies can cause drift in the images with relation to the core and result in the width of the core not being within the field of view of the images.
+Specifically for cores, a centering procedure was implemented to realign the center of the core to the center of the image frame when the machine moves 
+to a new sample.
+
+First, the camera is moved to to the $(X_0, Y_0, Z_0)$ of the new core. 
+From there, the camera is moved across a window, at constant velocity, from $(X_0 - 1/2*{range}, Y_0, Z_0)$ to $(X_0 + 1/2*{range}, Y_0, Z_0)$. 
+Images are captured at equal distance intervals analogously to the image focusing procedure. 
+Once again, the image with the highest normalized variance score is considered the best image and its location is used as the realigned $X_0^\prime$.
+This procedure drastically improves digitization quality of batches of cores.
+
 \subsection{Stitching}
 
-Image stitching is a well explored field, ranging from panoramic images taken on most smart phones to highly tuned microscopy slide stitching. 
-But the basis of stitching requires adjacent images to have a region of overlap. 
+Image stitching is a well explored field, ranging from panoramic images taken on smart phones to highly tuned microscopy slide stitching. 
+But the basis of stitching remains the same across algorithms. Adjacent images must have a region of overlap. 
 Of the stitching tools with a software API that we tested, only the Python package Stitch2D was able to stitch our images successfully into a grid. % need to reference this
-This package wraps OpenCV functions for finding distinctive image features from the SIFT feature detector and brute force feature matching \citep{lowe_distinctive_2004}. 
-The default implementation of the package works very well but has an $O(n^2)$ space complexity and fails due to out-of-memory errors when stitching hundreds of images.
+This package wraps OpenCV functions for finding distinctive image features using the SIFT feature detector and matches these feature within the overlapping region of adjacent images \citep{lowe_distinctive_2004}.
+The default implementation of the package works very well but has an $O(2n)$ space complexity and fails due to out-of-memory errors when stitching hundreds of images.
 %is this wording correct?, I'm trying to kindly say it was memory inefficient
@@
 Digitizing samples takes significantly longer for cookies than it does for cores (Figure \ref{fig:digitization_time}).
-This is due to the surface area of cookies requiring more images as their surface area has a square relationship to sample diameter.
+This is due to the surface area of cookies requiring more images as it has a square relationship to sample radius.
 For large cookies, the operator can choose to selectively scan a portion of the cookie. Scanning cookies at maximum resolution is unrealistic as the max cookie size is very small. 

### 2025-02-13 code/controller.py
Updating serial communication with GRBL and making some more methods direct injection

@@ -18,14 +18,16 @@ import numpy as np
 
 
 class Controller:
-    def __init__(self):
+    def __init__(self, g: gantry.Gantry, c: camera.Camera, f: focus.Focus):
         """Abstraction of the controller which moves the gantry, gets information from the GUI, operates the camera, and determines when to stitch.
+
+            g (gantry.Gantry): Gantry instance to interface with the GRBL controller.
+            c (camera.Camera): Camera instance to interface with the GStreamer pipeline and camera.
+            f (focus.Focus): Focus instance to capture in focus images.
         """
         #Settings for capturing images from multiple distances
-        # self.n_images = 11 #make sure you're not going faster than the frame rate of the GStreamer feed... 
         self.config = utils.load_config()
         self.n_images = self.config["controller"]["N_IMAGES_MULTIPLE_DISTANCES"] #make sure you're not going faster than the frame rate of the GStreamer feed... 
-        # self.height_range = 1
         self.height_range = self.config["controller"]["HEIGHT_RANGE_MM"]
         self.acceleration_buffer = self.config["controller"]["ACCELERATION_BUFFER_MM"]
         self.stitch_sizes = self.config["controller"]["STITCH_SIZES"]
@@ -36,9 +38,9 @@ class Controller:
     
         #Objects
         self.samples = []
-        self._gantry = gantry.Gantry()
-        self.camera = camera.Camera()
-        self.focus = focus.Focus(delete_flag=True, setpoint=math.floor(self.n_images / 2))
+        self._gantry = g
+        self.camera = c
+        self.focus = f
 
         #attributes
         self.image_height_mm = self.config["gui"]["DEFAULT_IMAGE_HEIGHT_MM"]
@@ -98,16 +100,12 @@ class Controller:
             self.focus.set_setpoint(round(self.n_images/2))
 
             #set directories
-            species = sample.species
-            id1 = sample.id1
-            id2 = sample.id2
-
             self.set_directory(sample.directory)
 
             start_time = time.time()
         
-            gantry_thread = Thread(target=self.capture_grid_photos, args=(sample.coordinates, sample.directory, sample.targets_top, sample.targets_bot, focus_queue, pid_queue, pid_lock, sample.rows, sample.cols, self.n_images, self.height_range, progress_callback, stop_capture, sample.is_core))
[…]

### 2025-02-13 code/focus.py

@@ -6,13 +6,14 @@ import os
 from pathlib import Path
 from PID import AsynchronousPID 
 import utils
+import sample
 
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
 class Focus:
 
     def __init__(self, delete_flag, setpoint):
-        self.config = utils.load_config()
+        self.config =  utils.load_config()
 
         self.DELETE_FLAG = delete_flag
         self.scale_factor = 0.05
@@ -24,11 +25,12 @@ class Focus:
         self.PID = AsynchronousPID(Kp=self.kp, Ki=self.ki, Kd=self.kd, setpoint=setpoint) 
         self.TESTINGLOG = []
 
-    def find_focus(self, focus_queue, pid_queue, pid_lock, directory, nvars: list, index:list, background:list, background_std:list):
+    def find_focus(self, sample: sample.Sample, focus_queue, pid_queue, pid_lock):
         while True:
             image_files = focus_queue.get()
             pid_lock.acquire()
-            log.info("Images For focus finding: {}".format(image_files))
+
+            #log.info("Images For focus finding: {}".format(image_files))
             if image_files == [-1]:
                 focus_queue.task_done()
                 break
@@ -43,12 +45,12 @@ class Focus:
                 self.delete_unfocused(image_files)
             else: 
                 Path("{}/focused_images".format(self.directory)).mkdir(exist_ok=True)
-                shutil.copy(focused_image_name, "{}/focused_images/{}".format(directory,filename))
+                shutil.copy(focused_image_name, "{}/focused_images/{}".format(sample.directory,filename))
             
-            background_std.append(std)
-            background.append(self.is_background(std))
-            index.append(stack_number)
-            nvars.append(nv)
+            sample.background_std.append(std)
+            sample.background.append(self.is_background(std))
+            sample.focus_index.append(stack_number)
+            sample.nvar.append(nv)
 
             if not self.is_background(std):
                 control_variable = self.PID.update(int(stack_number))
@@ -86,7 +88,7 @@ class Focus:
             
         std = np.std(vars)
 
-        log.info(image_filenames)
+        # log.info(image_filenames)
 
         if delete:
             self.delete_unfocused(image_filenames)

### 2025-02-13 code/gantry.py

@@ -8,9 +8,10 @@ import utils
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
 class Gantry:
+    """Abstraction of most of the GRBL protocols.
+    """
     def __init__(self, serial_port = "/dev/ttyUSB0", quiet=True):
         self.config = utils.load_config()
-
 
         self.quiet = quiet
         self._serial_port = serial_port # windows should be a "COM[X]" port which will vary per device
@@ -29,99 +30,186 @@ class Gantry:
         self._z = None
 
         self.position_lock = Lock()
+        self.send_command_lock = Lock() # adding lock for send command so that we can correctly associate GRBL responses to the corresponding command
         self.state = None
         
         self.thread = Thread(target=self.position_monitor)
 
     def position_monitor(self):
+        """Target function for thread to query GRBL state at approximately 5Hz.
+        """
         cmd = "?"
         # Set WPos status reports
         self._send_command("$10=2")
         while not self.stop_threads:
-            res_str = self._send_command(cmd)
+            grbl_out_list = self._send_command(cmd)
 
-            if res_str is None:
-                continue
-            elif res_str[-2:] == "ok":
-                continue
-            elif "WPos" in res_str:
-                self.position_lock.acquire()
-                try:
-                    _x, _y, _z = self.parse_coordinates(res_str)
-                    self._x = _x
-                    self._y = _y
-                    self._z = _z
-                except:
-                    log.info("Error reading parsing coordinates")
+            for grbl_out in grbl_out_list:
+                if "ok" in grbl_out:
                     continue
-                self.position_lock.release()
-                self.state = self.parse_state(res_str)
-                if not self.quiet:
-                    log.info("X {} \nY {}\nZ{}\n".format(_x, _y, _z))
-
+                elif "WPos" in grbl_out:
+                    self.position_lock.acquire()
+                    try:
+                        _x, _y, _z = self.parse_coordinates(grbl_out)
+                        self._x = _x
+                        self._y = _y
+                        self._z = _z
+                    except:
+                        log.error("Response with unparsable coordinates: {}".format(grbl_out))
+                    
+                    self.position_lock.release()
+                    self.state = self.parse_state(grbl_out)
+                    if not self.quiet:
[…]

### 2025-02-13 code/gui.py

@@ -8,6 +8,10 @@ import controller
 from datetime import datetime
 import time
 import utils
+import gantry
+import camera
+import focus
+import math
 
 log.basicConfig(format='%(process)d-%(levelname)s-%(message)s', level=log.INFO)
 
@@ -18,9 +22,10 @@ class App(Gtk.Window):
 
         self.set_default_size(self.config["gui"]["DEFAULT_WINDOW_SIZE"][0], self.config["gui"]["DEFAULT_WINDOW_SIZE"][1])
         self.set_size_request(self.config["gui"]["DEFAULT_WINDOW_SIZE"][0], self.config["gui"]["DEFAULT_WINDOW_SIZE"][1])
+        n_images = self.config["controller"]["N_IMAGES_MULTIPLE_DISTANCES"]
         self.connect("destroy", self.quit_program)
         
-        self.controller = controller.Controller()
+        self.controller = controller.Controller(gantry.Gantry(), camera.Camera(), focus.Focus(delete_flag=True, setpoint = math.floor(n_images / 2)))
 
         grid = Gtk.Grid()
         grid.set_row_homogeneous(False)
@@ -396,9 +401,9 @@ class App(Gtk.Window):
             overlap = 50
         else:
             overlap = float(overlap)
-            species = species.replace(" ", "_")
-            id1 = id1.replace(" ", "_")
-            id2 = id2.replace(" ", "_")
+            species = species.replace(" ", "_").replace("/","_").replace(".","_")
+            id1 = id1.replace(" ", "_").replace("/","_").replace(".","_")
+            id2 = id2.replace(" ", "_").replace("/","_").replace(".","_")
         self.controller.add_sample(width, height, overlap, species, id1, id2, notes, is_core)
         log.info("Adding Sample \nW: {}\nH: {}\nO: {}\nS:  {}\nID1:  {}\nID2:  {}\nNotes:  {}\n".format(width, height, overlap, species, id1, id2, notes))
     

### 2025-02-13 code/mosaic.py

@@ -13,7 +13,7 @@ import numpy as np
 
 from tile import Tile, OpenCVTile, MemmapOpenCVTile, MemmapTile
 
-from memory_profiler import profile
+#from memory_profiler import profile
 
 logger = logging.getLogger(__name__)
 

### 2025-02-15 code/camera.py
Adding correct DPI scaling, refactoring to have metadata be written as a method in Sample, tested on Tina. DPI gets added last so it shows up at the bottom of the metadata file. Might be annoying

@@ -19,6 +19,7 @@ class Camera:
         CROP_H = self.config["camera"]["CROP_H"]
         W_PIXELS = self.config["camera"]["W_PIXELS_NO_CROP"] - 2 * CROP_W
         H_PIXELS = self.config["camera"]["H_PIXELS_NO_CROP"] - 2 * CROP_H
+        MAX_DPI = self.config["camera"]["MAX_DPI"]
 
         Gst.init(None)
         # Create the pipeline with both display and save frame functionality

### 2025-02-15 code/controller.py

@@ -35,6 +35,7 @@ class Controller:
         self.slow_feed_rate_z = self.config["controller"]["SLOW_FEED_RATE_Z"]
         self.fast_feed_rate_xy = self.config["controller"]["FAST_FEED_RATE_XY"]
         self.fast_feed_rate_z = self.config["controller"]["FAST_FEED_RATE_Z"]
+        self.max_dpi = self.config["camera"]["MAX_DPI"]
     
         #Objects
         self.samples = []
@@ -103,7 +104,8 @@ class Controller:
             self.set_directory(sample.directory)
 
             start_time = time.time()
-        
+
+            sample.set_start_time_imaging(start_time)
             gantry_thread = Thread(target=self.capture_grid_photos, args=(sample, focus_queue, pid_queue, pid_lock, progress_callback, stop_capture))
             focus_thread = Thread(target=self.focus.find_focus, args=(sample, focus_queue, pid_queue, pid_lock))
             gantry_thread.start()
@@ -114,12 +116,9 @@ class Controller:
             focus_thread.join()
 
             end_time = time.time()
-            elapsed_time = end_time - start_time
-
-            self.create_metadata(sample, elapsed_time)
-            
-            return
-            #stop_capture.set()
+            sample.set_end_time_imaging(end_time)
+            sample.to_json()
+            break
 
     def capture_all_samples(self, progress_callback: Callable, stop_capture: Event):
         """Callable for the GUI to iterate through all samples. For multiple sample capture.
@@ -142,23 +141,25 @@ class Controller:
                 # Only stitch if the capture complete successfully
                 if not stop_capture.is_set():
                     print('stitching frames')
-                    self.stitch_frames(sample.directory)
+                    self.stitch_frames(sample)
                 if len(self.samples) == 0:
                     stop_capture.set()
 
             return
 
-    def stitch_frames(self, frame_dir: str):
+    def stitch_frames(self, sample: sample.Sample):
         """Stitch together frames into a mosaic. Not perfect yet as retrofitting Stitch2d is slightly cumbersome. If running into stitching issues, such as 'Killed' 
         or OOM, please try stitching using a more power computer with 16 GB of rame or something. This is a decently high priority to fix.
 
         Args:
-            frame_dir (str): Directory of where the frames are. 
+            sample (str): sample.Sample which is being stitched.. 
         """
+
         for size in self.stitch_sizes:
[…]

### 2025-02-15 code/sample.py

@@ -2,11 +2,12 @@ import cv2
 import time
 import numpy as np
 import math
+import utils
 from pathlib import Path
 from datetime import datetime
 
 class Sample:
-    def __init__(self, sample_width_mm: int, sample_height_mm: int, species:str, id1:str, id2:str, notes:str, image_width_mm:float, image_height_mm:float, is_core:bool,  percent_overlap:int = 50, is_vertical:bool = True, x:float = None, y:float = None, z:float = None):
+    def __init__(self, sample_width_mm: int, sample_height_mm: int, species:str, id1:str, id2:str, notes:str, image_width_mm:float, image_height_mm:float, image_width_pixels:float, image_height_pixels:float, is_core:bool, percent_overlap:int = 50, is_vertical:bool = True, x:float = None, y:float = None, z:float = None):
         """Class which contains the necessary information for each sample that needs to be digitized. Data comes from user inputs from GUI. Data is saved in metadata.json files alongside stitched images
 
         Args:
@@ -18,6 +19,8 @@ class Sample:
             notes (str): Additional notes for ambiguity in sample
             image_width_mm (float): Width in mm of the field of view of a single image. Helpful to print a 1mm x 1mm grid to measure
             image_height_mm (float): Height in mm of the field of view of a single image. Helpful to print a 1mm x 1mm grid to measure
+            image_width_pixels (float): Width in pixels of the field of view of a single image.
+            image_height_pixels (float): Height in mm of the field of view of a single image.
             is_core (bool): Is the sample a core? 
             percent_overlap (int, optional): Desired overlap between two images. Defaults to 50.
             is_vertical (bool, optional): If is_core, is the length of the core aligned vertically? Defaults to True.
@@ -27,6 +30,10 @@ class Sample:
         """
         self.width = sample_width_mm
         self.height = sample_height_mm
+        self.image_height_pixels = image_height_pixels
+        self.image_width_pixels = image_width_pixels
+        self.image_height_mm = image_height_mm
+        self.image_width_mm = image_width_mm
         self.percent_overlap = percent_overlap
         self._center = (round(x, 4),round(y, 4),round(z, 4))
         tl_x = x - (self.width/2)
@@ -54,6 +61,95 @@ class Sample:
         self.nvar = []
         self.is_core = is_core
         self.is_vertical = is_vertical
+        
+        # Adding rest of metadata after instantiation
+        self.start_time_imaging = None
+        self.end_time_imaging = None
[…]

### 2025-02-15 code/stitcher.py

@@ -11,6 +11,7 @@ import glob
 import gc
 import imutils
 import logging as log
+import sample
 # from memory_profiler import profile
 import sys 
 import time
@@ -27,12 +28,13 @@ class MaxFileSizeException(Exception):
         super().__init__(message)
 
 class Stitcher:
-    def __init__(self, frame_dir):
+    def __init__(self, sample: sample.Sample):
         config = utils.load_config()
+        self.sample = sample
         self._memmaps = []
         self._tiles = []
         self._mosaic = None
-        self._frame_dir = frame_dir
+        self._frame_dir = sample.directory
         self._metadata = None
         self._max_file_size = config["stitcher"]["MAX_FILE_SIZE_GB"] * 1000 # GBs
 
@@ -131,28 +133,6 @@ class Stitcher:
             self._metadata = j
             f.close()
 
-    def write_metadata(self, new_dir, shape, frame_shape, stitch_time):
-        new_metadata = {}
-        new_metadata["species"]=self._metadata["species"]
-        new_metadata["id1"]=self._metadata["id1"]
-        new_metadata["id2"]=self._metadata["id2"]
-        new_metadata["notes"]=self._metadata["notes"]
-        new_metadata["camera_pixels"]=self._metadata["camera_pixels"]
-        new_metadata["percent_overlap"]=self._metadata["percent_overlap"]
-        new_metadata["DPI"] = int((frame_shape[0] / self._metadata["image_height_mm"]) * 25.4) # fixing bug... cannot assume the overall mosaic height and sample height can be related perfectly
-        new_metadata["sample_height_mm"] = self._metadata["sample_height_mm"]
-        new_metadata["sample_width_mm"] = self._metadata["sample_width_mm"]
-        new_metadata["pixels_h"] = shape[0]
-        new_metadata["pixels_w"] = shape[1]
-        new_metadata["depth"] = shape[2]
-        new_metadata["stitch_time"] = stitch_time
-        new_metadata["frame_pixels_h"] = frame_shape[0]
-        new_metadata["frame_pixels_w"] = frame_shape[1]
-         
-
-        with open(os.path.join(new_dir, 'metadata.json'), 'w', encoding='utf-8') as f:
-            json.dump(new_metadata, f, ensure_ascii=False, indent=4)
-
     def memmap_to_tiff(self, memmap, path_tif):
         if memmap is not None:
             tifffile.imwrite(
@@ -179,7 +159,7 @@ class Stitcher:
     # @profile
[…]

### 2025-02-15 code/utils.py

@@ -9,13 +9,13 @@ def load_config(path = "."):
 
     return config["config"]
 
-def load_metadata(dir):
-    f = open(dir)
+def load_metadata(path):
+    f = open(path)
     metadata = json.load(f)
     f.close()
 
     return metadata
 
 def write_metadata(metadata, dir):
-    with open (dir, "w", encoding="utf-8") as f:
+    with open(os.path.join(dir, 'metadata.json'), 'w', encoding='utf-8') as f:
         json.dump(metadata, f, ensure_ascii=False, indent=4)

### 2025-02-24 code/gantry.py
Adding a state lock for thread safe state setting and getting

@@ -31,7 +31,8 @@ class Gantry:
 
         self.position_lock = Lock()
         self.send_command_lock = Lock() # adding lock for send command so that we can correctly associate GRBL responses to the corresponding command
-        self.state = None
+        self.state_lock = Lock()
+        self._state = None
         
         self.thread = Thread(target=self.position_monitor)
 
@@ -43,7 +44,7 @@ class Gantry:
         self._send_command("$10=2")
         while not self.stop_threads:
             grbl_out_list = self._send_command(cmd)
-
+        
             for grbl_out in grbl_out_list:
                 if "ok" in grbl_out:
                     continue
@@ -58,7 +59,11 @@ class Gantry:
                         log.error("Response with unparsable coordinates: {}".format(grbl_out))
                     
                     self.position_lock.release()
-                    self.state = self.parse_state(grbl_out)
+
+                    self.state_lock.acquire()
+                    self._state = self.parse_state(grbl_out)
+                    self.state_lock.release()
+
                     if not self.quiet:
                         log.info("X {} \nY {}\nZ{}\n".format(_x, _y, _z))
                 else:
@@ -107,37 +112,65 @@ class Gantry:
         Returns:
             str: The current GRBL machine state.
         """
-        if "ALARM" in grbl_status:
-            log.info("GRBL ALARM SOUNDED: {}".format(grbl_status))
-        elif "Idle" in grbl_status:
-            return "Idle"
-        elif "Jog" in grbl_status:
-            return "Jog"
-        elif "Alarm" in grbl_status:
-            return "Alarm"
-        elif "Home" in grbl_status:
-            return "Home"
-        elif "Run" in grbl_status:
-            return "Run"
-        elif "Hold" in grbl_status:
-            return "Hold"
-        elif "Door" in grbl_status:
-            return "Door"
-        elif "Check" in grbl_status:
-            return "Check"
-        elif "Sleep" in grbl_status:
-            return "Sleep"
+        acquired = self.state_lock.acquire(timeout = 3)
+        if acquired:
+            if "ALARM" in grbl_status:
+                log.info("GRBL ALARM SOUNDED: {}".format(grbl_status))
+                self._state = "Alarm"
+                self.state_lock.release()
+            elif "Idle" in grbl_status:
+                self._state = "Idle"
+                self.state_lock.release()
+            elif "Jog" in grbl_status:
+                self._state = "Jog"
+                self.state_lock.release()
+            elif "Alarm" in grbl_status:
+                self._state = "Alarm"
+                self.state_lock.release()
[…]

### 2025-02-24 code/controller.py
Removing PID lock because it was a redundant lock of the queue.get() blocking method. This fixed a deadlock issue

@@ -209,9 +209,8 @@ class Controller:
                         self._gantry.block_for_jog()
                         r = self.config["controller"]["CORE_CENTERING_RANGE"] #5
                         n_images_centering = self.config["controller"]["N_IMAGES_CORE_CENTERING"]
-                        filenames = self.capture_images_multiple_x(d, n_images_centering, self._gantry.feed_rate_z, r, self.acceleration_buffer)
+                        filenames = self.capture_images_multiple_x(sample.directory, n_images_centering, self._gantry.feed_rate_z, r, self.acceleration_buffer)
                         self.recenter_core_naive(filenames, r, self._gantry.feed_rate_z)
-
                 # Jog x and y and allow PID to handle the Z
                 elif img_num == len(sample.targets_top):
                     
@@ -220,15 +219,14 @@ class Controller:
                         self._gantry.jog_absolute_z(z)
                     else:
                         self._gantry.jog_absolute_xyz(x,y,z)
-
                 else:
                     if sample.is_core and sample.is_vertical:
                         self._gantry.jog_absolute_y(y)
                     else:
                         self._gantry.jog_absolute_xy(x, y)
-                    pid_lock.acquire()
+                    # pid_lock.acquire()
                     update_z = pid_queue.get()
-                    pid_lock.release()
+                    # pid_lock.release()
                     z += update_z
                     log.info(f"PID update Z by {update_z} mm")
                     if update_z != 0:

### 2025-02-24 docs/writing/paper.tex
Edits finally from me

@@
 % Title section
-\title{TIM: Tree Imaging Machine}
+\title{TIM: Tree Imaging Machine} %emw24Feb: Do we need a better title for the paper? LIke: TIM: A Tree Imaging Machine for low-cost digital images of tree cores and cookies ?
 \author{Your Name}
@@
 
+%emw24Feb: I added some text to abstract and intro that I am not 100% sure you would need for some journals, but I think you will need for MEE. 
+%emw24Feb: I was trying to add this tag '%emw24Feb:' to my comments, but got lazy and just added them below without the tag generally.
+Please delete my comments in the tex once you have dealt with them!
+
 \begin{abstract} 
-The Tree Imaging Machine (TIM) is a do-it-yourself scanning tool made to digitize tree cookies and cores.
+Studies of tree rings (dendrochronology) have benefitted recently from new methods to image wood rapidly and efficiently, helping to scale up the number and quality of image samples.
+Current open-source approaches are, however, limited in dimensionality and can be cost-prohibitive.
+The Tree Imaging Machine (TIM) is a do-it-yourself scanning tool made to digitize tree cookies and cores that address these limitations.
 TIM takes partially overlapping microscopic images of samples and stitches the individual images together to form
-a mosaic.
-These mosaics can be zoomed in to see distinct features on the scale of 0.01mm.
-With scans of up to 21,140 DPI, TIM produces one of the highest resolution images among similar tools for the competitive price of less than \$3,000 USD.
-We designed TIM to have a large working plane to allow the digitization of individual cores over 50cm and to allow the operator to set a batch of samples to be digitized without intervention. 
-All of the 3D printed parts are available for download on the NIH 3D printing repository and the software is 
-open source. Instructions to recreate this tool can be found on the code repository. 
+a mosaic, which can be zoomed in to visualize features on the scale of 0.01 mm.
+With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for the competitive price of less than \$3 000 USD.
+We designed TIM to have a large working plane to allow digitizing individual cores over 50 cm and to allow the operator to set a batch of samples to digitize without operator intervention.
+%emw24Feb: Could we say operator assistance?
+Or just: without any help from the operator?
+I also don't totally get what we mean here ...
+what are a batch of samples?
+Do we mean cores or multiple cookies?
+I might rephrase as: We designed TIM to have a large working plane to allow digitizing individual cores over 50 cm and batch sampling.
+With batch sampling ...
+All of the 3D printed parts are available for download on the NIH 3D printing repository (ADD LINK or reference?) and the software is 
+open source, with instructions to recreate this tool available at (ADD LINK).
+TIM provides an important advance in (resolution, cost, functionality, all-in-one package?) that advances tree ring sampling to tackle larger samples and collect data at finer resolutions in an open-source design that can be built on and improved as new technologies become available.
+% Add to the discussion something about next steps for TIM more clearly?
 \end{abstract}
@@
 \section{Introduction}
-The study of tree rings is useful across multiple fields, proving to be a reliable method for reconstructing past climates of regional and local environments, as well as a mechanism to understand tree growth response \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
-To obtain these data from the tree rings, it is necessary to measure tree rings widths from a tree cookie or core.
-The first high precision tool for this purpose was a stage micrometer, involving a trained technician to incrementally shift a tree core under the objective of a microscope - informing a computer when a new ring is encountered \citep{robinson_microcomputer_nodate}.
-While this method has very high precision, the data is only as accurate as the experience and knowledge of the technician at the time of recording \citep{levanic_atrics_2007}.
-The desire to remove repetition of errors in sampling and sampling bias across to individual technicians led researchers to an alternative - image analysis.
+%emw24Feb -- move around and add cites below to fit with current text.
+You should cite a Michael Mann paper, such as Proxy-based reconstructions of hemispheric and global surface temperature variations over the past two millennia
+Multiple fields---from climatology to forestry and ecology---rely on tree rings to understand how the environment affects tree growth. Tree rings have been critical to 
+reconstructing past climates of regional and local environments, informing our understanding of climate change today \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
+They are also increasingly used in ecology,  to understand how plant competition affects tree growth in addition to climate % cite https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2745.12782
+and with implications in forestry to managing stand dynamics.
+% maybe cite https://cdnsciencepub.com/doi/abs/10.1139/x03-232?casa_token=tzQCqy0zaGwAAAAA:5b-e-8-9OT4MZ_rxgxr4KrD7oXZ4KCcZi0S-e7ZaGCMB_gUfC8n_b4DcYuda5GT8DStxnWpA8BY7_A
+As tree growth and forest management becomes more critical to mitigating climate change, scaling up the collection and processing of tree ring samples has become increasingly important.
 
-The first step in measuring tree ring width from images requires the digitization of the sample from one of two major methods.  
-The original technique was as a flatbed scanner which can digitize the entire sample at once \citep{guay_new_1992}. 
-With a top of the line scanner, like the Epson Perfection V850 Pro, it's possible to scan at maximum resolutions of 4800 dpi and scan an area of up to 8.5" x 11.7" (21.6cm x 29.7cm).
-Analysis which relies on higher resolution or of larger samples require a different digitization approach.
+%One of many examples of removing PV 
+% You should ask Mao for the tree ring methods book ref and define these things quickly then cite it. 
+Data from the tree rings comes from measured tree rings widths from a tree cookie (DEFINE) or core (DEFINE).
+The first high precision tool for this purpose was a stage micrometer, which requires a trained technician to incrementally shift a tree core under the objective of a microscope---informing an attached computer when a new ring is encountered \citep{robinson_microcomputer_nodate}.
+% How does the 'informing' happen?
+I think you could make this more informative and use active voice ...
+do they click something or what?
+While this method has very high precision, the data are only as accurate as the experience and knowledge of the technician at the time of recording as no permanent image record is produced \citep{levanic_atrics_2007}.
 
-The second digitization model was introduced with ATRICS \citep{levanic_atrics_2007}. 
-Rather than scanning a whole sample at once, a high resolution camera takes multiple images across the surface of the sample and uses image stitching software techniques to combine them into one ultra high resolution image \citep{muhlich_stitching_2022}.
+Image analysis of tree rings was designed to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in (late 1980s to 1990?
+And add cites) to digitize tree ring samples using flatbed scanners.
+Early techniques used a flatbed scanner to digitize the entire sample at once \citep{guay_new_1992}, producing maximum resolutions of 4800 dpi and scan an area of up to  21.6 cm x 29.7 cm (e.g., using a top of the line scanner, like the Epson Perfection V850 Pro) % I think we can skip inches ....
+8.5 inches x 11.7 inches
+More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach, called ATRICS \citep{levanic_atrics_2007}.
+% Should we spell out ATRICS?
+Also I think we should introduce the METHOD, then say ATRICS is an example of it (for example, 'through a different digitization approach, which takes multiple images then stitches them together.
+One widely used version of this was ATRICS ...'), but I am not sure if that is true so will take a look on next version; feel free to add info here in the comments.
+Rather than scanning a whole sample at once, with ATRICS a high resolution camera takes multiple images across the surface of the sample then uses image stitching software to combine them into one ultra high resolution image \citep{muhlich_stitching_2022}.
 Stitching multiple images into one mosaic has been applied in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}. 
 This method requires either the camera objective to move relative to the sample, or the sample move underneath a stationary camera.
+
+% I would start a new paragraph here that opens with something like: Since XX was introduced in XX, image capture and stitching of tree rings has increased.
+Today CaptuRING offers a  more modern do-it-yourself alternative to ATRICS that is (open source?
+Has resolutions up to XX?
+Used XX type of camera?
+Give a little more here).
+Another approach, Gigapixel, moves the camera relative to the sample, allowing for multiple samples to be recorded in sequence \citep{griffin_gigapixel_2021}.
+% Then I would end on the cookies issue and also anything else to mention, like cost? You should also clearly mention when/how stitching happens in these approaches.
 For ATRICS and a more modern do-it-yourself alternative, CaptuRING, the sample is moved relative to the camera \citep{garcia-hidalgo_capturing_2022}. 
@@
 While these machines can all digitize cores, none have been shown to digitize cookies.
+% I think you could build up to this point by adding in that the above approaches move in one linear direction or such?
 
+% I will edit this a little next time, but for now fix up em-dashes, and simplify with less PV as possible. 
 TIM was made to combine the defining features of the previously mentioned machines into one while making the project open-source and open-hardware. 
@@
 The only specialized piece of equipment needed to build TIM is a 3D printer, but the parts can be readily ordered through 3D print shops if preferred.
-Excluding 3D printed parts, the total cost of the machine is approximately \$2,200 USD compared to the \$70,000 USD of the Gigapixel \citep{griffin_gigapixel_2021}.
+Excluding 3D printed parts, the total cost of the machine is approximately \$2 200 USD compared to the \$70 000 USD of the Gigapixel \citep{griffin_gigapixel_2021}.
 The total cost of the machine is almost comparable in price or less to many professional camera and macro lens combinations.
@@
 \subsection{System Overview} % this is so dry fix later
+% Can we have an a photo as a figure that we link to here?
 
+% Good topic sentence!
+Can you walk reader more carefully through each item?
+You may need to re-order them ..
+But basically, pick the most useful order and then write: For the camera ....
+For the gantry machine ...
+You might also use this space to say, in contrast to previous approaches that have used XX giant camera, we used ...
+and otherwise point out the novelty when it comes up.
 TIM can be thought of as a combination of multiple subsystems: the camera, computer, and gantry machine. 
@@
 On the linear actuator's build plate, adapters were made to hold a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens connected to an NVIDIA Jetson Orin Nano (Jetson) through a CSI cable.
-By choosing this combination, we were able to reduce the weight and cost of the camera significantly - trickling the budget into an efficient computer which can handle intensive image processing. 
+% Nice active voice below!
+By choosing this combination, we were able to reduce the weight and cost of the camera significantly, which allowed us to invest more in an efficient computer that can handle intensive image processing.
 The Jetson is a powerful edge computer which drives a monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for automatic control, and performs calculations to stitch individual images into one mosaic.
@@
 \subsection{Preparing Cookies and Cores}
-TIM can be used to scan both tree cookies and tree cores. Sample preparation for both are equally important but require different approaches.
-The scanning surface of the cookies and cores need to be sanded with incrementally increasing sand paper grit. Our cookies were sanded using an orbital sander with mesh sandpaper from 60 to 800 grit.
-We used a microscope to visually inspect the quality of the sanding and when vessels were easily identifiable, we considered the sample to be sufficiently prepared.
+% Tried to reduce PV below. 
+TIM can scan both tree cookies and tree cores. Sample preparation for both are equally important but require different approaches.
+Similar to general sample preparation methods for tree rings samples (CITE dendro manual), cookies and cores need to be sanded with incrementally increasing sand paper grit to produce an even surface for scanning.
+We sanded cookies using an orbital sander with mesh sandpaper from 60 to 800 grit, then used a microscope to visually inspect the quality of the sanding.
+We considered the sample to be sufficiently prepared when vessels were easily identifiable.
 
@@
 \subsection{Sample Digitization} % Methods
-The subsystems can be best understood by following the process from sample setup through through obtaining a stitched image. 
+% It would be super nice to have a concept image of this process! 
+The subsystems of TIM can be best understood by following the process from sample setup through through obtaining a stitched image. 
 To achieve this with a fixed focus camera, the samples must be nearly orthogonal to the camera lens. 
@@
 
-\subsection{Image Capturing}
+\subsection{Image Capturing} % Could this be Image Capture? 
 In the GUI the operator indicates the height and width of the field of view in an individual image from the camera. 
@@
 
+% I am not sure what 'implement automatic control to focus the images' ... I started an improved sentence below, but stalled because I am not sure what you mean.... 
+% Centering the design of TIM around a fixed focus camera and lens means we need to automatically focus the images.
 By centering the design around a fixed focus camera and lens, it is necessary to implement automatic control to focus the images. 
@@
 And with the microscope lens, the depth of field of the image is so sensitive that sub millimeter heights can move an entire image out of focus.
-Solving this in a time efficient manner involves two stages. 
+Solving this in a time-efficient manner involves two stages. 
 The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
-This allows the initial $Z_0$ value ,captured when adding a sample, to be an informed guess as to what value would have an in focus image across the entire surface. 
+This allows the initial $Z_0$ value, captured when adding a sample, to be an informed guess as to what value would have an in focus image across the entire surface. 
 From the first $(X_0, Y_0)$ coordinate captured, 11 images are taken at different $Z$ values in $\boldsymbol{Z_{\text{samples}}}$. 
-So long as $Z_{focused}$ is within this set, an in focus image exists.
+So long as $Z_{focused}$ is within this set, an in focus image exists. % why 11?
 
+% Something bad is happening with the i below. 
 \[
@@
 Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to improve speed and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
-This is a variation of the standard approach used in 3D printing and CNC control methods.
+This is a variation of the standard approach used in 3D printing and CNC control methods. % What is CNC?
 
@@
 These inaccuracies can cause drift in the images with relation to the core and result in the width of the core not being within the field of view of the images.
-Specifically for cores, a centering procedure was implemented to realign the center of the core to the center of the image frame when the machine moves 
+We thus implemented a centering procedure that realigns the center of the core to the center of the image frame when the machine moves 
 to a new sample.
 
-First, the camera is moved to to the $(X_0, Y_0, Z_0)$ of the new core. 
+First, the camera is moved to the $(X_0, Y_0, Z_0)$ of the new core. 
 From there, the camera is moved across a window, at constant velocity, from $(X_0 - 1/2*{range}, Y_0, Z_0)$ to $(X_0 + 1/2*{range}, Y_0, Z_0)$. 
@@
 
+% What do we mean below?
+I think we might need to use more words (and sentences) to spell this out for readers: Some tree ring projects/analuses may not need high resolution images and could be needlessly slowed by TIM's approach ...
 Depending on the analysis, it is unnecessary to digitize samples at a very high DPI. To address this, a parameter in the machine configuration file 
@@
 
-TIM has been used to digitize 90 cookies and is in the process of finishing 1,200 cores. 
+TIM has been used to digitize 90 cookies and is in the process of finishing 1,200 cores. % I would move this to the end of the results and/or just put it in the cover letter to the journal.
+
 Digitizing samples takes significantly longer for cookies than it does for cores (Figure \ref{fig:digitization_time}).
@@
 The range of sample included are from a 3mm x 220mm core to a 75mm x 56mm cookie.
-True sampling times are drastically influenced by configurable machine parameters.}
+True sampling times are drastically influenced by configurable machine parameters.} % Take the title on top off this image as we don't have titles on images in sci papers.
     \label{fig:digitization_time}
@@
 
-\section{Discussion}
-By taking advantage of a common 3 degree of freedom cartesian machine design, powerful edge computing, and a microscope camera we were able to design a cost effective and high resolution digitization tool for wood samples.
+\section{Discussion} % I think we could remind the reader here of what gaps TIM addresses, but I can help do that next time. 
+By taking advantage of a common three degree of freedom cartesian machine design, powerful edge computing, and a microscope camera we were able to design a cost effective and high resolution digitization tool for wood samples.
 We significantly increased the maximum sample length and maximum resolution compared to common alternatives in the field. 
@@
 \subsection{Strengths and Opportunities}
+% Try to reduce PV here please!
 Our design has tried to minimize the barriers to building this machine in smaller labs. After investing a few hours in the sourcing and assembly of parts, 
@@
 Further testing with vessel counting models or ring identification models would be an interesting extension to take advantage of high resolution scans.
+% I guess Sandy never sent in results we could use?
 

### 2025-02-25 docs/writing/paper.tex
Making changes per Lizzie's edits.

@@
 % Title section
-\title{TIM: Tree Imaging Machine} %emw24Feb: Do we need a better title for the paper? LIke: TIM: A Tree Imaging Machine for low-cost digital images of tree cores and cookies ?
+\title{TIM: Tree Imaging Machine for low-cost digital images of tree cores and cookies} 
 \author{Your Name}
@@
 With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for the competitive price of less than \$3 000 USD.
-We designed TIM to have a large working plane to allow digitizing individual cores over 50 cm and to allow the operator to set a batch of samples to digitize without operator intervention.
-%emw24Feb: Could we say operator assistance?
-Or just: without any help from the operator?
-I also don't totally get what we mean here ...
-what are a batch of samples?
-Do we mean cores or multiple cookies?
-I might rephrase as: We designed TIM to have a large working plane to allow digitizing individual cores over 50 cm and batch sampling.
-With batch sampling ...
-All of the 3D printed parts are available for download on the NIH 3D printing repository (ADD LINK or reference?) and the software is 
-open source, with instructions to recreate this tool available at (ADD LINK).
-TIM provides an important advance in (resolution, cost, functionality, all-in-one package?) that advances tree ring sampling to tackle larger samples and collect data at finer resolutions in an open-source design that can be built on and improved as new technologies become available.
+We designed TIM to have a large working plane to allow digitizing cores over 50 cm and complete cookies.
+Operators can prepare sample batches of multiple cores and cookies, letting the machine work until the queue is finished.
+All of the 3D printed parts are available for download on the [NIH 3D printing repository](https://doi.org/10.60705/3DPX/21561.3) and the software is 
+open source, with instructions to recreate this tool available on [GitHub](https://github.com/temporalecologylab/TreeRings).
+TIM provides an important advance in scanning resolution, tree cookie sampling, maximum core sample size, batch digitization, and cost in a single package.
+This promotes the digitization of larger samples and can collect data at finer resolutions in an open-source design which can be built on and improved as new technologies become available.
 % Add to the discussion something about next steps for TIM more clearly?
@@
 reconstructing past climates of regional and local environments, informing our understanding of climate change today \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
-They are also increasingly used in ecology,  to understand how plant competition affects tree growth in addition to climate % cite https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2745.12782
-and with implications in forestry to managing stand dynamics.
-% maybe cite https://cdnsciencepub.com/doi/abs/10.1139/x03-232?casa_token=tzQCqy0zaGwAAAAA:5b-e-8-9OT4MZ_rxgxr4KrD7oXZ4KCcZi0S-e7ZaGCMB_gUfC8n_b4DcYuda5GT8DStxnWpA8BY7_A
+They are also increasingly used in ecology, to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017}.
+and with implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}. 
+% Add Michael Mann citation
 As tree growth and forest management becomes more critical to mitigating climate change, scaling up the collection and processing of tree ring samples has become increasingly important.
@@
 % You should ask Mao for the tree ring methods book ref and define these things quickly then cite it. 
-Data from the tree rings comes from measured tree rings widths from a tree cookie (DEFINE) or core (DEFINE).
-The first high precision tool for this purpose was a stage micrometer, which requires a trained technician to incrementally shift a tree core under the objective of a microscope---informing an attached computer when a new ring is encountered \citep{robinson_microcomputer_nodate}.
-% How does the 'informing' happen?
-I think you could make this more informative and use active voice ...
-do they click something or what?
-While this method has very high precision, the data are only as accurate as the experience and knowledge of the technician at the time of recording as no permanent image record is produced \citep{levanic_atrics_2007}.
+Measuring the width between tree rings from tree cookies (DEFINE) or tree cores (DEFINE) is where the analysis begins.
+% reword, placeholder for now % OLD: Data from the tree rings comes from measured tree rings widths from a tree cookie (DEFINE) or core (DEFINE).
+The first method to measure tree ring widths used a microcomputer, a stage micrometer, and a push button.
+A trained technician shifted a core underneath a microscope objective and pushed a button connected to a microcomputer to record a ring's location \citep{robinson_microcomputer_1980}.
+This method was high precision but depended heavily on specific technicians and maintained no image record \citep{levanic_atrics_2007}.
 
-Image analysis of tree rings was designed to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in (late 1980s to 1990?
-And add cites) to digitize tree ring samples using flatbed scanners.
-Early techniques used a flatbed scanner to digitize the entire sample at once \citep{guay_new_1992}, producing maximum resolutions of 4800 dpi and scan an area of up to  21.6 cm x 29.7 cm (e.g., using a top of the line scanner, like the Epson Perfection V850 Pro) % I think we can skip inches ....
-8.5 inches x 11.7 inches
-More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach, called ATRICS \citep{levanic_atrics_2007}.
-% Should we spell out ATRICS?
-Also I think we should introduce the METHOD, then say ATRICS is an example of it (for example, 'through a different digitization approach, which takes multiple images then stitches them together.
-One widely used version of this was ATRICS ...'), but I am not sure if that is true so will take a look on next version; feel free to add info here in the comments.
-Rather than scanning a whole sample at once, with ATRICS a high resolution camera takes multiple images across the surface of the sample then uses image stitching software to combine them into one ultra high resolution image \citep{muhlich_stitching_2022}.
-Stitching multiple images into one mosaic has been applied in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}. 
-This method requires either the camera objective to move relative to the sample, or the sample move underneath a stationary camera.
+Image analysis of tree rings was designed to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in the 1980s and 1990s to digitize tree ring samples using optical scanning \citep{mcmillin_application_1982} \citep{guay_new_1992}.
+Optical scanners are still readily available for purchase, using a top of the line scanner, like the Epson Perfection V850 Pro produces maximum resolutions of 6 400 dpi and scan an area of up to 21.6 cm x 29.7 cm.
+More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach.
+Samples are digitized using an image-capturing system built using motors to move the sample, a microscope camera to take images, a computer to manage the capturing, and a second, more powerful, computer to stitch images into a larger composite image.
+Decreasing the field of view of each captured image increases the resolution of the final stitched image without restricting the size of the sample \citep{muhlich_stitching_2022}.
+Stitching multiple images into one mosaic is a common technique used in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}. 
+The first implementation of the capture and stitch method was the Advanced Tree-Ring Image Capturing System (ATRICS) \citep{levanic_atrics_2007}.
+% Adding ATRICS after introducing the image and stitch method
 
@@
 % Then I would end on the cookies issue and also anything else to mention, like cost? You should also clearly mention when/how stitching happens in these approaches.
-For ATRICS and a more modern do-it-yourself alternative, CaptuRING, the sample is moved relative to the camera \citep{garcia-hidalgo_capturing_2022}. 
-Gigapixel takes a different approach by moving the camera relative to the sample, allowing for multiple samples to be recorded in sequence \citep{griffin_gigapixel_2021}. 
-While these machines can all digitize cores, none have been shown to digitize cookies.
-% I think you could build up to this point by adding in that the above approaches move in one linear direction or such?
+Since ATRICS was introduced in 2007, image capture and stitching of tree rings has become more common.
+Today CaptuRING offers a more modern do-it-yourself alternative to ATRICS, creating scans up to 5 339 dpi using a DSLR camera and motorized stage \citep{garcia-hidalgo_capturing_2022}.
+Another alternative, Gigapixel, achieves an impressive 19 812 dpi with the same DSLR image capture and composite stitch method \citep{griffin_gigapixel_2021}. 
+While CaptuRING and Gigapixel both use DSLR cameras and motorized movement, they differ on two fronts. 
+First, Gigapixel's camera is traversed across a stationary sample while CaptuRING moves a core sample underneath a stationary camera.
+With a mobile camera, multiple samples can be recorded sequentially without needing to remove a sample from a stage upon image completion---significantly reducing the active setup time for an operator.
+Second, Gigapixel automatically stitches individual images together once imaging is complete.
+This removes the need to transfer images to a more powerful computer for stitching as is necessary with CaptuRING.
+Additionally, both of these systems are designed to capture images along a singular axis as the entire width of a core fits within a single image's field of view.
+This assumption fundamentally restricts the digitization to be limited only to cores, as cookies have a two dimensional scanning surface.
 
 % I will edit this a little next time, but for now fix up em-dashes, and simplify with less PV as possible. 
-TIM was made to combine the defining features of the previously mentioned machines into one while making the project open-source and open-hardware. 
-We designed TIM to digitize both cookies and cores, increase the maximum sample length, perform image stitching without operator intervention, and capture batches of samples in a queue - all while minimizing cost.
-The only specialized piece of equipment needed to build TIM is a 3D printer, but the parts can be readily ordered through 3D print shops if preferred.
+We made TIM to combine the open-source and open-hardware philosophies of CaptuRING with the functionality of Gigapixel.
+We added support to digitize both cookies and cores, increased the maximum sample length, perform image stitching directly without the need for manual file transfer, and capture batches of samples sequentially---all while minimizing cost.
+Our design requires the use of 3D printed components and common hand tools such as allen keys and wrenches. No specialized equipment or power tools are required beyond this.
 Excluding 3D printed parts, the total cost of the machine is approximately \$2 200 USD compared to the \$70 000 USD of the Gigapixel \citep{griffin_gigapixel_2021}.
-The total cost of the machine is almost comparable in price or less to many professional camera and macro lens combinations.
-Additional savings can also be had when factoring in the cost of a professional stitching software license such as PTGui.
+We chose to use a Raspberry Pi HQ camera with a microscope lens to save on purchasing a professional camera with a macro lens.
+Professional camera setups are comparable in price to the entire cost of TIM.
+Additional savings can also be had when factoring in the cost of a professional stitching software license such as PTGui, the suggested software to use with CaptuRING.
 
@@
 
-\subsection{System Overview} % this is so dry fix later
-% Can we have an a photo as a figure that we link to here?
+\subsection{TIM System Design} % this is so dry fix later. The engineer in me loves a dry title
+
+\begin{figure}
+  \centering
+  \includegraphics[height=0.5\linewidth]{../content/tina.jpg}
+  \caption{Complete assembly of TIM. Included in the work plane a few cookies ready to be scanned.}
+  \label{fig:tim_assembled}
+\end{figure}
 
@@
 and otherwise point out the novelty when it comes up.
-TIM can be thought of as a combination of multiple subsystems: the camera, computer, and gantry machine. 
-Cartesian movement in the X, Y and Z directions is a result of two machine kits and a motor controller from OpenBuilds - the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
+TIM can be thought of as a combination of multiple subsystems: a gantry machine, camera, and computer (See Figure \ref{fig:tim_assembled}). 
+For the gantry machine, cartesian movement in the $X$, $Y$ and $Z$ directions is a result of two machine kits and a motor controller from OpenBuilds---the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
 Smaller and larger working planes easily be achieved by purchasing a different sized ACRO kit.
@@
 To fit on the build plates of the ACRO system, we made an adapter to connect the linear actuator to the X and Y axis, thus creating motion in the Z axis.
-On the linear actuator's build plate, adapters were made to hold a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens connected to an NVIDIA Jetson Orin Nano (Jetson) through a CSI cable.
+For the camera, we chose a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens and connected this to the gantry machine using custom adapters.
+For the computer, we implemented all of the software control and image stitching to run on an NVIDIA Jetson Orin Nano (Jetson) which also connects to the camera via a CSI cable.
+
 % Nice active voice below!
-By choosing this combination, we were able to reduce the weight and cost of the camera significantly, which allowed us to invest more in an efficient computer that can handle intensive image processing.
-The Jetson is a powerful edge computer which drives a monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for automatic control, and performs calculations to stitch individual images into one mosaic.
+By choosing this combination, we were able to reduce the weight and cost of the camera significantly, which allowed us to invest more in a powerful computer that can handle intensive image processing.
+The Jetson is an edge computer which drives a monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for control, and performs calculations to stitch individual images into one mosaic.
 Despite the weight reductions, torsion on the gantry arm still resulted in a non-zero torsional deflection. A torsion correcting adapter was designed to counteract this rotation and level the lens.
@@
 Similarly to cookies, care should be given to achieving a scanning surface as parallel to the XY-plane as possible.
-Cores also need to be aligned to be parallel to the Y axis to be digitized, using squaring tool or similar to help with alignment.
+Cores also need to be aligned to be parallel to the $Y$ axis to be digitized, using squaring tool or similar to help with alignment.
 Any warping in core mounts need to be counteracted with the use of trigger clamps.
@@
 \subsection{Sample Digitization} % Methods
-% It would be super nice to have a concept image of this process! 
+
 The subsystems of TIM can be best understood by following the process from sample setup through through obtaining a stitched image. 
@@
 
-\subsection{Image Capturing} % Could this be Image Capture? 
+\subsection{Image Capture}  
+% It would be super nice to have a concept image of this process! A - I've struggled a bit with how to convey this concisely, let me know what you think about the new figure
+\begin{figure}
+  \centering
+  \includegraphics[height=0.4\linewidth]{../content/cookie_traverse.png}
+  \caption{Scanning a sample begins with the machine navigating to the center of the sample which was automatically recorded at sample add time.
+From there, a grid of target locations is calculated for the machine to capture an image at.
+The system systematically navigates to each location and captures an image until all target locations have been visited.
+Core samples are digitized in the same way but only capture images in one dimension.}
+  \label{fig:cookie_traverse}
+\end{figure}
+
 In the GUI the operator indicates the height and width of the field of view in an individual image from the camera. 
@@
 Once prompted the system begins to exhaustively traverse the surface area of the sample, using the sample height and width entered by the operator.  
-The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors - the basis of image stitching. 
+The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors---the basis of image stitching (See Figure \ref{fig:cookie_traverse}). 
 Digitizing cores can be done without needing to traverse in both the X and Y axis. 
-The optimal scanning path for a core is to ignore the Y axis and traverse the cores length.
+The optimal scanning path for a core is to align the length of cores to the Y axis and only take one column of images.
 Spanning the surface area of a cookie requires movement in both the X and Y axis.
 
-% I am not sure what 'implement automatic control to focus the images' ... I started an improved sentence below, but stalled because I am not sure what you mean.... 
-% Centering the design of TIM around a fixed focus camera and lens means we need to automatically focus the images.
-By centering the design around a fixed focus camera and lens, it is necessary to implement automatic control to focus the images. 
-The only way to control the focus of a fixed focus lens is by changing the lens's distance to the sample. 
-And with the microscope lens, the depth of field of the image is so sensitive that sub millimeter heights can move an entire image out of focus.
+Centering the design of TIM around a fixed focus camera and lens means we need to focus the images with a software procedure.
+The only controllable parameter to focus a fixed focus lens is by changing the lens's distance to the sample. 
+And with the microscope lens, the depth of field of the image is so sensitive that sub millimeter heights can move an image out of focus.
 Solving this in a time-efficient manner involves two stages. 
@@
 This allows the initial $Z_0$ value, captured when adding a sample, to be an informed guess as to what value would have an in focus image across the entire surface. 
-From the first $(X_0, Y_0)$ coordinate captured, 11 images are taken at different $Z$ values in $\boldsymbol{Z_{\text{samples}}}$. 
-So long as $Z_{focused}$ is within this set, an in focus image exists. % why 11?
+From the first $(X_0, Y_0)$ coordinate captured, multiple images are taken at different $Z$ values in $\boldsymbol{Z_{\text{samples}}}$. 
+We found that taking 11 images at different heights provides a sufficient set of in and out-of-focus images for comparison.
+The number of images taken is an editable parameter in the configuration file and can be increased, along with the height range, if in focus images are not being obtained.
 
-% Something bad is happening with the i below. 
+% Something bad is happening with the i below. <-- is this comment still relevant? I don't see i
 \[
@@
 \end{bmatrix}
-% \exists 
-% Z_{focused}
 \] % unsure if this makes sense
@@
 
-This focusing procedure works well alone when the sample alignment has a difference in height no greater than 0.5 mm, $d_1 - d_2 < 0.5 mm $ (See Figure \ref{fig:realistic_levelling}). 
-% And while 1mm may sound like a small range, it is important to note that increases in time at each $(X_k, Y_k)$ are repeated $k$ times. 
+The second component of a time-efficient focusing procedure is using a feedback loop to update the initial $Z_0$ at each image location. 
+Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm, $d_1 - d_2 < 0.5 mm $ (See Figure \ref{fig:realistic_levelling}). 
 Rather than adjusting this range, a greater alignment error can be managed by controlling the center of the range, $Z_{0,k}$, for $(X_k,Y_k)$. 
@@
 A PID control algorithm with a process variable of $i_{max}$ and control variable of $Z_{0,k}$ allows a negative feedback loop to improve focusing across the entire sample \citep{odwyer_summary_2000}.
-Instead of $(Z_{max} - Z_{min}) < 0.5mm$, now the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which allows for a more forgiving sample setup.
-After the capturing overlapping images across the entire surface area, the images are ready to be stitched.
+Instead of $(d_1 - d_2) < 0.5mm$ (See \ref{fig:realistic_levelling}), the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which allows for a more forgiving sample setup.
+%After the capturing overlapping images across the entire surface area, the images are ready to be stitched.
 
@@
 Difficulties arise when more than one core are added to the digitization queue.
-The ACRO has a theoretical 4.5 micron translational accuracy, but is hard to achieve.
+The ACRO has a theoretical 4.5 micron translational accuracy, but has not been achieved despite significant effort.
 These inaccuracies can cause drift in the images with relation to the core and result in the width of the core not being within the field of view of the images.
@@
 This package wraps OpenCV functions for finding distinctive image features using the SIFT feature detector and matches these feature within the overlapping region of adjacent images \citep{lowe_distinctive_2004}.
-The default implementation of the package works very well but has an $O(2n)$ space complexity and fails due to out-of-memory errors when stitching hundreds of images.
-%is this wording correct?, I'm trying to kindly say it was memory inefficient
-With a few key memory conscious changes to the algorithm, the space complexity was reduced to approximately $O(\log{n})$ and the package was able to run on the Jetson without a problem.
+The default implementation of the package produces high quality stitches but is memory inefficient. %is this wording correct?, I'm trying to kindly say it was memory inefficient
+We implemented a few key memory conscious changes to the package and were able to run it on the Jetson without a problem.
 
-% What do we mean below?
-I think we might need to use more words (and sentences) to spell this out for readers: Some tree ring projects/analuses may not need high resolution images and could be needlessly slowed by TIM's approach ...
-Depending on the analysis, it is unnecessary to digitize samples at a very high DPI. To address this, a parameter in the machine configuration file 
-allows the user to downsize images before they get stitched together to avoid unnecessary sample detail and decrease the size of the files.
+Some tree ring analyses may not need the maximum scanning resolution produced by TIM. At maximum resolution, TIM's scans quickly produce large file sizes and can be costly to store at scale. 
+To address this, a parameter in the machine configuration file allows the operator to choose a downsizing ratio to apply to images before they get stitched, decreasing the file size of the scans.
 With the same images, multiple final stitched resolutions can be made.
@@
 
-TIM has been used to digitize 90 cookies and is in the process of finishing 1,200 cores. % I would move this to the end of the results and/or just put it in the cover letter to the journal.
-
 Digitizing samples takes significantly longer for cookies than it does for cores (Figure \ref{fig:digitization_time}).
 This is due to the surface area of cookies requiring more images as it has a square relationship to sample radius.
-For large cookies, the operator can choose to selectively scan a portion of the cookie. Scanning cookies at maximum resolution is unrealistic as the max cookie size is very small. 
+For large cookies, the operator can choose to selectively scan a portion of the cookie instead of the whole surface.
+Scanning complete cookies at maximum resolution is unrealistic as the max cookie size is very small.
 To fit into the maximum 2.5 GB limits of a TIFF file, the maximum diameter of a cookie scanned at 30\% of maximum resolution is limited to approximately 130 mm.
-For cores at maximum resolution, this file size limit
+This is significantly smaller when compared to the maximum theoretical core sample size.
+At maximum resolution, the file size limit
 constrains samples to approximately 1,500 mm in length.
@@
 Samples that are larger than this file size limit are still possible to be scanned but they are no longer stored as a TIFF file.
-An uncompressed binary NumPy memory-mapped array file is produced.
-This implies that the operator knows how to splice into these files and is comfortable working with NumPy memory-mapped arrays in Python.
+An uncompressed binary NumPy memory-mapped array file is produced. These NumPy files are not supported by standard image viewers like ImageJ and would require custom tools to work with.
 
@@
 \subsection{Strengths and Opportunities}
-% Try to reduce PV here please!
-Our design has tried to minimize the barriers to building this machine in smaller labs. After investing a few hours in the sourcing and assembly of parts, 
-the rest of the implementation has been completed. Digitized samples from this machine have been able to have tree rings registered onto their images using CooRecorder.
-Integration into a cloud data storage system would be a great addition to remove the need for an operator to move the final data with a physical external drive. 
-Further testing with vessel counting models or ring identification models would be an interesting extension to take advantage of high resolution scans.
+% add better opportunities
+Our design minimizes the barriers to build a TIM in smaller labs without compromising scan quality.
+We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ.
+Moving forward, TIM could benefit from being integrated into a cloud data storage system with automatic uploads---removing the need to manually transfer completed scans with physical drives.
+Experimenting with sensors to decrease image focusing time would be a way to addition to the system.
+Additionally, developing a framework to run vessel counting or ring identification models while scanning could greatly decrease the time invested in manually identifying tree anatomy.
 % I guess Sandy never sent in results we could use?
+A- Sadly not.
 

### 2025-03-03 docs/writing/paper.tex
Some more changes

@@
 \section{Introduction}
-%emw24Feb -- move around and add cites below to fit with current text.
-You should cite a Michael Mann paper, such as Proxy-based reconstructions of hemispheric and global surface temperature variations over the past two millennia
-Multiple fields---from climatology to forestry and ecology---rely on tree rings to understand how the environment affects tree growth. Tree rings have been critical to 
+Incremental growth rings from trees provide a valuable proxy record of the environment and climate.
+They have proven to provide the highest accuracy, precision, and reliable dating among alternatives \citep{mann_northern_1999}.
+Multiple fields---from climatology to forestry and ecology---rely on tree rings to understand how the environment affects tree growth.
+Tree rings have been critical to
 reconstructing past climates of regional and local environments, informing our understanding of climate change today \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
-They are also increasingly used in ecology, to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017}.
-and with implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}. 
-% Add Michael Mann citation
+They are also increasingly used in ecology, to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017} and have implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}.
 As tree growth and forest management becomes more critical to mitigating climate change, scaling up the collection and processing of tree ring samples has become increasingly important.
 
-%One of many examples of removing PV 
-% You should ask Mao for the tree ring methods book ref and define these things quickly then cite it. 
-Measuring the width between tree rings from tree cookies (DEFINE) or tree cores (DEFINE) is where the analysis begins.
-% reword, placeholder for now % OLD: Data from the tree rings comes from measured tree rings widths from a tree cookie (DEFINE) or core (DEFINE).
-The first method to measure tree ring widths used a microcomputer, a stage micrometer, and a push button.
+% Can't find a proper citation with an explicit tree core/cookie definition. Will try harder later.
+Two types of tree sampling techniques are used to capture the variation of tree rings. The first sample type is collected using an incremental borer, used to remove a cylindrical core from a tree. 
+The second sample type is an entire cross section of a tree, often called a tree cookie.
+Tree cookies present an opportunity to measure as many radii from the sample as desired \citep{speer_fundamentals_2010}.
+Measuring the width between incremental rings from tree samples is first step towards gaining insights.
+
+The first computer based method to measure tree ring widths used a microcomputer, a stage micrometer, and a push button.
 A trained technician shifted a core underneath a microscope objective and pushed a button connected to a microcomputer to record a ring's location \citep{robinson_microcomputer_1980}.
@@
 
-Image analysis of tree rings was designed to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in the 1980s and 1990s to digitize tree ring samples using optical scanning \citep{mcmillin_application_1982} \citep{guay_new_1992}.
+Image analysis of tree rings was later implemented to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in the 1980s and 1990s to digitize tree ring samples using optical scanning \citep{mcmillin_application_1982} \citep{guay_new_1992}.
 Optical scanners are still readily available for purchase, using a top of the line scanner, like the Epson Perfection V850 Pro produces maximum resolutions of 6 400 dpi and scan an area of up to 21.6 cm x 29.7 cm.
 More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach.
-Samples are digitized using an image-capturing system built using motors to move the sample, a microscope camera to take images, a computer to manage the capturing, and a second, more powerful, computer to stitch images into a larger composite image.
+Samples are digitized using an image-capturing system built using a microscope camera to take images, motors to move the sample or camera, a computer to manage the capturing, and a second, more powerful, computer to stitch images into a larger composite image.
 Decreasing the field of view of each captured image increases the resolution of the final stitched image without restricting the size of the sample \citep{muhlich_stitching_2022}.
@@
 For the gantry machine, cartesian movement in the $X$, $Y$ and $Z$ directions is a result of two machine kits and a motor controller from OpenBuilds---the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
-Smaller and larger working planes easily be achieved by purchasing a different sized ACRO kit.
+The machine kit includes extruded aluminum rails, carriages to slide along the rails, stepper motors to control the $X$ and $Y$ movement, and all the hardware to assemble the machine. 
+To fit on the build plates of the ACRO system, we made an adapter to connect the linear actuator to the X and Y axis, thus creating motion in the Z axis.
+The size of the work plane is ultimately up to the size of the ACRO kit that is purchased---allowing for more flexibility.
 Building on top of kits allowed for quick assembly with sound instructions and saved development time. 
-To fit on the build plates of the ACRO system, we made an adapter to connect the linear actuator to the X and Y axis, thus creating motion in the Z axis.
 For the camera, we chose a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens and connected this to the gantry machine using custom adapters.
-For the computer, we implemented all of the software control and image stitching to run on an NVIDIA Jetson Orin Nano (Jetson) which also connects to the camera via a CSI cable.
+For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson) which also connects to the camera via a MIPI cable.
+Controlling the machine is done by launching a Python program and interacting with mouse button clicks on the GUI.
 
@@
 TIM can scan both tree cookies and tree cores. Sample preparation for both are equally important but require different approaches.
-Similar to general sample preparation methods for tree rings samples (CITE dendro manual), cookies and cores need to be sanded with incrementally increasing sand paper grit to produce an even surface for scanning.
+The general guidelines used for preparing cookies and cores applies---requiring samples to be sanded with incrementally increasing sand paper grit to produce an even surface for scanning \citep{speer_fundamentals_2010}.
 We sanded cookies using an orbital sander with mesh sandpaper from 60 to 800 grit, then used a microscope to visually inspect the quality of the sanding.
@@
 
-Special sanding considerations need to be taken for each sample type. 
+Additional sanding considerations need to be taken for each sample type. 
 For cookies, it is important to have the top and bottom surfaces be nearly parallel.
@@
 For cores, care should be taken to maximize the scanning surface. This means removing material to be coincident to the center of the cross section of the core. 
-Similarly to cookies, care should be given to achieving a scanning surface as parallel to the XY-plane as possible.
+Similarly to cookies, care should be given to achieving a scanning surface as parallel to the XY-plane as possible but generally do not need to be leveled using the levelling tables.
 Cores also need to be aligned to be parallel to the $Y$ axis to be digitized, using squaring tool or similar to help with alignment.
-Any warping in core mounts need to be counteracted with the use of trigger clamps.
+Any warping in core mounts needs to be counteracted with the use of trigger clamps.
 
@@
 The height, width, identifiers, sample type, and centering of the sample is sufficient information to digitize.
+
+\subsubsection{Exhaustive Sample Traverse}  
+% It would be super nice to have a concept image of this process! A - I've struggled a bit with how to convey this concisely, let me know what you think about the new figure
+\begin{figure}
+  \centering
+  \includegraphics[height=0.4\linewidth]{../content/cookie_traverse.png}
+  \caption{Scanning a sample begins with the machine navigating to the center of the sample which was automatically recorded at sample add time.
+From there, a grid of target locations is calculated for the machine to capture an image at.
+The system systematically navigates to each location and captures an image until all target locations have been visited.
+Core samples are digitized in the same way but only capture images in one dimension.}
+  \label{fig:cookie_traverse}
+\end{figure}
+
+In the GUI the operator indicates the height and width of the field of view in an individual image from the camera. 
+This varies on the focal length of the lens, but we have kept the field of view to be 3mm x 5mm. 
+Once prompted the system begins to exhaustively traverse the surface area of the sample, using the sample height and width entered by the operator.  
+The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors---the basis of image stitching (See Figure \ref{fig:cookie_traverse}). 
+Digitizing cores can be done without needing to traverse in both the X and Y axis. 
+The optimal scanning path for a core is to align the length of cores to the Y axis and only take one column of images.
+Spanning the surface area of a cookie requires movement in both the X and Y axis.
+
+\subsubsection{Image Focusing}
 
@@
 
-\subsection{Image Capture}  
-% It would be super nice to have a concept image of this process! A - I've struggled a bit with how to convey this concisely, let me know what you think about the new figure
-\begin{figure}
-  \centering
-  \includegraphics[height=0.4\linewidth]{../content/cookie_traverse.png}
-  \caption{Scanning a sample begins with the machine navigating to the center of the sample which was automatically recorded at sample add time.
-From there, a grid of target locations is calculated for the machine to capture an image at.
-The system systematically navigates to each location and captures an image until all target locations have been visited.
-Core samples are digitized in the same way but only capture images in one dimension.}
-  \label{fig:cookie_traverse}
-\end{figure}
-
-In the GUI the operator indicates the height and width of the field of view in an individual image from the camera. 
-This varies on the focal length of the lens, but we have kept the field of view to be 3mm x 5mm. 
-Once prompted the system begins to exhaustively traverse the surface area of the sample, using the sample height and width entered by the operator.  
-The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors---the basis of image stitching (See Figure \ref{fig:cookie_traverse}). 
-Digitizing cores can be done without needing to traverse in both the X and Y axis. 
-The optimal scanning path for a core is to align the length of cores to the Y axis and only take one column of images.
-Spanning the surface area of a cookie requires movement in both the X and Y axis.
-
 Centering the design of TIM around a fixed focus camera and lens means we need to focus the images with a software procedure.
@@
 From the first $(X_0, Y_0)$ coordinate captured, multiple images are taken at different $Z$ values in $\boldsymbol{Z_{\text{samples}}}$. 
-We found that taking 11 images at different heights provides a sufficient set of in and out-of-focus images for comparison.
+We found that it is important to have a distance of no more than 0.1 mm between each image in the stack to maximize the likelihood of an in focus image.
+And with a sufficiently wide height range of 1 mm, we need to take  11 images at different heights for comparison.
 The number of images taken is an editable parameter in the configuration file and can be increased, along with the height range, if in focus images are not being obtained.
@@
 A PID control algorithm with a process variable of $i_{max}$ and control variable of $Z_{0,k}$ allows a negative feedback loop to improve focusing across the entire sample \citep{odwyer_summary_2000}.
-Instead of $(d_1 - d_2) < 0.5mm$ (See \ref{fig:realistic_levelling}), the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which allows for a more forgiving sample setup.
+Instead of forcing a remarkably level sample alignment of $(d_1 - d_2) < 0.5mm$ (See Figure \ref{fig:realistic_levelling}), the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which is readily achievable.
 %After the capturing overlapping images across the entire surface area, the images are ready to be stitched.
@@
 Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to improve speed and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
-This is a variation of the standard approach used in 3D printing and CNC control methods. % What is CNC?
+This is a variation of the standard approach used in 3D printing and Computer Numerical Control (CNC) machine control.
 
-\subsection{Accounting for Translation Error Between Core Samples}
+\subsubsection{Accounting for Translation Error Between Core Samples}
 
@@
 First, the camera is moved to the $(X_0, Y_0, Z_0)$ of the new core. 
-From there, the camera is moved across a window, at constant velocity, from $(X_0 - 1/2*{range}, Y_0, Z_0)$ to $(X_0 + 1/2*{range}, Y_0, Z_0)$. 
+From there, the camera is moved across a window, at constant velocity.
+The width of the window, $w$ in millimeters is defined in the machine configuration file and is editable.
+$(X_0 - 1/2*{w}, Y_0, Z_0)$ to $(X_0 + 1/2*{w}, Y_0, Z_0)$.
 Images are captured at equal distance intervals analogously to the image focusing procedure. 
@@
 
-\subsection{Stitching}
+\subsubsection{Stitching}
 
 Image stitching is a well explored field, ranging from panoramic images taken on smart phones to highly tuned microscopy slide stitching. 
-But the basis of stitching remains the same across algorithms. Adjacent images must have a region of overlap. 
-Of the stitching tools with a software API that we tested, only the Python package Stitch2D was able to stitch our images successfully into a grid. % need to reference this
+Stitching one image to another requires both to share a region of overlap. When capturing a sample using TIM, the operator has the ability to choose a percentage amount of overlap between images.
+Most samples are digitized with a percentage overlap between 33\% and 50\%.
+
+After testing numerous image stitching software APIs, we found the Python package Stitch2D most capable of stitching our images successfully. % need to reference this
 This package wraps OpenCV functions for finding distinctive image features using the SIFT feature detector and matches these feature within the overlapping region of adjacent images \citep{lowe_distinctive_2004}.
@@
 This is due to the surface area of cookies requiring more images as it has a square relationship to sample radius.
-For large cookies, the operator can choose to selectively scan a portion of the cookie instead of the whole surface.
-Scanning complete cookies at maximum resolution is unrealistic as the max cookie size is very small.
+For large cookies, the operator can choose to selectively scan a radius of the cookie instead of the whole surface.
+A single radius contains a portion of each ring.
+Scanning complete cookies at maximum DPI is unrealistic as file size limits are quickly encountered with small diameter cookies.
 To fit into the maximum 2.5 GB limits of a TIFF file, the maximum diameter of a cookie scanned at 30\% of maximum resolution is limited to approximately 130 mm.
-This is significantly smaller when compared to the maximum theoretical core sample size.
+This maximum cookie diameter is significantly smaller than the maximum core sample length.
 At maximum resolution, the file size limit
-constrains samples to approximately 1,500 mm in length.
+constrains core samples to approximately 1,500 mm in length.
 Note that a larger ACRO frame would be needed to support this sample length.
-Samples that are larger than this file size limit are still possible to be scanned but they are no longer stored as a TIFF file.
+Samples that are larger than this file size limit can still be scanned but they are no longer stored as a TIFF file.
 An uncompressed binary NumPy memory-mapped array file is produced. These NumPy files are not supported by standard image viewers like ImageJ and would require custom tools to work with.
@@
 Moving forward, TIM could benefit from being integrated into a cloud data storage system with automatic uploads---removing the need to manually transfer completed scans with physical drives.
-Experimenting with sensors to decrease image focusing time would be a way to addition to the system.
 Additionally, developing a framework to run vessel counting or ring identification models while scanning could greatly decrease the time invested in manually identifying tree anatomy.
@@
 A- Sadly not.
+
+Our implementation of focusing images could be improved with the aid of additional sensors.
+Much of the imaging time of samples is the result of taking a vertical stack of images to obtain one in focus image.
+In advanced microscopy applications, it is common to see supplemental sensors used to measure the distance between the subject and lens.
+Adding sensor support and a control loop to focus images could significantly reduce sampling time on the scale of an order of magnitude.
 

### 2025-03-03 docs/writing/paper.tex
Adding better figure description and layout

@@
 \subsubsection{Exhaustive Sample Traverse}
-% It would be super nice to have a concept image of this process! A - I've struggled a bit with how to convey this concisely, let me know what you think about the new figure
+
 \begin{figure}
   \centering
-  \includegraphics[height=0.4\linewidth]{../content/cookie_traverse.png}
-  \caption{Scanning a sample begins with the machine navigating to the center of the sample which was automatically recorded at sample add time.
-From there, a grid of target locations is calculated for the machine to capture an image at.
-The system systematically navigates to each location and captures an image until all target locations have been visited.
-Core samples are digitized in the same way but only capture images in one dimension.}
-  \label{fig:cookie_traverse}
+  \begin{subfigure}{.3\textwidth}
+      \centering
+      \includegraphics[width=.95\linewidth]{../content/cookie_figure_grid.png}  
+      \caption{}
+      \label{SUBFIGURE LABEL 1}
+  \end{subfigure}
+  \begin{subfigure}{.3\textwidth}
+      \centering
+      \includegraphics[width=.95\linewidth]{../content/cookie_figure_traverse_start.png}  
+      \caption{}
+      \label{SUBFIGURE LABEL 2}
+  \end{subfigure}
+  \begin{subfigure}{.3\textwidth}
+      \centering
+      \includegraphics[width=.95\linewidth]{../content/cookie_figure_second_row.png}  
+      \caption{}
+      \label{SUBFIGURE LABEL 3}
+  \end{subfigure}
+  \caption{This figure describes the imaging process of a cookie sample after it has been added by a user.
+In this example, the height and width enclosed the entire surface area of the cookie.
+(a) A grid of target coordinates is generated to systematically capture partially overlapping images.
+(b) The first image is taken a the center coordinate of the sample.
+The rectangular field of view from the camera moves to adjacent target coordinates until it reaches the left boundary---capturing an image for each target coordinate.
+(c) After completing the row, the machine moves to a new row and continues capturing adjacent images.
+}
+  \label{FIGURE LABEL}
   \end{figure}

### 2025-03-10 docs/writing/paper.tex
New round of edits by EMW

@@
 Please delete my comments in the tex once you have dealt with them!
+%emw9Mar:
 
-\begin{abstract} 
-Studies of tree rings (dendrochronology) have benefitted recently from new methods to image wood rapidly and efficiently, helping to scale up the number and quality of image samples.
+\begin{abstract}  %emw9Mar: I believe dendro is tree ring DATING, so I adjusted below to be more inclusive. 
+Studies of tree rings from climatology to ecology have benefitted recently from new methods to image wood rapidly and efficiently, helping to scale up the number and quality of image samples.
 Current open-source approaches are, however, limited in dimensionality and can be cost-prohibitive.
@@
 
-\section{Introduction}
-Incremental growth rings from trees provide a valuable proxy record of the environment and climate.
-They have proven to provide the highest accuracy, precision, and reliable dating among alternatives \citep{mann_northern_1999}.
-Multiple fields---from climatology to forestry and ecology---rely on tree rings to understand how the environment affects tree growth.
-Tree rings have been critical to
-reconstructing past climates of regional and local environments, informing our understanding of climate change today \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
-They are also increasingly used in ecology, to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017} and have implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}.
-As tree growth and forest management becomes more critical to mitigating climate change, scaling up the collection and processing of tree ring samples has become increasingly important.
+\section{Introduction} %emw9Mar:  edits throughout to tighten things up... please check I did not add typos or nonsensical stuff. 
+Incremental growth rings from trees have provided valuable insights into the abiotic and biotic environment across multiple fields---from climatology to forestry and ecology.
+Dated tree rings from dendrochronology have been critical to
+reconstructing past climates of regional and local environments, providing the most accurate, precise, and reliable dating among alternatives \citep{mann_northern_1999}, and thus fundamnentally informing  our understanding of climate change today \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
+They are also increasingly used in ecology to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017} and have implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}.
+These fields have leveraged two basic types of tree sampling techniques to capture the variation of tree rings: cores, which are cylindrical samples collected using an incremental borer, and `cookies,' which are entire cross sections \citep[and thus allow measuring as many radii from the sample as desired,][]{speer_fundamentals_2010}.
+As understanding tree growth and managing forest becomes more critical to mitigating climate change, scaling up the collection and processing of tree ring samples has become increasingly important.
 
-% Can't find a proper citation with an explicit tree core/cookie definition. Will try harder later.
-Two types of tree sampling techniques are used to capture the variation of tree rings. The first sample type is collected using an incremental borer, used to remove a cylindrical core from a tree. 
-The second sample type is an entire cross section of a tree, often called a tree cookie.
-Tree cookies present an opportunity to measure as many radii from the sample as desired \citep{speer_fundamentals_2010}.
-Measuring the width between incremental rings from tree samples is first step towards gaining insights.
-
-The first computer based method to measure tree ring widths used a microcomputer, a stage micrometer, and a push button.
-A trained technician shifted a core underneath a microscope objective and pushed a button connected to a microcomputer to record a ring's location \citep{robinson_microcomputer_1980}.
-This method was high precision but depended heavily on specific technicians and maintained no image record \citep{levanic_atrics_2007}.
+The first computer-based method to measure tree ring widths used a microcomputer, a stage micrometer, and a push button.
+With this technique, still used widely today, a trained technician shifts a core underneath a microscope objective and pushes a button connected to a microcomputer to record a ring's location \citep{robinson_microcomputer_1980}.
+% 's okay here
+This method has high precision but depends heavily on specific technicians and maintains no image record \citep{levanic_atrics_2007}.
 
@@
 Optical scanners are still readily available for purchase, using a top of the line scanner, like the Epson Perfection V850 Pro produces maximum resolutions of 6 400 dpi and scan an area of up to 21.6 cm x 29.7 cm.
-More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach.
-Samples are digitized using an image-capturing system built using a microscope camera to take images, motors to move the sample or camera, a computer to manage the capturing, and a second, more powerful, computer to stitch images into a larger composite image.
+More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach---using an image-capturing system.
+These systems rely on a microscope camera to take images, motors to move the sample or camera, a computer to manage the capturing, and a second, more powerful, computer to stitch images into a larger composite image.
 Decreasing the field of view of each captured image increases the resolution of the final stitched image without restricting the size of the sample \citep{muhlich_stitching_2022}.
-Stitching multiple images into one mosaic is a common technique used in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}. 
-The first implementation of the capture and stitch method was the Advanced Tree-Ring Image Capturing System (ATRICS) \citep{levanic_atrics_2007}.
-% Adding ATRICS after introducing the image and stitch method
+Stitching multiple images into one mosaic is a common technique used in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}, which was
+first implemented for tree rings through the Advanced Tree-Ring Image Capturing System (ATRICS) \citep{levanic_atrics_2007}. % Adding ATRICS after introducing the image and stitch method
 
@@
 % Then I would end on the cookies issue and also anything else to mention, like cost? You should also clearly mention when/how stitching happens in these approaches.
-Since ATRICS was introduced in 2007, image capture and stitching of tree rings has become more common.
+Since ATRICS was introduced in 2007, image capture and stitching of tree rings have become more common.
 Today CaptuRING offers a more modern do-it-yourself alternative to ATRICS, creating scans up to 5 339 dpi using a DSLR camera and motorized stage \citep{garcia-hidalgo_capturing_2022}.
@@
 Additionally, both of these systems are designed to capture images along a singular axis as the entire width of a core fits within a single image's field of view.
-This assumption fundamentally restricts the digitization to be limited only to cores, as cookies have a two dimensional scanning surface.
+This assumption fundamentally restricts the digitization to be limited only to cores, as cookies have a two dimensional scanning surface. %emw9Mar: later 2/3 of paragraph very nice.
 
@@
 Our design requires the use of 3D printed components and common hand tools such as allen keys and wrenches. No specialized equipment or power tools are required beyond this.
-Excluding 3D printed parts, the total cost of the machine is approximately \$2 200 USD compared to the \$70 000 USD of the Gigapixel \citep{griffin_gigapixel_2021}.
+Excluding 3D printed parts, the total cost of the machine is approximately \$2 200 USD \citep[compared to the \$70 000 USD of the Gigapixel,][]{griffin_gigapixel_2021}.
 We chose to use a Raspberry Pi HQ camera with a microscope lens to save on purchasing a professional camera with a macro lens.
@@
 
-% Nice active voice below!
-By choosing this combination, we were able to reduce the weight and cost of the camera significantly, which allowed us to invest more in a powerful computer that can handle intensive image processing.
+%emw9Mar: check my edits are accurate, I think having a reason beyond getting a nicer computer would be good and is true to our decision making. 
+By choosing this combination, we were able to reduce the weight and cost of the camera significantly, which reduced design challenges in stabilizing movement of a heavier DSLR and allowed us to invest more in a powerful computer that can handle intensive image processing.
 The Jetson is an edge computer which drives a monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for control, and performs calculations to stitch individual images into one mosaic.
@@
 
-Additional sanding considerations need to be taken for each sample type. 
-For cookies, it is important to have the top and bottom surfaces be nearly parallel.
-Small differences in plane angle can be corrected using the 3D printed levelling table we designed (See Figure \ref{fig:ideal_levelling}).
+%emw9Mar: You have to number figures in the order they are introduced, but I think you could introduce Fig 2 below under 'nearly parallel' and then you don't need to re-order. 
+TIM requires additional sanding considerations for each sample type. 
+For cookies, it is important to have the top and bottom surfaces nearly parallel.
+Small differences in plane angle can be corrected using the 3D printed leveling table we designed (See Figure \ref{fig:ideal_levelling}).
 For cores, care should be taken to maximize the scanning surface. This means removing material to be coincident to the center of the cross section of the core. 
-Similarly to cookies, care should be given to achieving a scanning surface as parallel to the XY-plane as possible but generally do not need to be leveled using the levelling tables.
+Similarly to cookies, the scanning surface should be as close to parallel to the XY-plane as possible, but generally does not need to be leveled using the leveling tables.
 Cores also need to be aligned to be parallel to the $Y$ axis to be digitized, using squaring tool or similar to help with alignment.
@@
 
-The subsystems of TIM can be best understood by following the process from sample setup through through obtaining a stitched image. 
+The subsystems of TIM can be best understood by following the process from sample setup through obtaining a stitched image. 
 To achieve this with a fixed focus camera, the samples must be nearly orthogonal to the camera lens. 
@@
       \label{SUBFIGURE LABEL 3}
-  \end{subfigure}
-  \caption{This figure describes the imaging process of a cookie sample after it has been added by a user.
-In this example, the height and width enclosed the entire surface area of the cookie.
-(a) A grid of target coordinates is generated to systematically capture partially overlapping images.
-(b) The first image is taken a the center coordinate of the sample.
-The rectangular field of view from the camera moves to adjacent target coordinates until it reaches the left boundary---capturing an image for each target coordinate.
-(c) After completing the row, the machine moves to a new row and continues capturing adjacent images.
+  \end{subfigure} %emw9Mar: Avoid 'this figure shows ...' or anything similar. Also, check my edits here closely!
+  \caption{An example imaging process of a cookie sample.
+Here, the height and width are enclosed the entire surface area of the cookie and TIM begins by generating a grid of target coordinates to systematically capture partially overlapping images (a).
+TIM takes its first image at the center coordinate of the sample, then the rectangular field of view from the camera moves to adjacent target coordinates until it reaches the left boundary---capturing an image for each target coordinate (b).
+After completing the row, TIM moves to a new row and continues capturing adjacent images (c).
 }
@@
 In the GUI the operator indicates the height and width of the field of view in an individual image from the camera. 
-This varies on the focal length of the lens, but we have kept the field of view to be 3mm x 5mm. 
+This varies on the focal length of the lens, but we have kept the field of view to be 3 mm x 5 mm. 
 Once prompted the system begins to exhaustively traverse the surface area of the sample, using the sample height and width entered by the operator.  
@@
 
-Centering the design of TIM around a fixed focus camera and lens means we need to focus the images with a software procedure.
-The only controllable parameter to focus a fixed focus lens is by changing the lens's distance to the sample. 
-And with the microscope lens, the depth of field of the image is so sensitive that sub millimeter heights can move an image out of focus.
+Centering the design of TIM around a fixed focus camera and lens meant we needed to focus the images with a software procedure.
+The only controllable parameter to focus a fixed focus lens is by changing the distance of the lens relative to the sample. 
+With TIM's microscope lens, the depth of field of the image is sensitive enough that sub-millimeter heights can move an image out of focus.
 Solving this in a time-efficient manner involves two stages. 
 The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
-This allows the initial $Z_0$ value, captured when adding a sample, to be an informed guess as to what value would have an in focus image across the entire surface. 
+This allows the initial $Z_0$ value, captured when adding a sample, to be an informed guess as to what value would have an in-focus image across the entire surface. 
 From the first $(X_0, Y_0)$ coordinate captured, multiple images are taken at different $Z$ values in $\boldsymbol{Z_{\text{samples}}}$.
-We found that it is important to have a distance of no more than 0.1 mm between each image in the stack to maximize the likelihood of an in focus image.
-And with a sufficiently wide height range of 1 mm, we need to take  11 images at different heights for comparison.
-The number of images taken is an editable parameter in the configuration file and can be increased, along with the height range, if in focus images are not being obtained.
+%emw9Mar: I think we need to be a little more careful in defining the math as we introduce each new term.
+One way to start on this would be to make a supplemental table with ALL the math we introduce and a brief description of it.
+If you can work that up then we could meet and fix this quickly I suspect.
+We found that it is important to have a distance of no more than 0.1 mm between each image in the stack to maximize the likelihood of an in-focus image.
+With a sufficiently wide height range of 1 mm, TIM needs to take  11 images at different heights for comparison.
+The number of images taken is an editable parameter in the configuration file and can be increased, along with the height range, if in-focus images are not being obtained.
 
@@
 The 11 images then have their normalized variance, $NV$, calculated in a separate thread to measure the image sharpness \citep{sampat_extensive_2014}.
-The image with the maximum score is saved and can be considered the most in focus image, while the rest are deleted from storage.
+The image with the maximum score is saved and can be considered the most in-focus image, while the rest are deleted.
 
@@
 
-The second component of a time-efficient focusing procedure is using a feedback loop to update the initial $Z_0$ at each image location. 
+The second stage of our time-efficient focusing procedure is using a feedback loop to update the initial $Z_0$ at each image location. 
 Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm, $d_1 - d_2 < 0.5 mm $ (See Figure \ref{fig:realistic_levelling}). 
 Rather than adjusting this range, a greater alignment error can be managed by controlling the center of the range, $Z_{0,k}$, for $(X_k,Y_k)$. 
-The likelihood of an adjacent $(X_{k+1}, Y_{k+1})$ containing an in focus image is highest when the current $i_{{\max,k}}$ is at the middle index of $\boldsymbol{Z_{samples}}$. 
+The likelihood of an adjacent %emw9Mar:  we need a noun here instead of just $(X_{k+1}, Y_{k+1})$ ... maybe we need a term for this?
+$(X_{k+1}, Y_{k+1})$ containing an in-focus image is highest when the current $i_{{\max,k}}$ is at the middle index of $\boldsymbol{Z_{samples}}$. 
 A PID control algorithm with a process variable of $i_{max}$ and control variable of $Z_{0,k}$ allows a negative feedback loop to improve focusing across the entire sample \citep{odwyer_summary_2000}.
@@
 To reduce motion blur in the images, the range of $\boldsymbol{Z_{\text{samples}}}$ is traversed at constant velocity and images are captured without stopping. 
-Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to improve speed and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
+Decoupling the auxiliary camera control from the G-code commands to control the machine improved speed and decreased vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
 This is a variation of the standard approach used in 3D printing and Computer Numerical Control (CNC) machine control.
@@
 First, the camera is moved to the $(X_0, Y_0, Z_0)$ of the new core. 
-From there, the camera is moved across a window, at constant velocity.
-The width of the window, $w$ in millimeters is defined in the machine configuration file and is editable.
+From there, the camera is moved across a window, at constant velocity. %emw9Mar:  w is introduced nicely below. Need to try to follow this format when possible. 
+The width of the window, $w$, in millimeters is defined in the machine configuration file and is editable. %emw9Mar:  we need a noun below ...
 $(X_0 - 1/2*{w}, Y_0, Z_0)$ to $(X_0 + 1/2*{w}, Y_0, Z_0)$. 
@@
 After testing numerous image stitching software APIs, we found the Python package Stitch2D most capable of stitching our images successfully. % need to reference this
-This package wraps OpenCV functions for finding distinctive image features using the SIFT feature detector and matches these feature within the overlapping region of adjacent images \citep{lowe_distinctive_2004}.
+This package wraps OpenCV functions for finding distinctive image features using the SIFT feature detector and matches these features within the overlapping region of adjacent images \citep{lowe_distinctive_2004}.
 The default implementation of the package produces high quality stitches but is memory inefficient. %is this wording correct?, I'm trying to kindly say it was memory inefficient
@@
 Some tree ring analyses may not need the maximum scanning resolution produced by TIM. At maximum resolution, TIM's scans quickly produce large file sizes and can be costly to store at scale. 
-To address this, a parameter in the machine configuration file allows the operator to choose a downsizing ratio to apply to images before they get stitched, decreasing the file size of the scans.
+To address this, a parameter in the machine configuration file allows the operator to choose a downsizing ratio to apply to images before they are stitched, decreasing the file size of the scans.
 With the same images, multiple final stitched resolutions can be made.
 
-\section{Results}
+\section{Results} %emw9Mar: I wondered about this being Results & Discussion ...
+would MEE be okay with that?
+It would work better for the paper I think.
+What do the guidelines say on formatting for articles?
 
@@
   \end{subfigure}
-  \caption{Example of a single DPI measurement using a 0.01mm slide scale.
-The DPI was measured at multiple locations in the field of view both vertically and horizontally.
-They converged on the same value.}
+  \caption{Example of a single DPI measurement using a 0.01 mm slide scale.
+We found the DPI measured at multiple locations in the field of view vertically and horizontally  converged on the same value.}
   \label{fig:dpi_measurements}
@@
 TIM is capable of producing extremely high resolution scans.
-A microscopy slide scale with 0.01mm graduations was used to calculate the dots per inch (DPI) of individual images captured from the system (Figure \ref{fig:dpi_measurements}).  
-Using this scale we calculated the maximum DPI to be 21,140.
-This calculation can be replicated using the Python tool we created as necessary for varied lens focal lengths.
+We calculated the maximum dots per inch (DPI) at 21 140 of individual images captured from the system using a microscopy slide scale (with 0.01 mm graduations, Figure \ref{fig:dpi_measurements}), and developed a Python tool to replicate this for varied lens focal lengths.
 When downscaling the final stitch,
@@
 At maximum resolution, the file size limit
-constrains core samples to approximately 1,500 mm in length.
-Note that a larger ACRO frame would be needed to support this sample length.
-Samples that are larger than this file size limit can still be scanned but they are no longer stored as a TIFF file.
-An uncompressed binary NumPy memory-mapped array file is produced. These NumPy files are not supported by standard image viewers like ImageJ and would require custom tools to work with.
+constrains core samples to approximately 1 500 mm in length (though a larger ACRO frame than shown here would be needed to support this sample length).
+Samples larger than this limit can still be scanned, but TIM saves them as an uncompressed binary NumPy memory-mapped array file is produced.
+Currently, NumPy files are not supported by standard image viewers like ImageJ and would require custom tools to work with.
+%emw9Mar: Like custom Python scripts?
+I think we might be more explicit here if we mean that.
 
@@
 \section{Discussion} % I think we could remind the reader here of what gaps TIM addresses, but I can help do that next time.
+%emw9Mar: This section could move down and become conclusions if we merge the Results & Discussion.
 By taking advantage of a common three degree of freedom cartesian machine design, powerful edge computing, and a microscope camera we were able to design a cost effective and high resolution digitization tool for wood samples.
@@
 
-\subsection{Strengths and Opportunities}
+\subsection{Strengths and Opportunities} %emw9Mar: And this section could become part of the Results & Discussion.... 
 % add better opportunities
-Our design minimizes the barriers to build a TIM in smaller labs without compromising scan quality.
+TIM's design minimizes the barriers to build an affordable, efficient and high-resolution image capture system in smaller labs without compromising scan quality, and while allowing scanning of cookies and cores.
 We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ.
-Moving forward, TIM could benefit from being integrated into a cloud data storage system with automatic uploads---removing the need to manually transfer completed scans with physical drives.
-Additionally, developing a framework to run vessel counting or ring identification models while scanning could greatly decrease the time invested in manually identifying tree anatomy.
-% I guess Sandy never sent in results we could use?
-A- Sadly not.
+Extensions to TIM could allow automatic uploads of images into cloud data storage system with automatic uploads, though this would likely require significant effort for it to work across diverse cloud platforms.
+% removing the need to manually transfer completed scans with physical drives.
+%emw9Mar: But we just do this because of our wifi issues ...
+others could manually upload them to the cloud, right?
+So I suggest we remove this.
+%emw9Mar: Any citations for developing repos to share tree ring images?
+Or just more on building better open shared data for tree rings?
+We should also cite some of the new vessel and ring ID papers in CITES2
+One opportunity for this development could come as more universal and large-scale open-data storage of tree ring images come online CITES.
+Such open repositories would also increase the value of developing new frameworks count vessel or identify rings, which are already starting and have the potential to greatly decrease the time invested in manually identifying tree anatomy CITES2.
+With these more fully developed TIM could be extended to be run such frameworks while scanning, possibly taking additional scans only when needed to reach sufficient algorithm accuracy.
+%emw9Mar -- check my riffing here.
 
-Our implementation of focusing images could be improved with the aid of additional sensors.
-Much of the imaging time of samples is the result of taking a vertical stack of images to obtain one in focus image.
-In advanced microscopy applications, it is common to see supplemental sensors used to measure the distance between the subject and lens.
-Adding sensor support and a control loop to focus images could significantly reduce sampling time on the scale of an order of magnitude.
+Finally, an exciting frontier for machines such as TIM is integrating additional sensors.
+While much of the current imaging time of samples comes from taking a vertical stack of images to obtain one in-focus image, advanced microscopy applications have overcome similar challenges through supplemental sensors that measure the distance between the subject and lens.
+Adding such sensor support and a control loop to focus images could significantly reduce sampling time on the scale of an order of magnitude.
 

### 2025-03-19 docs/writing/paper.tex
Mentioning adding encoders for more accurate movement

--- 2025-03-03 docs/writing/paper.tex
@@
 To reduce motion blur in the images, the range of $\boldsymbol{Z_{\text{samples}}}$ is traversed at constant velocity and images are captured without stopping. 
-Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to improve speed and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
+Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to decrease digitization time and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
 This is a variation of the standard approach used in 3D printing and Computer Numerical Control (CNC) machine control.
@@
 Adding sensor support and a control loop to focus images could significantly reduce sampling time on the scale of an order of magnitude. 
+Using rotary encoders on each of the motors controlling the X and Y movement would be a welcomed addition---this would greatly increase the 
+accuracy of the the location of the camera on the work plane.
 

### 2025-03-24 docs/writing/paper.tex

--- 2025-03-10 docs/writing/paper.tex
@@
 To reduce motion blur in the images, the range of $\boldsymbol{Z_{\text{samples}}}$ is traversed at constant velocity and images are captured without stopping. 
-Decoupling the auxiliary camera control from the G-code commands to control the machine improved speed and decreased vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
+Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to decrease digitization time and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
 This is a variation of the standard approach used in 3D printing and Computer Numerical Control (CNC) machine control.
@@
 
+Our implementation of focusing images could be improved with the aid of additional sensors.
+Much of the imaging time of samples is the result of taking a vertical stack of images to obtain one in focus image.
+In advanced microscopy applications, it is common to see supplemental sensors used to measure the distance between the subject and lens.
+Adding sensor support and a control loop to focus images could significantly reduce sampling time on the scale of an order of magnitude. 
+Using rotary encoders on each of the motors controlling the X and Y movement would be a welcomed addition---this would greatly increase the 
+accuracy of the the location of the camera on the work plane.  
 Finally, an exciting frontier for machines such as TIM is integrating additional sensors.

### 2025-03-27 docs/writing/paper.tex
Addressing Lizzies Mar 9 edits.

@@
-% !TeX root = paper.tex
+% !TeX root = ./paper.tex
 \documentclass[a4paper,12pt]{article}
@@
 \usepackage{subcaption}     
+\usepackage{multirow}
+
 % Page layout
@@
 
-\begin{abstract}  %emw9Mar: I believe dendro is tree ring DATING, so I adjusted below to be more inclusive. 
+\begin{abstract}  
 Studies of tree rings from climatology to ecology have benefitted recently from new methods to image wood rapidly and efficiently, helping to scale up the number and quality of image samples.
-Current open-source approaches are, however, limited in dimensionality and can be cost-prohibitive.
+Current approaches are, however, limited in scanning dimensionality and can be cost-prohibitive.
+% Getting rid of open-source as CaptuRING is cheap but GigaPixel is expensive and not open source
 The Tree Imaging Machine (TIM) is a do-it-yourself scanning tool made to digitize tree cookies and cores that address these limitations.
@@
 Operators can prepare sample batches of multiple cores and cookies, letting the machine work until the queue is finished.
-All of the 3D printed parts are available for download on the [NIH 3D printing repository](https://doi.org/10.60705/3DPX/21561.3) and the software is 
+All of the 3D printed parts are available for download on the [NIH 3D printing repository](https://doi.org/10.60705/3DPX/21561.3) and the software is  % is this best practice for including links? 
 open source, with instructions to recreate this tool available on [GitHub](https://github.com/temporalecologylab/TreeRings).
@@
 This promotes the digitization of larger samples and can collect data at finer resolutions in an open-source design which can be built on and improved as new technologies become available.
-% Add to the discussion something about next steps for TIM more clearly?
+% Archive GitHub with Zenodo?
+I'm nervous as the code base is updated relatively frequently...
+will this be a problem?
+Or do I only ever make one 'release' https://docs.github.com/en/repositories/archiving-a-github-repository/referencing-and-citing-content
 \end{abstract}
@@
 
-% I would start a new paragraph here that opens with something like: Since XX was introduced in XX, image capture and stitching of tree rings has increased.
-Today CaptuRING offers a  more modern do-it-yourself alternative to ATRICS that is (open source?
-Has resolutions up to XX?
-Used XX type of camera?
-Give a little more here).
-Another approach, Gigapixel, moves the camera relative to the sample, allowing for multiple samples to be recorded in sequence \citep{griffin_gigapixel_2021}.
-% Then I would end on the cookies issue and also anything else to mention, like cost? You should also clearly mention when/how stitching happens in these approaches.
 Since ATRICS was introduced in 2007, image capture and stitching of tree rings have become more common.
@@
 
-% I will edit this a little next time, but for now fix up em-dashes, and simplify with less PV as possible. 
 We made TIM to combine the open-source and open-hardware philosophies of CaptuRING with the functionality of Gigapixel.
@@
 and otherwise point out the novelty when it comes up.
-TIM can be thought of as a combination of multiple subsystems: a gantry machine, camera, and computer (See Figure \ref{fig:tim_assembled}). 
+TIM can be thought of as a combination of multiple subsystems: a gantry machine, camera, and computer (Figure \ref{fig:tim_assembled}). 
 For the gantry machine, cartesian movement in the $X$, $Y$ and $Z$ directions is a result of two machine kits and a motor controller from OpenBuilds---the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
@@
 
-%emw9Mar: check my edits are accurate, I think having a reason beyond getting a nicer computer would be good and is true to our decision making. 
 By choosing this combination, we were able to reduce the weight and cost of the camera significantly, which reduced design challenges in stabilizing movement of a heavier DSLR and allowed us to invest more in a powerful computer that can handle intensive image processing.
@@
 \subsection{Preparing Cookies and Cores}
-% Tried to reduce PV below. 
 TIM can scan both tree cookies and tree cores. Sample preparation for both are equally important but require different approaches.
@@
 
-%emw9Mar: You have to number figures in the order they are introduced, but I think you could introduce Fig 2 below under 'nearly parallel' and then you don't need to re-order. 
 TIM requires additional sanding considerations for each sample type. 
 For cookies, it is important to have the top and bottom surfaces nearly parallel. 
-Small differences in plane angle can be corrected using the 3D printed leveling table we designed (See Figure \ref{fig:ideal_levelling}).
+\begin{figure}
+  \centering
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[height=0.5\linewidth]{../diagrams/sample_setup_ideal.png}
+    \caption{Ideal sample leveling}
+    \label{fig:ideal_levelling}
+  \end{subfigure}%
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[height=0.5\linewidth]{../diagrams/sample_setup_realistic.png}
+    \caption{Realistic sample leveling}
+    \label{fig:realistic_levelling}
+  \end{subfigure}
+  \caption{Side view of the camera and sample on top of a leveling table.
+The ideal sample leveling shows a uniform distance d at all $(X,Y)$ coordinates on the sample.
+This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique $(X_i, Y_i)$.}
+  \label{fig:sample_levelling}
+\end{figure}
+Small differences in plane angle can be corrected using the 3D printed leveling table we designed (Figure \ref{fig:ideal_levelling}). 
 For cores, care should be taken to maximize the scanning surface. This means removing material to be coincident to the center of the cross section of the core. 
@@
 
-In the GUI the operator indicates the height and width of the field of view in an individual image from the camera. 
+In the GUI, the operator indicates the height and width of the field of view in an individual image from the camera. 
 This varies on the focal length of the lens, but we have kept the field of view to be 3 mm x 5 mm. 
 Once prompted the system begins to exhaustively traverse the surface area of the sample, using the sample height and width entered by the operator.  
-The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors---the basis of image stitching (See Figure \ref{fig:cookie_traverse}). 
+The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
 Digitizing cores can be done without needing to traverse in both the X and Y axis. 
@@
 \subsubsection{Image Focusing}
-
-\begin{figure}
-  \centering
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[height=0.5\linewidth]{../diagrams/sample_setup_ideal.png}
-    \caption{Ideal sample leveling}
-    \label{fig:ideal_levelling}
-  \end{subfigure}%
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[height=0.5\linewidth]{../diagrams/sample_setup_realistic.png}
-    \caption{Realistic sample leveling}
-    \label{fig:realistic_levelling}
-  \end{subfigure}
-  \caption{Side view of the camera and sample on top of a leveling table.
-The ideal sample leveling shows a uniform distance d at all $(X,Y)$ coordinates on the sample.
-This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique $(X_i, Y_i)$.}
-  \label{fig:sample_levelling}
-\end{figure}
 
@@
 \[
-  i_{\max} = \arg\max_{i} NV(image(\boldsymbol{{Z_\text{focus}})})
+  i_{\max} = \arg\max_{i} NV(image(\boldsymbol{{Z_\text{samples}})})
   \]
@@
 The second stage of our time-efficient focusing procedure is using a feedback loop to update the initial $Z_0$ at each image location. 
-Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm, $d_1 - d_2 < 0.5 mm $ (See Figure \ref{fig:realistic_levelling}). 
+Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm, $d_1 - d_2 < 0.5 mm $ (Figure \ref{fig:realistic_levelling}). 
 Rather than adjusting this range, a greater alignment error can be managed by controlling the center of the range, $Z_{0,k}$, for $(X_k,Y_k)$. 
@@
 A PID control algorithm with a process variable of $i_{max}$ and control variable of $Z_{0,k}$ allows a negative feedback loop to improve focusing across the entire sample \citep{odwyer_summary_2000}.
-Instead of forcing a remarkably level sample alignment of $(d_1 - d_2) < 0.5mm$ (See Figure \ref{fig:realistic_levelling}), the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which is readily achievable.
+Instead of forcing a remarkably level sample alignment of $(d_1 - d_2) < 0.5mm$ (Figure \ref{fig:realistic_levelling}), the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which is readily achievable.
 %After the capturing overlapping images across the entire surface area, the images are ready to be stitched.
@@
 
-\section{Results} %emw9Mar: I wondered about this being Results & Discussion ...
-would MEE be okay with that?
-It would work better for the paper I think.
-What do the guidelines say on formatting for articles?
+\section{Results \& Discussion} 
+%emw9Mar: I wondered about this being Results & Discussion ... would MEE be okay with that? It would work better for the paper I think. What do the guidelines say on formatting for articles? 
+% MEE's guidelines have a distinct Results section from Discussion although maybe they're flexible on this? Conclusion is optional for MEE
 
@@
 The range of sample included are from a 3mm x 220mm core to a 75mm x 56mm cookie.
-True sampling times are drastically influenced by configurable machine parameters.} % Take the title on top off this image as we don't have titles on images in sci papers.
+True sampling times are drastically influenced by configurable machine parameters.}
     \label{fig:digitization_time}
@@
 
-\section{Discussion} % I think we could remind the reader here of what gaps TIM addresses, but I can help do that next time.
+\subsection{Strengths and Opportunities} %emw9Mar: And this section could become part of the Results & Discussion.... 
+% add better opportunities
+TIM's design minimizes the barriers to build an affordable, efficient and high-resolution image capture system in smaller labs without compromising scan quality, and while allowing scanning of cookies and cores.
+We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ.
+%emw9Mar: Any citations for developing repos to share tree ring images? Or just more on building better open shared data for tree rings? 
+% It appears databases store data after images are processed and measured, potentially too expensive to store image datasets at high resolution?
+Publicly available tree-ring datasets can be found on multiple online repositories such as TreeSource's Tree-Rings Database for Canadian sites or the International Tree-Ring Data Bank (ITRDB) for worldwide data \citep{girardin_national_2021} \citep{grissino-mayer_international_1997}.
+Both of these databases offer metadata rich data on measured tree-rings but do not store the source images of samples.
+One opportunity for this development could come as more universal and large-scale open-data storage of tree ring images come online. 
+Such open repositories would also increase the value of developing new frameworks count vessel or identify rings, which are already starting and have the potential to greatly decrease the time invested in manually identifying tree anatomy \citep{resente_mask_2021} \citep{polacek_automation_2023}.
+With these more fully developed TIM could be extended to be run such frameworks while scanning, possibly taking additional scans only when needed to reach sufficient algorithm accuracy.
+%emw9Mar -- check my riffing here.
+
+% Sandy got back to me! :)
+We tested a scanned cookie from TIM with the R-CNN presented by \citeauthor{polacek_automation_2023} on two servers.
+The first being the high-performance UBC ARC Sockeye which completed the task in 15 minutes.
+As many do not have access to such powerful computing, we also tested
+the same image on an RTX A6000 GPU server---taking 3.5 hours to complete the task for approximately \$3.50 USD. 
+While the results of automated tree rings are promising, further testing is necessary to test the feasibility of running this model on an edge computer or developing integrations for TIM with a remote server.
+
+Finally, an exciting frontier for machines such as TIM is integrating additional sensors.
+While much of the current imaging time of samples comes from taking a vertical stack of images to obtain one in-focus image, advanced microscopy applications have overcome similar challenges through supplemental sensors that measure the distance between the subject and lens.
+Adding such sensor support and a control loop to focus images could significantly reduce sampling time on the scale of an order of magnitude. 
+Using rotary encoders on each of the motors controlling the X and Y movement would be a welcomed addition---this would greatly increase the 
+accuracy of the the location of the camera on the work plane.
+
+\section{Conclusion} % I think we could remind the reader here of what gaps TIM addresses, but I can help do that next time.
 %emw9Mar: This section could move down and become conclusions if we merge the Results & Discussion.
@@
 
-\subsection{Strengths and Opportunities} %emw9Mar: And this section could become part of the Results & Discussion.... 
-% add better opportunities
-TIM's design minimizes the barriers to build an affordable, efficient and high-resolution image capture system in smaller labs without compromising scan quality, and while allowing scanning of cookies and cores.
-We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ.
-Extensions to TIM could allow automatic uploads of images into cloud data storage system with automatic uploads, though this would likely require significant effort for it to work across diverse cloud platforms.
-% removing the need to manually transfer completed scans with physical drives.
-%emw9Mar: But we just do this because of our wifi issues ...
-others could manually upload them to the cloud, right?
-So I suggest we remove this.
-%emw9Mar: Any citations for developing repos to share tree ring images?
-Or just more on building better open shared data for tree rings?
-We should also cite some of the new vessel and ring ID papers in CITES2
-One opportunity for this development could come as more universal and large-scale open-data storage of tree ring images come online CITES.
-Such open repositories would also increase the value of developing new frameworks count vessel or identify rings, which are already starting and have the potential to greatly decrease the time invested in manually identifying tree anatomy CITES2.
-With these more fully developed TIM could be extended to be run such frameworks while scanning, possibly taking additional scans only when needed to reach sufficient algorithm accuracy.
-%emw9Mar -- check my riffing here.
+\subsection{Supplemental Information}
 
-Our implementation of focusing images could be improved with the aid of additional sensors.
-Much of the imaging time of samples is the result of taking a vertical stack of images to obtain one in focus image.
-In advanced microscopy applications, it is common to see supplemental sensors used to measure the distance between the subject and lens.
-Adding sensor support and a control loop to focus images could significantly reduce sampling time on the scale of an order of magnitude. 
-Using rotary encoders on each of the motors controlling the X and Y movement would be a welcomed addition---this would greatly increase the 
-accuracy of the the location of the camera on the work plane.  
-Finally, an exciting frontier for machines such as TIM is integrating additional sensors.
-While much of the current imaging time of samples comes from taking a vertical stack of images to obtain one in-focus image, advanced microscopy applications have overcome similar challenges through supplemental sensors that measure the distance between the subject and lens.
-Adding such sensor support and a control loop to focus images could significantly reduce sampling time on the scale of an order of magnitude.
+\begin{tabular}{ |p{3cm}||p{12cm}|  }
+  \hline
+  \multicolumn{2}{|c|}{Math Notation} \\
+  \hline
+  Terms & Definition\\
+  \hline
+  $X_0, Y_0, Z_0$   &The initial cartesian coordinates of the camera when the sample is saved by the user. \\
+  $(X_k, Y_k, Z_k)$ & A unique set of $X$, $Y$, and $Z$ coordinates for the center point of an individual image $k$. \\
+  $(X_{k+1}, Y_{k+1})$ & The $X$ and $Y$ coordinates of an image that neighbors image $k$ at $(X_k, Y_k)$. \\
+  $Z_{0,i}$ & The starting $Z$ value to move to when moving to the adjacent image. This value is also the center value of $\boldsymbol{Z_{\text{samples}}}$ \\
+  $NV()$ & Normalized variance 'function.' \\
+  $Z_{focused}$ & The $Z$ value of the most in focus image in $\boldsymbol{Z_{\text{samples}}}$\\
+  $\boldsymbol{Z_{\text{samples}}}$ & A vector containing the $Z$ coordinate for multiple images at the same $X_k, Y_k$. \\
+  $i_{\max,k}$ & The index of the image with the largest normalized variance at the same $(X_k, Y_k)$. \\
+  $w$ & The width of the window to traverse when trying to center a core. \\
+  $image()$ & The image at a corresponding $Z$ value from $\boldsymbol{Z_{\text{samples}}}$ \\
+  $d_1$ & The relative distance from the camera to one edge of a levelled sample. \\
+  $d_2$ & The relative distance from the camera to the opposite edge of a levelled sample. \\
+
+  \hline
+ \end{tabular}
 

### 2025-04-01 docs/writing/paper.tex
Updated bibliography to besjournals.bst. Figured out the build problems was from the CaptuRING reference. Accented 'A' upset everything
Fixing some bibtex stuff

@@
 \usepackage{geometry}           
-\usepackage[numbers]{natbib}            
+\usepackage{natbib}   
+\bibliographystyle{besjournals}
+
 \usepackage{hyperref}          
@@
 Dated tree rings from dendrochronology have been critical to
-reconstructing past climates of regional and local environments, providing the most accurate, precise, and reliable dating among alternatives \citep{mann_northern_1999}, and thus fundamnentally informing  our understanding of climate change today \citep{fritts_dendroclimatology_1971} \citep{williams_using_2010} \citep{guibal_dendrochronology_2021} \citep{sheppard_dendroclimatology_2010}.
+reconstructing past climates of regional and local environments, providing the most accurate, precise, and reliable dating among alternatives \citep{mann_northern_1999}, and thus fundamnentally informing  our understanding of climate change today \citep{fritts_dendroclimatology_1971,williams_using_2010,guibal_dendrochronology_2021,sheppard_dendroclimatology_2010}.
 They are also increasingly used in ecology to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017} and have implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}.
@@
 
-\bibliographystyle{plainnat}
 \bibliography{references}

### 2025-04-02 docs/writing/notation_glossary_table.tex
Addressing math notation. Adding glossary for notation in supplemental tables. Adding new file for table

\begin{table}[h]
    \centering
    \caption{Mathematical notation used in this paper.}
    \label{tab:math_notation}
    \begin{tabular}{ |p{3cm}||p{12cm}|  }
      \hline
      \multicolumn{2}{|c|}{Math Notation} \\
      \hline
      Terms & Definition\\
      \hline
      $X_0, Y_0, Z_0$   &The initial cartesian coordinates of the camera when the sample is saved by the user. \\
      $(X_k, Y_k, Z_k)$ & A unique set of $X$, $Y$, and $Z$ coordinates for the center point of an individual image $k$. \\
      $(X_{k+1}, Y_{k+1})$ & The $X$ and $Y$ coordinates of an image that neighbors image $k$ at $(X_k, Y_k)$. \\
      $\boldsymbol{\delta}$ & Equally spaced heights, relative to $Z_0$, to capture images at and use to find a sharp image. \\
      $\alpha_k$ & The response variable of the PID control algorithm which adjusts the initial height of an image to gradually align $Z_{focused}$ to be at the center index of $\boldsymbol{Z_{samples}}$.\\
      $NV()$ & Normalized variance 'function.' \\
      $Z_{focused}$ & The $Z$ value of the most in-focus image in $\boldsymbol{Z_{\text{samples}}}$. \\
      $\boldsymbol{Z_{\text{samples}}}$ & A vector containing the $Z$ coordinate for multiple images at the same $X_k, Y_k$. \\
      $i_{\max,k}$ & The index of the image with the largest normalized variance at the same $(X_k, Y_k)$. \[
      i_{\max} = \arg\max_{i} NV(image(\boldsymbol{{Z_\text{samples}})})
      \] \\
      $w$ & The width of the window to traverse when trying to center a core. \\
      $image()$ & The image at a corresponding $Z$ value from $\boldsymbol{Z_{\text{samples}}}$. \\
      $d_1$ & The relative distance from the camera to one edge of a leveled sample. \\
      $d_2$ & The relative distance from the camera to the opposite edge of a leveled sample. \\
      \hline
    \end{tabular}
    \end{table}

### 2025-04-02 docs/writing/paper.tex
Addressing math notation. Adding glossary for notation in supplemental tables. Adding new file for table

@@
   \caption{Side view of the camera and sample on top of a leveling table.
-The ideal sample leveling shows a uniform distance d at all $(X,Y)$ coordinates on the sample.
-This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique $(X_i, Y_i)$.}
+The ideal sample leveling shows a uniform distance, $d$, at all $XY$ coordinates on the sample.
+This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique $XY$.}
   \label{fig:sample_levelling}
@@
 To achieve this with a fixed focus camera, the samples must be nearly orthogonal to the camera lens. 
-Once the sample is level, the operator interacts with the machine through the GUI to navigate the camera to the center of the sample, and focuses the preview image to be sharp by moving in the Z axis.
-The height and width dimensions are then entered in the GUI
-to be saved along with the detected center coordinate of the sample. This procedure can be repeated to create a queue of samples to digitize as a batch.
-The height, width, identifiers, sample type, and centering of the sample is sufficient information to digitize.
+Once the sample is level, the operator must follow a procedure to inform TIM of where the sample is, and the it's size. 
+To do this, the operator interacts with the machine using buttons on the GUI---navigating and focusing the camera on the center of the sample. 
+The sample height and width in the X and Y dimensions are then entered in the GUI to be saved.
+Finally, the user must click the 'Add Sample' button which provides more fields to add sample identifiers and save the center coordinate.
+This procedure can be repeated to create a queue of samples to digitize as a batch.
+One additional parameter that needs to be defined is the height and width of the field of view of one image from the camera.
+We adjusted the focal length of the camera using the zoom ring to have a constant field of view of approximately 3 mm x 5 mm when in focus.  
+The sample height and width, sample identifiers, sample type, sample center coordinate, and the height and width of an image provides sufficient information to digitize a sample.
 
@@
   \caption{An example imaging process of a cookie sample.
-Here, the height and width are enclosed the entire surface area of the cookie and TIM begins by generating a grid of target coordinates to systematically capture partially overlapping images (a).
+Here, the height and width are enclosed the entire surface area of the cookie and TIM begins by generating a grid of target coordinates to systematically capture a grid of continuous and partially overlapping images (a).
 TIM takes its first image at the center coordinate of the sample, then the rectangular field of view from the camera moves to adjacent target coordinates until it reaches the left boundary---capturing an image for each target coordinate (b).
@@
 
-In the GUI, the operator indicates the height and width of the field of view in an individual image from the camera. 
-This varies on the focal length of the lens, but we have kept the field of view to be 3 mm x 5 mm. 
-Once prompted the system begins to exhaustively traverse the surface area of the sample, using the sample height and width entered by the operator.  
-The goal of this traversal is to obtain in-focus images that have a region of overlap with all its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
+The goal of this traversal is to obtain in-focus images that have a region of overlap with each of its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
 Digitizing cores can be done without needing to traverse in both the X and Y axis. 
@@
 With TIM's microscope lens, the depth of field of the image is sensitive enough that sub-millimeter heights can move an image out of focus.
-Solving this in a time-efficient manner involves two stages. 
+Solving this in a time-efficient manner involves two stages. We provide a complete glossary of notation used to describe the methods, see Table \ref{tab:math_notation}.
 The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
 This allows the initial $Z_0$ value, captured when adding a sample, to be an informed guess as to what value would have an in-focus image across the entire surface. 
-From the first $(X_0, Y_0)$ coordinate captured, multiple images are taken at different $Z$ values in $\boldsymbol{Z_{\text{samples}}}$.
+From the first coordinate captured $(X_0, Y_0, Z_0)$, multiple images are taken at different heights (saved as a vector $\boldsymbol{Z_{\text{samples}}}$).
 %emw9Mar: I think we need to be a little more careful in defining the math as we introduce each new term.
@@
 If you can work that up then we could meet and fix this quickly I suspect.
-We found that it is important to have a distance of no more than 0.1 mm between each image in the stack to maximize the likelihood of an in-focus image.
+The difference in height between each image are equally spaced, and are described by the relative distance to $Z_0$ in the vector $\boldsymbol{\delta}$. 
+We found that it is important to have a distance of no more than 0.1 mm between each image in the stack to maximize the likelihood of an in-focus image at height $Z_{focused}$.
 With a sufficiently wide height range of 1 mm, TIM needs to take 11 images at different heights for comparison.
@@
 
-% Something bad is happening with the i below. <-- is this comment still relevant? I don't see i
+Our focusing procedure begins at 0.5 mm above $Z_0$ and finishes with the last image at 0.5 mm below $Z_0$.
+The 11 images then have their normalized variance calculated in a separate thread to measure the image sharpness \citep{sampat_extensive_2014}.
+The image with the maximum score is saved and can be considered the most in-focus image, while the rest are deleted.
+
+The second stage of our time-efficient focusing procedure is uses a feedback loop to update the initial $Z_0$ at each image location $(X_k, Y_k, Z_k)$. 
+Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm ($d_1 - d_2 < 0.5 mm $ in Figure \ref{fig:realistic_levelling}). 
+Rather than increasing this height range, a greater alignment error can be managed by adjusting the starting image height until $Z_{focused}$ is at the center index of $\boldsymbol{Z_{samples}}$. 
+The likelihood of an adjacent target
+$(X_{k+1}, Y_{k+1})$ containing an in-focus image is highest in this scenario. 
+A negative feedback PID control algorithm is used to adjust the initial height of each image, with $\alpha_k$, as the camera navigates the sample \citep{odwyer_summary_2000}. 
+This alleviates the need of having remarkably level sample alignment ($d_1 - d_2 < 0.5mm$ in Figure \ref{fig:realistic_levelling}).
+
+\[\boldsymbol{\delta} = 
+\begin{bmatrix}
+  0.5 \\
+  0.4 \\
+  \vdots \\
+  -0.4 \\
+  -0.5 \\
+  \end{bmatrix}\]
 \[
@@
 \boldsymbol{Z_{\text{samples}}} = 
-Z_0 + 
-\begin{bmatrix}
--0.5 \\
--0.4 \\
-\vdots \\
-0.4 \\
-0.5 \\
-\end{bmatrix}
-\] % unsure if this makes sense
-
-Our focusing procedure begins at 0.5 mm above $Z_0$ and finishes with the last image at 0.5 mm below $Z_0$.
-The 11 images then have their normalized variance, $NV$, calculated in a separate thread to measure the image sharpness \citep{sampat_extensive_2014}.
-The image with the maximum score is saved and can be considered the most in-focus image, while the rest are deleted.
-
-\[
-  i_{\max} = \arg\max_{i} NV(image(\boldsymbol{{Z_\text{samples}})})
+Z_0 + \boldsymbol{\delta} + \alpha_k
 \]
-
-The second stage of our time-efficient focusing procedure is using a feedback loop to update the initial $Z_0$ at each image location. 
-Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm, $d_1 - d_2 < 0.5 mm $ (Figure \ref{fig:realistic_levelling}). 
-Rather than adjusting this range, a greater alignment error can be managed by controlling the center of the range, $Z_{0,k}$, for $(X_k,Y_k)$. 
-The likelihood of an adjacent %emw9Mar:  we need a noun here instead of just $(X_{k+1}, Y_{k+1})$ ... maybe we need a term for this?
-$(X_{k+1}, Y_{k+1})$ containing an in-focus image is highest when the current $i_{{\max,k}}$ is at the middle index of $\boldsymbol{Z_{samples}}$. 
-A PID control algorithm with a process variable of $i_{max}$ and control variable of $Z_{0,k}$ allows a negative feedback loop to improve focusing across the entire sample \citep{odwyer_summary_2000}.
-Instead of forcing a remarkably level sample alignment of $(d_1 - d_2) < 0.5mm$ (Figure \ref{fig:realistic_levelling}), the system can handle $(Z_{k} - Z_{k+1} < 0.5mm)$ which is readily achievable.
-%After the capturing overlapping images across the entire surface area, the images are ready to be stitched.
 
@@
 Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to decrease digitization time and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.
-This is a variation of the standard approach used in 3D printing and Computer Numerical Control (CNC) machine control.
+This is a variation of the standard approach used in 3D printing and Computer Numerical Control (CNC) machining.
 
@@
 
-First, the camera is moved to the $(X_0, Y_0, Z_0)$ of the new core. 
+First, the camera is moved to the $(X_0, Y_0, Z_0)$ (Table \ref{tab:math_notation}) of the new core. 
 From there, the camera is moved across a window, at constant velocity. %emw9Mar:  w is introduced nicely below. Need to try to follow this format when possible. 
@@
 Images are captured at equal distance intervals analogously to the image focusing procedure. 
-Once again, the image with the highest normalized variance score is considered the best image and its location is used as the realigned $X_0^\prime$.
+Once again, the image with the highest normalized variance score is considered the best image and its location is used as the realigned $X_0$.
 This procedure drastically improves digitization quality of batches of cores.
@@
 % Sandy got back to me! :)
-We tested a scanned cookie from TIM with the R-CNN presented by \citeauthor{polacek_automation_2023} on two servers.
+We tested a scanned cookie of size (HEIGHT)X(WIDTH) (WAITING FOR SANDY) pixels from TIM with the R-CNN presented by \citeauthor{polacek_automation_2023} on two servers.
 The first being the high-performance UBC ARC Sockeye which completed the task in 15 minutes.
@@
 \subsection{Supplemental Information}
+\renewcommand{\thetable}{S\arabic{table}}
 
-\begin{tabular}{ |p{3cm}||p{12cm}|  }
-  \hline
-  \multicolumn{2}{|c|}{Math Notation} \\
-  \hline
-  Terms & Definition\\
-  \hline
-  $X_0, Y_0, Z_0$   &The initial cartesian coordinates of the camera when the sample is saved by the user. \\
-  $(X_k, Y_k, Z_k)$ & A unique set of $X$, $Y$, and $Z$ coordinates for the center point of an individual image $k$. \\
-  $(X_{k+1}, Y_{k+1})$ & The $X$ and $Y$ coordinates of an image that neighbors image $k$ at $(X_k, Y_k)$. \\
-  $Z_{0,i}$ & The starting $Z$ value to move to when moving to the adjacent image. This value is also the center value of $\boldsymbol{Z_{\text{samples}}}$ \\
-  $NV()$ & Normalized variance 'function.' \\
-  $Z_{focused}$ & The $Z$ value of the most in focus image in $\boldsymbol{Z_{\text{samples}}}$\\
-  $\boldsymbol{Z_{\text{samples}}}$ & A vector containing the $Z$ coordinate for multiple images at the same $X_k, Y_k$. \\
-  $i_{\max,k}$ & The index of the image with the largest normalized variance at the same $(X_k, Y_k)$. \\
-  $w$ & The width of the window to traverse when trying to center a core. \\
-  $image()$ & The image at a corresponding $Z$ value from $\boldsymbol{Z_{\text{samples}}}$ \\
-  $d_1$ & The relative distance from the camera to one edge of a levelled sample. \\
-  $d_2$ & The relative distance from the camera to the opposite edge of a levelled sample. \\
-
-  \hline
- \end{tabular}
+\include{notation_glossary_table}
 

### 2025-04-03 docs/writing/notation_glossary_table.tex
Changing the phrasing of image focusing and updating some of the math notation. Doubling down on the k

@@
     \label{tab:math_notation}
-    \begin{tabular}{ |p{3cm}||p{12cm}|  }
+    \begin{tabular}{ |p{3.5cm}||p{11.5cm}|  }
       \hline
@@
       $X_0, Y_0, Z_0$   &The initial cartesian coordinates of the camera when the sample is saved by the user. \\
-      $(X_k, Y_k, Z_k)$ & A unique set of $X$, $Y$, and $Z$ coordinates for the center point of an individual image $k$. \\
+      $Z_k$ & The center height of $\boldsymbol{Z_k}$ which was the best approximation of $Z_{focused,k}$ from the previous image.\\ 
+      $(X_k, Y_k, Z_{focused,k})$ & A unique set of $X$, $Y$, and $Z$ coordinates for a unique image of the sample, $k$. \\
       $(X_{k+1}, Y_{k+1})$ & The $X$ and $Y$ coordinates of an image that neighbors image $k$ at $(X_k, Y_k)$. \\
-      $\boldsymbol{\delta}$ & Equally spaced heights, relative to $Z_0$, to capture images at and use to find a sharp image. \\
-      $\alpha_k$ & The response variable of the PID control algorithm which adjusts the initial height of an image to gradually align $Z_{focused}$ to be at the center index of $\boldsymbol{Z_{samples}}$.\\
-      $NV()$ & Normalized variance 'function.' \\
-      $Z_{focused}$ & The $Z$ value of the most in-focus image in $\boldsymbol{Z_{\text{samples}}}$. \\
-      $\boldsymbol{Z_{\text{samples}}}$ & A vector containing the $Z$ coordinate for multiple images at the same $X_k, Y_k$. \\
-      $i_{\max,k}$ & The index of the image with the largest normalized variance at the same $(X_k, Y_k)$. \[
-      i_{\max} = \arg\max_{i} NV(image(\boldsymbol{{Z_\text{samples}})})
-      \] \\
+      $\boldsymbol{\delta}$ & Equally spaced heights, relative to $Z_k$, to capture images at and use to find a sharp image. \\
+      $\alpha_k$ & The response variable of the PID control algorithm which adjusts the initial height of an image to gradually align $Z_{focused,k}$ to be at the center index of $\boldsymbol{Z_k}$.\\
+      $Z_{focused,k}$ & The $Z$ value of the most in-focus image in $\boldsymbol{Z_k}$. \\
+      $Z_{focused,0}$ & The $Z$ value of the center coordinate, and first image of the sample. \\
+      $\boldsymbol{Z_k}$ & A vector containing the $Z$ coordinates for multiple images at the same $(X_k, Y_k)$. \\
+      $\boldsymbol{Z_0}$ & A vector containing the $Z$ coordinates for multiple images at $(X_0, Y_0)$. \\
       $w$ & The width of the window to traverse when trying to center a core. \\
-      $image()$ & The image at a corresponding $Z$ value from $\boldsymbol{Z_{\text{samples}}}$. \\
-      $d_1$ & The relative distance from the camera to one edge of a leveled sample. \\
-      $d_2$ & The relative distance from the camera to the opposite edge of a leveled sample. \\
+      $d_1$ & The relative distance from the camera to one edge of a leveled sample (Figure \ref{fig:sample_levelling}).\\
+      $d_2$ & The relative distance from the camera to the opposite edge of a leveled sample (Figure \ref{fig:sample_levelling}). \\
       \hline

### 2025-04-03 docs/writing/paper.tex
Changing the phrasing of image focusing and updating some of the math notation. Doubling down on the k

@@
 One additional parameter that needs to be defined is the height and width of the field of view of one image from the camera.
-We adjusted the focal length of the camera using the zoom ring to have a constant field of view of approximately 3 mm x 5 mm when in focus.  
+We adjusted the focal length of the camera using the zoom ring to have a constant field of view of approximately 3 mm x 5 mm when in-focus.  
 The sample height and width, sample identifiers, sample type, sample center coordinate, and the height and width of an image provides sufficient information to digitize a sample.
@@
 The goal of this traversal is to obtain in-focus images that have a region of overlap with each of its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
-Digitizing cores can be done without needing to traverse in both the X and Y axis. 
-The optimal scanning path for a core is to align the length of cores to the Y axis and only take one column of images.
-Spanning the surface area of a cookie requires movement in both the X and Y axis.
+The coordinates of each image-to-be-captured are derived from the sample's center coordinate, the height and width of the sample, the height and width of an image frame, and the user provided percentage overlap between images.
+Our design can capture cores by defining the sample width to be zero and the height of the core to be used.
+So long as the cores are aligned to be parallel to the Y axis, this minimizes imaging time as only one column of images is captured.
+Digitizing cookies requires a rectangular grid of images to be taken in both the X and Y axes. 
+The user provides enough information to find the X and Y coordinates of each image, but the Z coordinate must be found with an algorithmic focusing procedure.
 
@@
 
-Centering the design of TIM around a fixed focus camera and lens meant we needed to focus the images with a software procedure.
-The only controllable parameter to focus a fixed focus lens is by changing the distance of the lens relative to the sample. 
-With TIM's microscope lens, the depth of field of the image is sensitive enough that sub-millimeter heights can move an image out of focus.
-Solving this in a time-efficient manner involves two stages. We provide a complete glossary of notation used to describe the methods, see Table \ref{tab:math_notation}.
-The first stage takes advantage of the requirement to navigate and focus the camera to the center of the sample.
-This allows the initial $Z_0$ value, captured when adding a sample, to be an informed guess as to what value would have an in-focus image across the entire surface. 
-From the first coordinate captured $(X_0, Y_0, Z_0)$, multiple images are taken at different heights (saved as a vector $\boldsymbol{Z_{\text{samples}}}$).
-%emw9Mar: I think we need to be a little more careful in defining the math as we introduce each new term.
-One way to start on this would be to make a supplemental table with ALL the math we introduce and a brief description of it.
-If you can work that up then we could meet and fix this quickly I suspect.
-The difference in height between each image are equally spaced, and are described by the relative distance to $Z_0$ in the vector $\boldsymbol{\delta}$. 
-We found that it is important to have a distance of no more than 0.1 mm between each image in the stack to maximize the likelihood of an in-focus image at height $Z_{focused}$.
-With a sufficiently wide height range of 1 mm, TIM needs to take 11 images at different heights for comparison.
+With TIM's microscope lens, the depth of field of the image is sensitive enough that sub-millimeter heights can move an image out-of-focus.
+This means that the height collected when adding a sample, $Z_0$, does not produce in-focus images throughout the entire surface area of the sample.
+To capture in in-focus image at height $Z_{focused,k}$, for each unique image coordinate $(X_k, Y_k)$, we implemented a time conscious software procedure. 
+Our procedure involves two stages and starts at the known center coordinate of the sample (we provide a complete glossary of notation used to describe the methods, see Table \ref{tab:math_notation}).
+The first stage takes advantage of the requirement that the sample is in focus when adding it to be scanned.
+Starting at the center coordinate allows the initial $Z_0$ value, captured when adding a sample, to find $Z_{focused,0}$. 
+After finding an in-focus image, TIM navigates to the next neighboring image coordinate to find it's in-focus image, using the previous neighbor's in-focus height as an initial approximation.
+
+Finding an in-focus image is completed by comparing the image sharpness of a set of images at varying heights but the same $(X_k, Y_k)$ (saved as a vector $\boldsymbol{Z_k}$). 
+The difference in height between each image are equally spaced, and are described by the relative distance to the starting height, $Z_k$, in the vector $\boldsymbol{\delta}$. 
+We found that it is important to have a distance of no more than 0.1 mm between each image in the stack of images to maximize the likelihood of an in-focus image at height $Z_{focused,k}$.
+A height range of 1 mm provides a sufficient distribution of image sharpness.
+TIM needs to take 11 images at different heights to navigate this height range with the given step size between each image.
 The number of images taken is an editable parameter in the configuration file and can be increased, along with the height range, if in-focus images are not being obtained.
 
-Our focusing procedure begins at 0.5 mm above $Z_0$ and finishes with the last image at 0.5 mm below $Z_0$.
+Our focusing procedure begins at 0.5 mm above $Z_k$ and finishes with the last image at 0.5 mm below $Z_k$.
 The 11 images then have their normalized variance calculated in a separate thread to measure the image sharpness \citep{sampat_extensive_2014}.
 The image with the maximum score is saved and can be considered the most in-focus image, while the rest are deleted. 
+Now $Z_{focused,k}$ has been found for $(X_k, Y_k)$ and the camera can now be moved to the next image coordinates $(X_{k+1}, Y_{k+1})$ and repeat the process to find a new in-focus height.
 
-The second stage of our time-efficient focusing procedure is uses a feedback loop to update the initial $Z_0$ at each image location $(X_k, Y_k, Z_k)$. 
+The second stage of our time-efficient focusing procedure uses a feedback loop to update the initial $Z_k$ at each image location $(X_k, Y_k, Z_k)$. 
 Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm ($d_1 - d_2 < 0.5 mm $ in Figure \ref{fig:realistic_levelling}). 
-Rather than increasing this height range, a greater alignment error can be managed by adjusting the starting image height until $Z_{focused}$ is at the center index of $\boldsymbol{Z_{samples}}$. 
-The likelihood of an adjacent target
-$(X_{k+1}, Y_{k+1})$ containing an in-focus image is highest in this scenario. 
-A negative feedback PID control algorithm is used to adjust the initial height of each image, with $\alpha_k$, as the camera navigates the sample \citep{odwyer_summary_2000}. 
+Rather than increasing this height range, a greater alignment error can be managed by adjusting the starting image height until $Z_{focused,k}$ is at the center index of $\boldsymbol{Z_{samples,k}}$. 
+The likelihood of an adjacent image, at the coordinate $(X_{k+1}, Y_{k+1})$, containing an in-focus image is highest in this scenario. 
+A negative feedback PID control algorithm is used to adjust the initial height of each image, with $\alpha_k$, whenever the camera moves in the X or Y dimensions \citep{odwyer_summary_2000}. 
 This alleviates the need of having remarkably level sample alignment ($d_1 - d_2 < 0.5mm$ in Figure \ref{fig:realistic_levelling}).
-
+\[
+Z_k = \begin{cases}
+  Z_0 & \text{if } k = 0,\\
+  Z_{k-1} + \alpha_k \\
+  \end{cases}
+\]
 \[\boldsymbol{\delta} = 
@@
 \[
-Z_{focused} \in
-\boldsymbol{Z_{\text{samples}}} = 
-Z_0 + \boldsymbol{\delta} + \alpha_k
+Z_{focused,k} \in
+\boldsymbol{Z_k} = 
+Z_k + \boldsymbol{\delta}
 \]
 
-To reduce motion blur in the images, the range of $\boldsymbol{Z_{\text{samples}}}$ is traversed at constant velocity and images are captured without stopping. 
+To reduce motion blur in the images, the range of $\boldsymbol{Z_k}$ is traversed at constant velocity and images are captured without stopping. 
 Decoupling the auxiliary camera control from the G-code commands to control the machine has shown to decrease digitization time and decrease vibrational effects from acceleration and deceleration \citep{propst_time_2025}.

### 2025-04-04 docs/writing/paper.tex
Adding some context to the introduction of automatic tree ring detection

@@
 % Sandy got back to me! :)
+In it's current state, TIM only provides scanned images and requires users to import them into other tools to measure tree-ring widths or count vessels.
+A great addition to TIM would be to use computer vision models to automatically detect tree-ring widths or other metrics while capturing images.  
 We tested a scanned cookie of size (HEIGHT)X(WIDTH) (WAITING FOR SANDY) pixels from TIM with the R-CNN presented by \citeauthor{polacek_automation_2023} on two servers.
 The first being the high-performance UBC ARC Sockeye which completed the task in 15 minutes.
-As many do not have access to such powerful computing, we also tested
-the same image on an RTX A6000 GPU server---taking 3.5 hours to complete the task for approximately \$3.50 USD. 
-While the results of automated tree rings are promising, further testing is necessary to test the feasibility of running this model on an edge computer or developing integrations for TIM with a remote server.
+As many do not have access to such powerful computing, we also tested the same image on an RTX A6000 GPU server---taking 3.5 hours to complete the task for approximately \$3.50 USD. 
+Although the Jetson is not as powerful as these servers, running these models in parallel to the image capturing algorithms could be a great extension to further reduce manual data annotation of tree-rings.
+Further development would be required to run the deep learning models on the Jetson itself or with a web server to offload the heavier computation.
 

### 2025-04-11 code/gantry.py
Removing command lock release logging

@@ -197,7 +197,7 @@ class Gantry:
             grbl_out_list.append(grbl_out)
 
         self.send_command_lock.release()
-        log.info("Release send command lock. {}".format(time.time()))
+        # log.info("Release send command lock. {}".format(time.time()))
         return grbl_out_list
 
     def grbl_handshake(self, grbl_out:str):

### 2025-04-14 docs/writing/paper.tex
Edits from Lizzie! See Supp and main text

--- 2025-04-03 docs/writing/paper.tex
@@
 \title{TIM: Tree Imaging Machine for low-cost digital images of tree cores and cookies} 
-\author{Your Name}
+\author{Adam Fong,$^{1}$ Chloe Curry,$^{1}$ Sandy Zhang$^{1,2}$ and E.M. Wolkovich$^{1}$} %emw14Apr2025: You should ask Chloe and Sandy who they want their names displayed (middle initial etc.)
 \date{\today}
+
+\noindent $^{1}$Forest and Conservation Sciences, University of British Columbia, Vancouver, BC, Canada\\
+$^{2}$Sandy's current affiliation\\ %emw14Apr2025 -- I am not sure you need to add this, usually the affiliation is where you did most of the work; it can be multiple if you use a significant # of employed at another job working on it (basically, the other university PAID your time to do this work)
 
@@
 % Getting rid of open-source as CaptuRING is cheap but GigaPixel is expensive and not open source
-The Tree Imaging Machine (TIM) is a do-it-yourself scanning tool made to digitize tree cookies and cores that address these limitations.
+To address these limitations, we designed the Tree Imaging Machine (TIM)---a do-it-yourself scanning tool to digitize tree cookies and cores.
 TIM takes partially overlapping microscopic images of samples and stitches the individual images together to form
 a mosaic, which can be zoomed in to visualize features on the scale of 0.01 mm.
-With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for the competitive price of less than \$3 000 USD.
+With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for less than \$3 000 USD.
+%emw14Apr2025 -- we're not good arbitrators of what is a competitive price
 We designed TIM to have a large working plane to allow digitizing cores over 50 cm and complete cookies.
 Operators can prepare sample batches of multiple cores and cookies, letting the machine work until the queue is finished.
+%emw14Apr2025 -- add \href command below to make the URLs work 
 All of the 3D printed parts are available for download on the [NIH 3D printing repository](https://doi.org/10.60705/3DPX/21561.3) and the software is  % is this best practice for including links? 
@@
 Dated tree rings from dendrochronology have been critical to
-reconstructing past climates of regional and local environments, providing the most accurate, precise, and reliable dating among alternatives \citep{mann_northern_1999}, and thus fundamnentally informing  our understanding of climate change today \citep{fritts_dendroclimatology_1971,williams_using_2010,guibal_dendrochronology_2021,sheppard_dendroclimatology_2010}.
-They are also increasingly used in ecology to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017} and have implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}.
+reconstructing past climates of regional and local environments, providing the most accurate, precise, and reliable dating among alternatives \citep{mann_northern_1999}, and thus informing  our understanding of climate change today \citep{fritts_dendroclimatology_1971,williams_using_2010,guibal_dendrochronology_2021,sheppard_dendroclimatology_2010}.
+Tree rings are also increasingly used in ecology to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017} and have implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}.
 These fields have leveraged two basic types of tree sampling techniques to capture the variation of tree rings: cores, which are cylindrical samples collected using an incremental borer, and `cookies,' which are entire cross sections \citep[and thus allow measuring as many radii from the sample as desired,][]{speer_fundamentals_2010}.
@@
 Image analysis of tree rings was later implemented to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in the 1980s and 1990s to digitize tree ring samples using optical scanning \citep{mcmillin_application_1982} \citep{guay_new_1992}.
-Optical scanners are still readily available for purchase, using a top of the line scanner, like the Epson Perfection V850 Pro produces maximum resolutions of 6 400 dpi and scan an area of up to 21.6 cm x 29.7 cm.
+Optical scanners are still readily available for purchase, using a top of the line scanner (e.g., Epson Perfection V850 Pro) produces a maximum resolutions of 6 400 dpi and scan area of up to 21.6 cm x 29.7 cm.
 More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach---using an image-capturing system.
@@
 Additionally, both of these systems are designed to capture images along a singular axis as the entire width of a core fits within a single image's field of view.
-This assumption fundamentally restricts the digitization to be limited only to cores, as cookies have a two dimensional scanning surface. %emw9Mar: later 2/3 of paragraph very nice.
+This assumption fundamentally restricts the digitization only to cores, as cookies have a two dimensional scanning surface. %emw9Mar: later 2/3 of paragraph very nice.
 
-We made TIM to combine the open-source and open-hardware philosophies of CaptuRING with the functionality of Gigapixel.
+To address some of the current limitations, while building upon the open-source and open-hardware philosophies of CaptuRING with the functionality of Gigapixel, we designed the Tree Imaging Machine (TIM).
 We added support to digitize both cookies and cores, increased the maximum sample length, perform image stitching directly without the need for manual file transfer, and capture batches of samples sequentially---all while minimizing cost.
-Our design requires the use of 3D printed components and common hand tools such as allen keys and wrenches. No specialized equipment or power tools are required beyond this.
+Our design requires the use of 3D printed components and common hand tools such as Allen keys and wrenches.
+No specialized equipment or power tools are required beyond this.
+%emw14Apr2025 I think it's Allen wrenches and Hex keys --- check and update
 Excluding 3D printed parts, the total cost of the machine is approximately \$2 200 USD \citep[compared to the \$70 000 USD of the Gigapixel,][]{griffin_gigapixel_2021}.
@@
 Professional camera setups are comparable in price to the entire cost of TIM.
-Additional savings can also be had when factoring in the cost of a professional stitching software license such as PTGui, the suggested software to use with CaptuRING.
+These savings combined with not needing to purchase a professional stitching software license (such as PTGui, the suggested software to use with CaptuRING) makes TIM a more economical option than most products current available and its open-source, open-hardware design makes it easy to adapt and adjust.
+%emw14Apr2025 -- add citation for PTGui?
+Check if their documentation suggests one....
 
@@
   \includegraphics[height=0.5\linewidth]{../content/tina.jpg}
-  \caption{Complete assembly of TIM. Included in the work plane a few cookies ready to be scanned.}
+  \caption{Complete assembly of TIM, shown with a few cookies ready to be scanned (in the work plane).}
   \label{fig:tim_assembled}
@@
 For the camera, we chose a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens and connected this to the gantry machine using custom adapters.
-For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson) which also connects to the camera via a MIPI cable.
+For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson), which also connects to the camera via a MIPI cable.
+%emw14Apr2025 -- I think we should avoid referring to products just by their brandname ...
+consider replacing Jetson with `Jetson edge computer' throughout.
 Controlling the machine is done by launching a Python program and interacting with mouse button clicks on the GUI.
 
-By choosing this combination, we were able to reduce the weight and cost of the camera significantly, which reduced design challenges in stabilizing movement of a heavier DSLR and allowed us to invest more in a powerful computer that can handle intensive image processing.
+By choosing this combination, we were able to reduce the weight and cost of the camera significantly.
+This minimized the design challenges of stabilizing the movement of a heavier DSLR and allowed us to invest more in a powerful computer that can handle intensive image processing.
 The Jetson is an edge computer which drives a monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for control, and performs calculations to stitch individual images into one mosaic.
-Despite the weight reductions, torsion on the gantry arm still resulted in a non-zero torsional deflection. A torsion correcting adapter was designed to counteract this rotation and level the lens.
+Despite the weight reductions, torsion on the gantry arm still resulted in a non-zero torsional deflection.
+We therefore designed a torsion correcting adapter to counteract this rotation and level the lens.
 
@@
   \end{subfigure}
-  \caption{Side view of the camera and sample on top of a leveling table.
+  \caption{Side view of the camera (top, in black) and sample on top of a leveling table.
 The ideal sample leveling shows a uniform distance, $d$, at all $XY$ coordinates on the sample.
-This is impossible to achieve in reality, the true sample leveling has a non uniform distance at unique $XY$.}
+This is impossible to achieve in reality, the true sample leveling has a non-uniform distance at unique $XY$.}
   \label{fig:sample_levelling}
@@
 Similarly to cookies, the scanning surface should be as close to parallel to the XY-plane as possible, but generally does not need to be leveled using the leveling tables.
-Cores also need to be aligned to be parallel to the $Y$ axis to be digitized, using squaring tool or similar to help with alignment.
+Cores also need to be aligned to be parallel to the $Y$ axis to be digitized, using a squaring tool (or similar) to help with alignment.
 Any warping in core mounts needs to be counteracted with the use of trigger clamps.
@@
 To achieve this with a fixed focus camera, the samples must be nearly orthogonal to the camera lens. 
-Once the sample is level, the operator must follow a procedure to inform TIM of where the sample is, and the it's size. 
-To do this, the operator interacts with the machine using buttons on the GUI---navigating and focusing the camera on the center of the sample. 
-The sample height and width in the X and Y dimensions are then entered in the GUI to be saved.
-Finally, the user must click the 'Add Sample' button which provides more fields to add sample identifiers and save the center coordinate.
+Once the sample is level, the operator follows a procedure to inform TIM of where the sample is, and the its size using buttons on the GUI to navigate and focus the camera on the center of the sample.
+The sample height and width in the $X$ and $Y$ dimensions are then entered in the GUI to be saved.
+Finally, the user must click the `Add Sample' button which provides more fields to add sample identifiers and save the center coordinate.
 This procedure can be repeated to create a queue of samples to digitize as a batch.
@@
       \label{SUBFIGURE LABEL 3}
-  \end{subfigure} %emw9Mar: Avoid 'this figure shows ...' or anything similar. Also, check my edits here closely!
+  \end{subfigure} %emw14Apr2025: check my edits here please (capture region as a phrase okay?)!
   \caption{An example imaging process of a cookie sample.
-Here, the height and width are enclosed the entire surface area of the cookie and TIM begins by generating a grid of target coordinates to systematically capture a grid of continuous and partially overlapping images (a).
+Here, the height and width of the capture region enclose the entire surface area of the cookie.
+TIM begins by generating a grid of target coordinates to systematically capture a grid of continuous and partially overlapping images (a).
 TIM takes its first image at the center coordinate of the sample, then the rectangular field of view from the camera moves to adjacent target coordinates until it reaches the left boundary---capturing an image for each target coordinate (b).
@@
 
-The goal of this traversal is to obtain in-focus images that have a region of overlap with each of its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
+%emw14Apr2025: Careful with its (versus it's -- it is)!
+The goal of this sample transverse is to obtain in-focus images that have a region of overlap with each of its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
 The coordinates of each image-to-be-captured are derived from the sample's center coordinate, the height and width of the sample, the height and width of an image frame, and the user provided percentage overlap between images.
@@
 With TIM's microscope lens, the depth of field of the image is sensitive enough that sub-millimeter heights can move an image out-of-focus.
-This means that the height collected when adding a sample, $Z_0$, does not produce in-focus images throughout the entire surface area of the sample.
-To capture in in-focus image at height $Z_{focused,k}$, for each unique image coordinate $(X_k, Y_k)$, we implemented a time conscious software procedure. 
-Our procedure involves two stages and starts at the known center coordinate of the sample (we provide a complete glossary of notation used to describe the methods, see Table \ref{tab:math_notation}).
-The first stage takes advantage of the requirement that the sample is in focus when adding it to be scanned.
+%emw14Apr2025: $Z_0$ is 'height' but in the notation table it's a vector ... I think maybe the table is wrong? 
+This means that the height collected when adding a sample ($Z_0$, we provide a complete glossary of notation used to describe the methods, see Table \ref{tab:math_notation}) does not produce in-focus images throughout the entire surface area of the sample.
+To capture an in-focus image at height $Z_{focused,k}$, for each unique image coordinate $(X_k, Y_k)$, we implemented a time-conscious software procedure. 
+Our procedure involves two stages and starts at the known center coordinate of the sample.
+The first stage takes advantage of the requirement that the sample is in-focus when adding it to be scanned.
 Starting at the center coordinate allows the initial $Z_0$ value, captured when adding a sample, to find $Z_{focused,0}$. 
-After finding an in-focus image, TIM navigates to the next neighboring image coordinate to find it's in-focus image, using the previous neighbor's in-focus height as an initial approximation.
+After finding an in-focus image, TIM navigates to the next neighboring image coordinate to find its in-focus image, using the previous neighbor's in-focus height as an initial approximation.
 
-Finding an in-focus image is completed by comparing the image sharpness of a set of images at varying heights but the same $(X_k, Y_k)$ (saved as a vector $\boldsymbol{Z_k}$). 
-The difference in height between each image are equally spaced, and are described by the relative distance to the starting height, $Z_k$, in the vector $\boldsymbol{\delta}$. 
+Finding an in-focus image is completed by comparing the image sharpness of a set of images at varying heights (a vector $\boldsymbol{Z_k}$) but the same coordinate $(X_k, Y_k)$. 
+The differences in height between each image are equally spaced, and are described by the relative distance to the starting height, $Z_k$, in the vector $\boldsymbol{\delta}$. 
 We found that it is important to have a distance of no more than 0.1 mm between each image in the stack of images to maximize the likelihood of an in-focus image at height $Z_{focused,k}$.
@@
 TIM needs to take 11 images at different heights to navigate this height range with the given step size between each image.
-The number of images taken is an editable parameter in the configuration file and can be increased, along with the height range, if in-focus images are not being obtained.
+The number of images taken is an editable parameter in the configuration file and can be increased, along with the height range, if TIM Is not obtaining in-focus images.
 
@@
 The likelihood of an adjacent image, at the coordinate $(X_{k+1}, Y_{k+1})$, containing an in-focus image is highest in this scenario. 
-A negative feedback PID control algorithm is used to adjust the initial height of each image, with $\alpha_k$, whenever the camera moves in the X or Y dimensions \citep{odwyer_summary_2000}. 
-This alleviates the need of having remarkably level sample alignment ($d_1 - d_2 < 0.5mm$ in Figure \ref{fig:realistic_levelling}).
+A negative feedback  %emw14Apr2025: check!
+proportional-integral-derivative (PID_ control algorithm is used to adjust the initial height of each image, with $\alpha_k$, whenever the camera moves in the X or Y dimensions \citep{odwyer_summary_2000}.
+This alleviates the need of having remarkably level sample alignment ($d_1 - d_2 < 0.5\text{mm}$ in Figure \ref{fig:realistic_levelling}).
+%emw14Apr2025 -- check I did not mess up math notation here, if not fix above also so that mm is not italics
 \[
@@
 
-First, the camera is moved to the $(X_0, Y_0, Z_0)$ (Table \ref{tab:math_notation}) of the new core. 
+First, the camera is moved to the initial coordinates of the camera when the sample is first saved by the user $(X_0, Y_0, Z_0$, Table \ref{tab:math_notation}) of the new core. 
 From there, the camera is moved across a window, at constant velocity. %emw9Mar:  w is introduced nicely below. Need to try to follow this format when possible. 
-The width of the window, $w$, in millimeters is defined in the machine configuration file and is editable. %emw9Mar:  we need a noun below ...
-$(X_0 - 1/2*{w}, Y_0, Z_0)$ to $(X_0 + 1/2*{w}, Y_0, Z_0)$. 
+The width of the window, $w$, in millimeters is defined in the machine configuration file and is editable %emw9Mar:  we need a noun below ...
+(from $X_0 - 1/2*{w}, Y_0, Z_0$ to $X_0 + 1/2*{w}, Y_0, Z_0)$. 
 Images are captured at equal distance intervals analogously to the image focusing procedure. 
@@
 The default implementation of the package produces high quality stitches but is memory inefficient. %is this wording correct?, I'm trying to kindly say it was memory inefficient
-We implemented a few key memory conscious changes to the package and were able to run it on the Jetson without a problem.
+We implemented a few key memory-conscious changes to the package and were able to run it on the Jetson without a problem.
 
@@
 
+%emw14Apr2025 -- this figure is not critical to the paper, so I would move it to supp and just reference it here (Fig S1 etc.)
 \begin{figure}
@@
 We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ.
+%emw14Apr2025 cite ImageJ when first used in paper, see options here https://imagej.net/ij/docs/faqs.html
 %emw9Mar: Any citations for developing repos to share tree ring images? Or just more on building better open shared data for tree rings? 
@@
 Publicly available tree-ring datasets can be found on multiple online repositories such as TreeSource's Tree-Rings Database for Canadian sites or the International Tree-Ring Data Bank (ITRDB) for worldwide data \citep{girardin_national_2021} \citep{grissino-mayer_international_1997}.
-Both of these databases offer metadata rich data on measured tree-rings but do not store the source images of samples.
+Both of these databases offer metadata-rich data on measured tree-rings but do not store the source images of samples.
 One opportunity for this development could come as more universal and large-scale open-data storage of tree ring images come online. 
@@
 % Sandy got back to me! :)
-We tested a scanned cookie of size (HEIGHT)X(WIDTH) (WAITING FOR SANDY) pixels from TIM with the R-CNN presented by \citeauthor{polacek_automation_2023} on two servers.
-The first being the high-performance UBC ARC Sockeye which completed the task in 15 minutes.
-As many do not have access to such powerful computing, we also tested
-the same image on an RTX A6000 GPU server---taking 3.5 hours to complete the task for approximately \$3.50 USD. 
-While the results of automated tree rings are promising, further testing is necessary to test the feasibility of running this model on an edge computer or developing integrations for TIM with a remote server.
+%emw14Apr2025  -- finalize the info below before sending outside the lab for review
+In its current state, TIM only provides scanned images and requires users to import them into other tools for additional visual analyses, such as measuring tree-ring widths or counting vessels.
+A useful addition to TIM would be to automatically detect tree-ring widths or other metrics while capturing images using computer vision models.
+We expect this may be possible as these methods improve in efficiency.
+Currently, we found that tree-ring detection by a recent machine learning algorithm  \citeauthor[R-CNN presented by][]{polacek_automation_2023} for a scanned cookie of size (HEIGHT)X(WIDTH) (WAITING FOR SANDY) pixels from TIM  took 15 minutes on a high-performance cluster (UBC ARC Sockeye) and 3.5 hours on a smaller server (RTX A6000 GPU server, which currently costs approximately \$3.50 USD for this).
+Although the Jetson is not as powerful as these servers, running these models in parallel to the image capturing algorithms would further reduce manual data annotation of tree-rings, and thus could help scale-up data collection.
+% Further development would be required to run the deep learning models on the Jetson itself or with a web server to offload the heavier computation.
 
-Finally, an exciting frontier for machines such as TIM is integrating additional sensors.
+An additional major frontier for machines such as TIM is integrating additional sensors.
 While much of the current imaging time of samples comes from taking a vertical stack of images to obtain one in-focus image, advanced microscopy applications have overcome similar challenges through supplemental sensors that measure the distance between the subject and lens.
@@
 By taking advantage of a common three degree of freedom cartesian machine design, powerful edge computing, and a microscope camera we were able to design a cost effective and high resolution digitization tool for wood samples.
-We significantly increased the maximum sample length and maximum resolution compared to common alternatives in the field. 
+Our design significantly increases the maximum sample length and maximum resolution compared to common alternatives in the field. 
 In addition, the machine design was intended to be readily replicated by labs with minimal engineering experience and equipment.
+
+%emw14Apr2025: Add acknowledgments here: Deirdre, Hoai Huong Nguyen Phan, Fredi, CRD, Mao etc.
 

### 2025-05-12 docs/writing/paper.tex

@@
 \usepackage{multirow}
+\usepackage{float}
 
@@
 \title{TIM: Tree Imaging Machine for low-cost digital images of tree cores and cookies} 
-\author{Adam Fong,$^{1}$ Chloe Curry,$^{1}$ Sandy Zhang$^{1,2}$ and E.M. Wolkovich$^{1}$} %emw14Apr2025: You should ask Chloe and Sandy who they want their names displayed (middle initial etc.)
+\author{Adam Fong,$^{1}$ Chloe Curry,$^{1}$ Yue Zhang$^{1, 2}$ and E.M. Wolkovich$^{1}$} %emw14Apr2025: You should ask Chloe and Sandy who they want their names displayed (middle initial etc.)
 \date{\today}
-
-\noindent $^{1}$Forest and Conservation Sciences, University of British Columbia, Vancouver, BC, Canada\\
-$^{2}$Sandy's current affiliation\\ %emw14Apr2025 -- I am not sure you need to add this, usually the affiliation is where you did most of the work; it can be multiple if you use a significant # of employed at another job working on it (basically, the other university PAID your time to do this work)
 
@@
 \maketitle
+
+\noindent $^{1}$Forest and Conservation Sciences, University of British Columbia, Vancouver, BC, Canada\\
+$^{2}$Department of Computer Science and Engineering, University of California San Diego, La Jolla, CA, USA\\
 
@@
 With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for less than \$3 000 USD.
-%emw14Apr2025 -- we're not good arbitrators of what is a competitive price
 We designed TIM to have a large working plane to allow digitizing cores over 50 cm and complete cookies.
 Operators can prepare sample batches of multiple cores and cookies, letting the machine work until the queue is finished.
-%emw14Apr2025 -- add \href command below to make the URLs work 
-All of the 3D printed parts are available for download on the [NIH 3D printing repository](https://doi.org/10.60705/3DPX/21561.3) and the software is  % is this best practice for including links? 
-open source, with instructions to recreate this tool available on [GitHub](https://github.com/temporalecologylab/TreeRings).
+
+All of the 3D printed parts are available for download on the \href{https://doi.org/10.60705/3DPX/21561.3}{NIH 3D printing repository} and the software is  % is this best practice for including links?
+open source, with instructions to recreate this tool available on \href{https://github.com/temporalecologylab/TreeRings}{GitHub}.
 TIM provides an important advance in scanning resolution, tree cookie sampling, maximum core sample size, batch digitization, and cost in a single package.
@@
 
-Image analysis of tree rings was later implemented to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in the 1980s and 1990s to digitize tree ring samples using optical scanning \citep{mcmillin_application_1982} \citep{guay_new_1992}.
+Image analysis of tree rings was later implemented to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in the 1980s and 1990s to digitize tree ring samples using optical scanning \citep{mcmillin_application_1982,guay_new_1992}.
 Optical scanners are still readily available for purchase, using a top of the line scanner (e.g., Epson Perfection V850 Pro) produces a maximum resolutions of 6 400 dpi and scan area of up to 21.6 cm x 29.7 cm.
@@
 We added support to digitize both cookies and cores, increased the maximum sample length, perform image stitching directly without the need for manual file transfer, and capture batches of samples sequentially---all while minimizing cost.
-Our design requires the use of 3D printed components and common hand tools such as Allen keys and wrenches.
-No specialized equipment or power tools are required beyond this.
-%emw14Apr2025 I think it's Allen wrenches and Hex keys --- check and update
+Our design requires the use of 3D printed components and common hand tools such as hex keys and wrenches. No specialized equipment or power tools are required beyond this. 
 Excluding 3D printed parts, the total cost of the machine is approximately \$2 200 USD \citep[compared to the \$70 000 USD of the Gigapixel,][]{griffin_gigapixel_2021}.
@@
 Professional camera setups are comparable in price to the entire cost of TIM.
-These savings combined with not needing to purchase a professional stitching software license (such as PTGui, the suggested software to use with CaptuRING) makes TIM a more economical option than most products current available and its open-source, open-hardware design makes it easy to adapt and adjust.
-%emw14Apr2025 -- add citation for PTGui?
-Check if their documentation suggests one....
+These savings combined with not needing to purchase a professional stitching software license (such as PTGui (New House Internet Services BV, Rotterdam, NL), the suggested software to use with CaptuRING) makes TIM a more economical option than most products current available and its open-source, open-hardware design makes it easy to adapt and adjust.
 
@@
 For the camera, we chose a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens and connected this to the gantry machine using custom adapters.
-For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson), which also connects to the camera via a MIPI cable.
-%emw14Apr2025 -- I think we should avoid referring to products just by their brandname ...
-consider replacing Jetson with `Jetson edge computer' throughout.
+For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson edge computer), which also connects to the camera via a MIPI cable.
 Controlling the machine is done by launching a Python program and interacting with mouse button clicks on the GUI.
@@
       \label{SUBFIGURE LABEL 3}
-  \end{subfigure} %emw14Apr2025: check my edits here please (capture region as a phrase okay?)!
+  \end{subfigure} 
   \caption{An example imaging process of a cookie sample.
@@
 
-%emw14Apr2025: Careful with its (versus it's -- it is)!
 The goal of this sample transverse is to obtain in-focus images that have a region of overlap with each of its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
@@
 With TIM's microscope lens, the depth of field of the image is sensitive enough that sub-millimeter heights can move an image out-of-focus.
-%emw14Apr2025: $Z_0$ is 'height' but in the notation table it's a vector ... I think maybe the table is wrong? 
+
 This means that the height collected when adding a sample ($Z_0$, we provide a complete glossary of notation used to describe the methods, see Table \ref{tab:math_notation}) does not produce in-focus images throughout the entire surface area of the sample.
@@
 The second stage of our time-efficient focusing procedure uses a feedback loop to update the initial $Z_k$ at each image location $(X_k, Y_k, Z_k)$. 
-Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm ($d_1 - d_2 < 0.5 mm $ in Figure \ref{fig:realistic_levelling}). 
+Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm ($d_1 - d_2 < 0.5\,\text{mm}$ in Figure \ref{fig:realistic_levelling}). 
 Rather than increasing this height range, a greater alignment error can be managed by adjusting the starting image height until $Z_{focused,k}$ is at the center index of $\boldsymbol{Z_{samples,k}}$. 
 The likelihood of an adjacent image, at the coordinate $(X_{k+1}, Y_{k+1})$, containing an in-focus image is highest in this scenario. 
-A negative feedback  %emw14Apr2025: check!
-proportional-integral-derivative (PID_ control algorithm is used to adjust the initial height of each image, with $\alpha_k$, whenever the camera moves in the X or Y dimensions \citep{odwyer_summary_2000}.
-This alleviates the need of having remarkably level sample alignment ($d_1 - d_2 < 0.5\text{mm}$ in Figure \ref{fig:realistic_levelling}).
-%emw14Apr2025 -- check I did not mess up math notation here, if not fix above also so that mm is not italics
+A negative feedback  
+proportional-integral-derivative (PID) control algorithm is used to adjust the initial height of each image, with $\alpha_k$, whenever the camera moves in the X or Y dimensions \citep{odwyer_summary_2000}.
+This alleviates the need of having remarkably level sample alignment ($d_1 - d_2 < 0.5\,\text{mm}$ in Figure \ref{fig:realistic_levelling}). 
 \[
@@
 The default implementation of the package produces high quality stitches but is memory inefficient. %is this wording correct?, I'm trying to kindly say it was memory inefficient
-We implemented a few key memory-conscious changes to the package and were able to run it on the Jetson without a problem.
+We implemented a few key memory-conscious changes to the package and were able to run it on the Jetson edge computer without a problem.
 
@@
 % MEE's guidelines have a distinct Results section from Discussion although maybe they're flexible on this? Conclusion is optional for MEE
-
-%emw14Apr2025 -- this figure is not critical to the paper, so I would move it to supp and just reference it here (Fig S1 etc.)
-\begin{figure}
-  \centering
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_whole.png}
-    \caption{Scale Bar for DPI measurements}
-    \label{fig:dpi_measurement_whole}
-  \end{subfigure}%
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_zoomed_2.png}
-    \caption{Zoomed in Scale Bar}
-    \label{fig:dpi_measurement_zoomed}
-  \end{subfigure}
-  \caption{Example of a single DPI measurement using a 0.01 mm slide scale.
-We found the DPI measured at multiple locations in the field of view vertically and horizontally  converged on the same value.}
-  \label{fig:dpi_measurements}
-\end{figure}
 
@@
 TIM's design minimizes the barriers to build an affordable, efficient and high-resolution image capture system in smaller labs without compromising scan quality, and while allowing scanning of cookies and cores.
-We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ.
-%emw14Apr2025 cite ImageJ when first used in paper, see options here https://imagej.net/ij/docs/faqs.html
+We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ \citep{schneider_nih_2012}.
 %emw9Mar: Any citations for developing repos to share tree ring images? Or just more on building better open shared data for tree rings? 
@@
 
-% Sandy got back to me! :)
-%emw14Apr2025  -- finalize the info below before sending outside the lab for review
 In its current state, TIM only provides scanned images and requires users to import them into other tools for additional visual analyses, such as measuring tree-ring widths or counting vessels.
@@
 We expect this may be possible as these methods improve in efficiency.
-Currently, we found that tree-ring detection by a recent machine learning algorithm  \citeauthor[R-CNN presented by][]{polacek_automation_2023} for a scanned cookie of size (HEIGHT)X(WIDTH) (WAITING FOR SANDY) pixels from TIM  took 15 minutes on a high-performance cluster (UBC ARC Sockeye) and 3.5 hours on a smaller server (RTX A6000 GPU server, which currently costs approximately \$3.50 USD for this).
-Although the Jetson is not as powerful as these servers, running these models in parallel to the image capturing algorithms would further reduce manual data annotation of tree-rings, and thus could help scale-up data collection.
-% Further development would be required to run the deep learning models on the Jetson itself or with a web server to offload the heavier computation.
+Currently, we found that tree-ring detection by a recent machine learning algorithm  \citeauthor[R-CNN presented by][]{polacek_automation_2023} for a scanned core (23294 x 897 pixels) took 15 minutes on a high-performance cluster (UBC ARC Sockeye) and 3.5 hours on a smaller server (RTX A6000 GPU server, which currently costs approximately \$3.50 USD for this).
+Although the Jetson edge computer is not as powerful as these servers, running these models in parallel to the image capturing algorithms would further reduce manual data annotation of tree-rings, and thus could help scale-up data collection.
 
@@
 %emw14Apr2025: Add acknowledgments here: Deirdre, Hoai Huong Nguyen Phan, Fredi, CRD, Mao etc. 
+\section{Acknowledgments}
+The authors thank Deirdre Loughnan, Hoai Huong Nguyen Phan, Frederik Baumgarten for conversations and ideas to help bring this project to fruition.
+Additionally, we thank Xiaomao Wang and Christophe Rouleau-Desrochers for their help and patience in operating TIM during development.
 
-\subsection{Supplemental Information}
+\section{Supplemental Information}
+
+{
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}  % Start supplemental figures from S1
+
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_whole.png}
+    \caption{Scale Bar for DPI measurements}
+    \label{fig:dpi_measurement_whole}
+  \end{subfigure}%
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_zoomed_2.png}
+    \caption{Zoomed in Scale Bar}
+    \label{fig:dpi_measurement_zoomed}
+  \end{subfigure}
+  \caption{Example of a single DPI measurement using a 0.01 mm slide scale.
+We found the DPI measured at multiple locations in the field of view vertically and horizontally converged on the same value.}
+  \label{fig:dpi_measurements}
+\end{figure}
+}
+
 \renewcommand{\thetable}{S\arabic{table}}
-
 \include{notation_glossary_table}

### 2025-07-01 docs/writing/paper.tex
Adding Victor's comments to git

@@
 With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for less than \$3 000 USD.
+%vvdm24Apr: briefly compare with the price of other machines?
+e.g.
+"up to 95% less expensive than other machines" (70,000 USD)
 We designed TIM to have a large working plane to allow digitizing cores over 50 cm and complete cookies.
@@
 As understanding tree growth and managing forest becomes more critical to mitigating climate change, scaling up the collection and processing of tree ring samples has become increasingly important.
+%vvdm24Apr: maybe cite Williams et al.
+2020 DOI: 10.1126/science.aaz9600?
 
@@
 This method has high precision but depends heavily on specific technicians and maintains no image record \citep{levanic_atrics_2007}.
+%vvdm24Apr: maybe add "preventing later measurement reproducibility and uncertainty quantification"
 
@@
 These savings combined with not needing to purchase a professional stitching software license (such as PTGui (New House Internet Services BV, Rotterdam, NL), the suggested software to use with CaptuRING) makes TIM a more economical option than most products current available and its open-source, open-hardware design makes it easy to adapt and adjust.
+%vvdm24Apr: adapt and adjust seem a bit redundant, perhaps "to adapt to specific requirements and improve with new...
+(something)"?
+You could also cut the sentence at "its open-source ...", to clearly separate the idea of cost reduction and the idea of open-source!
 
@@
 For the gantry machine, cartesian movement in the $X$, $Y$ and $Z$ directions is a result of two machine kits and a motor controller from OpenBuilds---the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware.
+%vvdm24Apr: I think the Figure 1 could be improve, by indicating the different components (camera, motors...)?
+I also wonder wheter a schematic diagram would be better than a photo, or maybe at least edit the photo (e.g.
+remove the wall with the fire extinguisher!
+Unless, of course, there's a high probability the machine catches fire...
+:-) )
 The machine kit includes extruded aluminum rails, carriages to slide along the rails, stepper motors to control the $X$ and $Y$ movement, and all the hardware to assemble the machine. 
@@
 For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson edge computer), which also connects to the camera via a MIPI cable.
+%vvdm24Apr: maybe add some specification for the lens?
+like the level of zoom?
 Controlling the machine is done by launching a Python program and interacting with mouse button clicks on the GUI.
@@
 We therefore designed a torsion correcting adapter to counteract this rotation and level the lens.
+%vvdm24Apr: the last two sentences are a bit hard for me to follow, though it might be because I'm far from being an hardware engineer
 
@@
 One additional parameter that needs to be defined is the height and width of the field of view of one image from the camera.
+%vvdm24Apr: this is a great feature!
 We adjusted the focal length of the camera using the zoom ring to have a constant field of view of approximately 3 mm x 5 mm when in-focus.  
@@
 The default implementation of the package produces high quality stitches but is memory inefficient. %is this wording correct?, I'm trying to kindly say it was memory inefficient
-We implemented a few key memory-conscious changes to the package and were able to run it on the Jetson edge computer without a problem.
+We implemented a few key memory-conscious changes to the package and were able to run it on the Jetson edge computer without a problem. %vvdm24Apr: maybe mention what are these changes?
 
@@
 Scanning complete cookies at maximum DPI is unrealistic as file size limits are quickly encountered with small diameter cookies.
+%vvdm24Apr: maybe you should mention the file format (TIFF or NumPy array) and the filesize constraints before?
+in the Methods
 To fit into the maximum 2.5 GB limits of a TIFF file, the maximum diameter of a cookie scanned at 30\% of maximum resolution is limited to approximately 130 mm.
@@
 Currently, we found that tree-ring detection by a recent machine learning algorithm  \citeauthor[R-CNN presented by][]{polacek_automation_2023} for a scanned core (23294 x 897 pixels) took 15 minutes on a high-performance cluster (UBC ARC Sockeye) and 3.5 hours on a smaller server (RTX A6000 GPU server, which currently costs approximately \$3.50 USD for this).
+%vvdm24Apr: what costs 3.50USD?
+Not sure I understood
 Although the Jetson edge computer is not as powerful as these servers, running these models in parallel to the image capturing algorithms would further reduce manual data annotation of tree-rings, and thus could help scale-up data collection.
@@
 %emw14Apr2025: Add acknowledgments here: Deirdre, Hoai Huong Nguyen Phan, Fredi, CRD, Mao etc. 
+%vvdm24Apr: maybe add something like: "Such an open-source device could facilitate the widespread sampling of tree rings across diverse ecosystems, offering new insights on..."
+%vvdm24Apr: and also something like "It will also support tree-ring research in countries where budget constraints may limit access to more expensive equipment..."?
+
+%vvdm24Apr: More general comment: I think it would be helpful to further justify (from an ecological perspective) why high-resolution scans are necessary?
+Not only to better distinguish the rings, but also for example to differentiate early wood and late wood...
+or something else (I'm not an expert)?
 \section{Acknowledgments}

### 2025-07-12 docs/writing/paper.tex
Editing according to Victor's comments

--- 2025-05-12 docs/writing/paper.tex
@@
 %emw24Feb: I added some text to abstract and intro that I am not 100% sure you would need for some journals, but I think you will need for MEE. 
-%emw24Feb: I was trying to add this tag '%emw24Feb:' to my comments, but got lazy and just added them below without the tag generally.
-Please delete my comments in the tex once you have dealt with them!
 %emw9Mar:
@@
 Dated tree rings from dendrochronology have been critical to
-reconstructing past climates of regional and local environments, providing the most accurate, precise, and reliable dating among alternatives \citep{mann_northern_1999}, and thus informing  our understanding of climate change today \citep{fritts_dendroclimatology_1971,williams_using_2010,guibal_dendrochronology_2021,sheppard_dendroclimatology_2010}.
+reconstructing past climates of regional and local environments, providing the most accurate, precise, and reliable dating among alternatives \citep{mann_northern_1999}, and thus informing  our understanding of climate change today \citep{fritts_dendroclimatology_1971,williams_using_2010,guibal_dendrochronology_2021,sheppard_dendroclimatology_2010,williams_large_2020}.
 Tree rings are also increasingly used in ecology to understand how plant competition affects tree growth in addition to climate \citep{buechling_climate_2017} and have implications in forestry to managing stand dynamics \citep{canham_neighborhood_2004}.
@@
 With this technique, still used widely today, a trained technician shifts a core underneath a microscope objective and pushes a button connected to a microcomputer to record a ring's location \citep{robinson_microcomputer_1980}.
-% 's okay here
-This method has high precision but depends heavily on specific technicians and maintains no image record \citep{levanic_atrics_2007}.
+This method has high precision but depends heavily on specific technicians and maintains no image record, preventing later measurement reproducibility and uncertainty quantification \citep{levanic_atrics_2007}.
 
@@
 Professional camera setups are comparable in price to the entire cost of TIM.
-These savings combined with not needing to purchase a professional stitching software license (such as PTGui (New House Internet Services BV, Rotterdam, NL), the suggested software to use with CaptuRING) makes TIM a more economical option than most products current available and its open-source, open-hardware design makes it easy to adapt and adjust.
+These savings combined with not needing to purchase a professional stitching software license (such as PTGui (New House Internet Services BV, Rotterdam, NL), the suggested software to use with CaptuRING) makes TIM a more economical option than most products current available.
+Additionally the design is open-source and open-hardware allowing for contributors to add new sensors or alter the control methods.
 
@@
 Building on top of kits allowed for quick assembly with sound instructions and saved development time. 
-For the camera, we chose a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens and connected this to the gantry machine using custom adapters.
+For the camera, we chose a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens with 150X zoom. The camera was connected to the gantry machine using custom, 3D printed, adapters.
 For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson edge computer), which also connects to the camera via a MIPI cable.
@@
 The Jetson is an edge computer which drives a monitor for a GUI, sends commands to the motor controller to move the machine, runs image processing calculations for control, and performs calculations to stitch individual images into one mosaic.
-Despite the weight reductions, torsion on the gantry arm still resulted in a non-zero torsional deflection.
-We therefore designed a torsion correcting adapter to counteract this rotation and level the lens.
 
@@
 The default implementation of the package produces high quality stitches but is memory inefficient. %is this wording correct?, I'm trying to kindly say it was memory inefficient
-We implemented a few key memory-conscious changes to the package and were able to run it on the Jetson edge computer without a problem.
+We implemented a few key memory-conscious changes such as loading only one image into RAM at a time and writing the final image to a memory-mapped NumPy file before converting to a TIFF file.
+These changes allowed for the package to run on a Jetson edge computer without a problem.
 
@@
 Scanning complete cookies at maximum DPI is unrealistic as file size limits are quickly encountered with small diameter cookies.
+
 To fit into the maximum 2.5 GB limits of a TIFF file, the maximum diameter of a cookie scanned at 30\% of maximum resolution is limited to approximately 130 mm.
@@
 We expect this may be possible as these methods improve in efficiency.
-Currently, we found that tree-ring detection by a recent machine learning algorithm  \citeauthor[R-CNN presented by][]{polacek_automation_2023} for a scanned core (23294 x 897 pixels) took 15 minutes on a high-performance cluster (UBC ARC Sockeye) and 3.5 hours on a smaller server (RTX A6000 GPU server, which currently costs approximately \$3.50 USD for this).
+Currently, we found that tree-ring detection by a recent machine learning algorithm  \citeauthor[R-CNN presented by][]{polacek_automation_2023} for a scanned core (23294 x 897 pixels) took 15 minutes on a high-performance cluster (UBC ARC Sockeye) and 3.5 hours on a smaller server (RTX A6000 GPU server).
 Although the Jetson edge computer is not as powerful as these servers, running these models in parallel to the image capturing algorithms would further reduce manual data annotation of tree-rings, and thus could help scale-up data collection.
@@
 In addition, the machine design was intended to be readily replicated by labs with minimal engineering experience and equipment. 
+Such an open-source device could facilitate the widespread sampling of tree rings across diverse ecosystems and reduce barriers for researchers with tight budgets.
 
 %emw14Apr2025: Add acknowledgments here: Deirdre, Hoai Huong Nguyen Phan, Fredi, CRD, Mao etc.
+
+%vvdm24Apr: and also something like "It will also support tree-ring research in countries where budget constraints may limit access to more expensive equipment..."?
+
+%vvdm24Apr: More general comment: I think it would be helpful to further justify (from an ecological perspective) why high-resolution scans are necessary?
+Not only to better distinguish the rings, but also for example to differentiate early wood and late wood...
+or something else (I'm not an expert)?
 \section{Acknowledgments}

### 2025-08-11 docs/writing/paper.tex
Updates to abstract in tex and 50 word thing in md

@@
 \begin{abstract}  
+1.
 Studies of tree rings from climatology to ecology have benefitted recently from new methods to image wood rapidly and efficiently, helping to scale up the number and quality of image samples.
-Current approaches are, however, limited in scanning dimensionality and can be cost-prohibitive.
+Current approaches, however, are limited in dimensionality, the amount of user-time involved per sample, and can be cost-prohibitive.
 % Getting rid of open-source as CaptuRING is cheap but GigaPixel is expensive and not open source
-To address these limitations, we designed the Tree Imaging Machine (TIM)---a do-it-yourself scanning tool to digitize tree cookies and cores.
+2.
+To address these limitations, we designed the Tree Imaging Machine (TIM)---a do-it-yourself scanning tool to digitize tree cookies and cores with open-source hardware, software and related instructions.
+% emwAug10 -- check my addition: correct?
+You could add the links here if you want
 TIM takes partially overlapping microscopic images of samples and stitches the individual images together to form 
 a mosaic, which can be zoomed in to visualize features on the scale of 0.01 mm. 
+3.
 With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for less than \$3 000 USD.
-We designed TIM to have a large working plane to allow digitizing cores over 50 cm and complete cookies.
-Operators can prepare sample batches of multiple cores and cookies, letting the machine work until the queue is finished.
-
-All of the 3D printed parts are available for download on the \href{https://doi.org/10.60705/3DPX/21561.3}{NIH 3D printing repository} and the software is  % is this best practice for including links?
-open source, with instructions to recreate this tool available on \href{https://github.com/temporalecologylab/TreeRings}{GitHub}.
+Operators can prepare sample batches of multiple samples, letting the machine work until the queue is finished.
+Given this batch approach and automatic stitching, TIM provides a major speed advance for high-resolution image capture of tree rings.
+% We designed TIM to have a large working plane to allow digitizing cores over 50 cm and complete cookies.
+% emwAug10 -- I would cut the above as suggested for MEE and put back in for more tree ring focused journals.
+I moved around something below and edited above because of that ...
+4.
 TIM provides an important advance in scanning resolution, tree cookie sampling, maximum core sample size, batch digitization, and cost in a single package.
 This promotes the digitization of larger samples and can collect data at finer resolutions in an open-source design which can be built on and improved as new technologies become available.
+Designed as completely open-source, we hope TIM will facilitate the widespread sampling of tree rings across diverse ecosystems and reduce barriers for researchers with tight budgets.
 % Archive GitHub with Zenodo?
@@
 will this be a problem?
-Or do I only ever make one 'release' https://docs.github.com/en/repositories/archiving-a-github-repository/referencing-and-citing-content
+Or do I only ever make one 'release' https://docs.github.com/en/repositories/archiving-a-github-repository/referencing-and-citing-content  % emwAug10 -- I think you can leave it on GitHub and see what whichever journal you end up in wants
+% emwAug10 -- this is a great sentence below, I would get it into cover letter or abstract! I tagged on one version above for now ... 
+% Such an open-source device could facilitate the widespread sampling of tree rings across diverse ecosystems and reduce barriers for researchers with tight budgets.
+
 \end{abstract}
@@
   \includegraphics[height=0.5\linewidth]{../content/tina.jpg}
-  \caption{Complete assembly of TIM, shown with a few cookies ready to be scanned (in the work plane).}
+  \caption{Complete assembly of TIM, shown with a few cookies ready to be scanned (in the work plane).
+All of the 3D printed parts are available for download on the \href{https://doi.org/10.60705/3DPX/21561.3}{NIH 3D printing repository} and the software is
+open source, with instructions to recreate this tool available on \href{https://github.com/temporalecologylab/TreeRings}{GitHub}.
+} % emwAug10 -- In addition to putting the links here, I would make sure each is sprinkled in around where most of the info related to it is introduced in the main text.
   \label{fig:tim_assembled}
@@
 \subsection{Strengths and Opportunities} %emw9Mar: And this section could become part of the Results & Discussion....
+% emwAug10 -- check all your refs are working!
+I search for ?
+before I submit.
+Also you need to through and fix anything as shown below, I fixed one but there are others...
+If you have other ref issues, ask Victor or me for help.
+% \citep{girardin_national_2021} \citep{grissino-mayer_international_1997}. --> \citep{girardin_national_2021,grissino-mayer_international_1997}.
 % add better opportunities
@@
 One opportunity for this development could come as more universal and large-scale open-data storage of tree ring images come online. 
-Such open repositories would also increase the value of developing new frameworks count vessel or identify rings, which are already starting and have the potential to greatly decrease the time invested in manually identifying tree anatomy \citep{resente_mask_2021} \citep{polacek_automation_2023}.
+Such open repositories would also increase the value of developing new frameworks count vessel or identify rings, which are already starting and have the potential to greatly decrease the time invested in manually identifying tree anatomy \citep{resente_mask_2021,polacek_automation_2023}.
 With these more fully developed TIM could be extended to be run such frameworks while scanning, possibly taking additional scans only when needed to reach sufficient algorithm accuracy.
@@
 or something else (I'm not an expert)?
-\section{Acknowledgments}
-The authors thank Deirdre Loughnan, Hoai Huong Nguyen Phan, Frederik Baumgarten for conversations and ideas to help bring this project to fruition.
+\section{Acknowledgments} %emwAug10 -- also thank Victor added 
+The authors thank Deirdre Loughnan, Hoai Huong Nguyen Phan, Frederik Baumgarten and Victor Van der Meersch for conversations, ideas and review to help bring this project to fruition.
 Additionally, we thank Xiaomao Wang and Christophe Rouleau-Desrochers for their help and patience in operating TIM during development.
